### 2013-08-28 paper_id1.tex
initial draft working tex file

% This is "sig-alternate.tex" V2.0 May 2012
% This file should be compiled with V2.5 of "sig-alternate.cls" May 2012
%
% This example file demonstrates the use of the 'sig-alternate.cls'
% V2.5 LaTeX2e document class file. It is for those submitting
% articles to ACM Conference Proceedings WHO DO NOT WISH TO
% STRICTLY ADHERE TO THE SIGS (PUBS-BOARD-ENDORSED) STYLE.
% The 'sig-alternate.cls' file will produce a similar-looking,
% albeit, 'tighter' paper resulting in, invariably, fewer pages.
%
% ----------------------------------------------------------------------------------------------------------------
% This .tex file (and associated .cls V2.5) produces:
%       1) The Permission Statement
%       2) The Conference (location) Info information
%       3) The Copyright Line with ACM data
%       4) NO page numbers
%
% as against the acm_proc_article-sp.cls file which
% DOES NOT produce 1) thru' 3) above.
%
% Using 'sig-alternate.cls' you have control, however, from within
% the source .tex file, over both the CopyrightYear
% (defaulted to 200X) and the ACM Copyright Data
% (defaulted to X-XXXXX-XX-X/XX/XX).
% e.g.
% \CopyrightYear{2007} will cause 2007 to appear in the copyright line.
% \crdata{0-12345-67-8/90/12} will cause 0-12345-67-8/90/12 to appear in the copyright line.
%
% ---------------------------------------------------------------------------------------------------------------
% This .tex source is an example which *does* use
% the .bib file (from which the .bbl file % is produced).
% REMEMBER HOWEVER: After having produced the .bbl file,
% and prior to final submission, you *NEED* to 'insert'
% your .bbl file into your source .tex file so as to provide
% ONE 'self-contained' source file.
%
% ================= IF YOU HAVE QUESTIONS =======================
% Questions regarding the SIGS styles, SIGS policies and
% procedures, Conferences etc. should be sent to
% Adrienne Griscti (griscti@acm.org)
%
% Technical questions _only_ to
% Gerald Murray (murray@hq.acm.org)
% ===============================================================
%
% For tracking purposes - this is V2.0 - May 2012

\documentclass{sig-alternate}
% *** CITATION PACKAGES ***
%
\usepackage{cite}
% cite.sty was written by Donald Arseneau
% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
% \cite{} output to follow that of IEEE. Loading the cite package will
% result in citation numbers being automatically sorted and properly
% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
% \cite will automatically add leading space, if needed. Use cite.sty's
% noadjust option (cite.sty V3.8 and later) if you want to turn this off.
% cite.sty is already installed on most LaTeX systems. Be sure and use
% version 4.0 (2003-05-27) and later if using hyperref.sty. cite.sty does
% not currently provide for hyperlinked citations.
% The latest version can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/cite/
% The documentation is contained in the cite.sty file itself.

% *** GRAPHICS RELATED PACKAGES ***
%
%\if CLASSINFOpdf
  \usepackage{graphicx}
  \usepackage{epstopdf}
 % \usepackage{auto-pst-pdf}
  % declare the path(s) where your graphic files are
  % \graphicspath{{./eps}{./pdf/}{./jpeg/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  \DeclareGraphicsExtensions{.eps,.pdf,.jpeg,.png,.ps}
%\else
  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
  % will default to the driver specified in the system graphics.cfg if no
  % driver is specified.
 % \usepackage[dvips]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{./eps/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
 % \DeclareGraphicsExtensions{.eps}
%\fi
% graphicx was written by David Carlisle and Sebastian Rahtz. It is
% required if you want graphics, photos, etc. graphicx.sty is already
% installed on most LaTeX systems. The latest version and documentation can
% be obtained at: 
% http://www.ctan.org/tex-archive/macros/latex/required/graphics/
% Another good source of documentation is "Using Imported Graphics in
% LaTeX2e" by Keith Reckdahl which can be found as epslatex.ps or
% epslatex.pdf at: http://www.ctan.org/tex-archive/info/
%
% latex, and pdflatex in dvi mode, support graphics in encapsulated
% postscript (.eps) format. pdflatex in pdf mode supports graphics
% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
% not a bitmapped formats (.jpeg, .png). IEEE frowns on bitmapped formats
% which can result in "jaggedy"/blurry rendering of lines and letters as
% well as large increases in file sizes.
%
% You can find documentation about the pdfTeX application at:
% http://www.tug.org/applications/pdftex

% *** ALIGNMENT PACKAGES ***
%
%\usepackage{array}
% Frank Mittelbach's and David Carlisle's array.sty patches and improves
% the standard LaTeX2e array and tabular environments to provide better
% appearance and additional user controls. As the default LaTeX2e table
% generation code is lacking to the point of almost being broken with
% respect to the quality of the end results, all users are strongly
% advised to use an enhanced (at the very least that provided by array.sty)
% set of table tools. array.sty is already installed on most systems. The
% latest version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/required/tools/
\usepackage{balance}
\usepackage{mdframed}
%\usepackage{mdwmath}
%\usepackage{mdwtab}
% Also highly recommended is Mark Wooding's extremely powerful MDW tools,
% especially mdwmath.sty and mdwtab.sty which are used to format equations
% and tables, respectively. The MDWtools set is already installed on most
% LaTeX systems. The lastest version and documentation is available at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/mdwtools

\begin{document}
%
% --- Author Metadata here ---
\conferenceinfo{WOODSTOCK}{'97 El Paso, Texas USA}
%\CopyrightYear{2007} % Allows default copyright year (20XX) to be over-ridden - IF NEED BE.
%\crdata{0-12345-67-8/90/01}  % Allows default copyright data (0-89791-88-6/97/05) to be over-ridden - IF NEED BE.
% --- End of Author Metadata ---

\title{Effectiveness of Instant Feedback on Developer Navigation Practices}
\subtitle{Does Instant Feedback Make Us Better Developers?}
%
% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in
% the body of your article BEFORE References or any Appendices.

\numberofauthors{3} %  in this sample file, there are a *total*
% of EIGHT authors. SIX appear on the 'first-page' (for formatting
% reasons) and the remaining two appear in the \additionalauthors section.
%
% author names and affiliations
% use a multiple column layout for up to three different
% affiliations
\author{
\alignauthor
Will Snipes
\affaddr{ABB Corporate Research}\\
\affaddr{Industrial Software Systems}\\
\affaddr{Raleigh, NC USA}\\
\email{will.snipes@us.abb.com}
\and
\alignauthor
Anil R. Nair
\affaddr{ABB Corporate Research}\\
\affaddr{Industrial Software Systems}\\
\affaddr{Bangalore, India}\\
\email{anil.nair@in.abb.com}
\and
\alignauthor
Emerson Murphy-Hill
\affaddr{North Carolina State University}\\
\affaddr{Department of Computer Science}\\
\affaddr{Raleigh, N.C. USA}\\
\email{emerson@csc.ncsu.edu}\\
}

\maketitle
\begin{abstract}
As software development practices evolve, we face the continuous challenge of communicating new practices and tools to the community.  As developers, we both like to try new things and like to stick with the familiar.  In addition to training, discussing, and presenting software engineering practices and tools, we seek to communicate them through active two-way information embedded in the Integrated Development Environment (IDE).  While communicating, we also collect data on new and current practices and tools used by each developer by recording actions in the IDE.  With a rich data source, we evaluate what practices are used, define metrics that give individual feedback, and create an environment where developers can see how they are doing compared with their teammates.

This approach helps understand practice and tool use over the long term and judges the lasting change a communication effort makes in everyday habits.  Tool support provides a mechanism to embed into Visual Studio an extension that provides the capture of usage data while giving feedback to users on their practices both in general and specific to the ones being promoted.  Through a points score system that assigns points to specific commands and tools we encourage everyone to use the new practices.  Providing a light score ranking view should help us stay motivated as we compare our command usage to our peers.  To demonstrate how this works, we implement a longitudinal study that demonstrates the results from this approach as we engage with developers over a month-long period.

Results from this demonstrative study show that in general the approach is <blank> with a few developers <blank> of the study.  The effectiveness of this approach for communicating about practices is demonstrated through the continued use of the commands as we communicate them.  Evaluation of metrics for navigation ratio over time show that developer use the commands <blank> than they did during the benchmarking phase.  The result of the study shows that this method for communicating and promoting software engineering tools and practices <blank>.
\end{abstract}

% A category with the (minimum) three required fields
\category{H.4}{Information Systems Applications}{Miscellaneous}
%A category including the fourth, optional field follows...
\category{D.2.8}{Software Engineering}{Metrics}[complexity measures, performance measures]

\terms{Theory}

\keywords{ACM proceedings, \LaTeX, text tagging}

\section{Introduction}

\section{Study Design}
Creating a longitudinal study involves defining a long-term goal that is measurable in increments, determining the discrete steps that the study will use to communicate the goal to the participants, distribute interventions to participants, and track participants progress over time.  Many software engineering studies are conducted at discrete points in time via surveys of sampled participants.  Perhaps prior to an intervention, and multiple points following an intervention.  This study demonstrates a method to perform an intervention, continuously monitor the progress of participant response to that intervention, and provide continuous feedback to the participant and researchers about the progress.  Continuous meaning that no information gaps exist in the data collection, there are no periods of unmonitored activity by the participant.  Continuous also applies to the intervention being evaluated, where the participant and researchers receive constant feedback about the conformance to the intervention's purpose.  Studying an intervention in a continuous context determines how sticky the intervention is in the mind of the developer.  It can remove bias that might be introduced by periodic surveys or audits and enhance the habits formed in the participant receiving the intervention.

Designing the study with continuous measurement but no control group presents a traditional study design with some challenges.  We overcame this by dividing the study in to phases where the participants themselves form a control group.  The first phase baselines their activities determining the current state of their practices.  The second phase presents the intervention to the participants.  This phase the participants do not receive continuous feedback on their compliance with the intervention.  This phase simulates the traditional approach of communicating an intervention then returning with an audit to determine whether participants are complaint with it.   The last phases turns on continuous feedback to the participants on their compliance with the intervention.  As they receive the feedback, the participants can determine for themselves how much they are implementing the intervention.  In the last phase they possess complete knowledge of the intervention and their compliance with it making it plausible that they can continuously control their compliance.

Developing the study with the available tooling means that we follow steps as follows.
1. define the practice we wish to measure compliance with
2. Develop a mapping of the practice to actions that are captured by the blaze tool from the IDE
3. Weight the actions according to their importance to the practice and assign points to the actions in the configuration file
4. Develop the intervention method
5. Determine the phases and length of the phase in the study according to baseline, intervention, and continuous feedback
6. Select participants to deploy
7. Define a pre-study questionnaire if one is necessary to collect associated information
8. Define a post-study questionnaire if necessary to collect post study information
9. Communicate with the participants about the study
10. Execute the study and report results

\subsection{Type Changes and {\subsecit Special} Characters}
We have already seen several typeface changes in this sample.  You
can indicate italicized words or phrases in your text with
the command \texttt{{\char'134}textit}; emboldening with the
command \texttt{{\char'134}textbf}
and typewriter-style (for instance, for computer code) with
\texttt{{\char'134}texttt}.  But remember, you do not
have to indicate typestyle changes when such changes are
part of the \textit{structural} elements of your
article; for instance, the heading of this subsection will
be in a sans serif\footnote{A third footnote, here.
Let's make this a rather short one to
see how it looks.} typeface, but that is handled by the
document class file. Take care with the use
of\footnote{A fourth, and last, footnote.}
the curly braces in typeface changes; they mark
the beginning and end of
the text that is to be in the different typeface.

You can use whatever symbols, accented characters, or
non-English characters you need anywhere in your document;
you can find a complete list of what is
available in the \textit{\LaTeX\
User's Guide}\cite{Lamport:LaTeX}.

\subsection{Math Equations}
You may want to display math equations in three distinct styles:
inline, numbered or non-numbered display.  Each of
the three are discussed in the next sections.

\subsubsection{Inline (In-text) Equations}
A formula that appears in the running text is called an
inline or in-text formula.  It is produced by the
\textbf{math} environment, which can be
invoked with the usual \texttt{{\char'134}begin. . .{\char'134}end}
construction or with the short form \texttt{\$. . .\$}. You
can use any of the symbols and structures,
from $\alpha$ to $\omega$, available in
\LaTeX\cite{Lamport:LaTeX}; this section will simply show a
few examples of in-text equations in context. Notice how
this equation: \begin{math}\lim_{n\rightarrow \infty}x=0\end{math},
set here in in-line math style, looks slightly different when
set in display style.  (See next section).

\subsubsection{Display Equations}
A numbered display equation -- one set off by vertical space
from the text and centered horizontally -- is produced
by the \textbf{equation} environment. An unnumbered display
equation is produced by the \textbf{displaymath} environment.

Again, in either environment, you can use any of the symbols
and structures available in \LaTeX; this section will just
give a couple of examples of display equations in context.
First, consider the equation, shown as an inline equation above:
\begin{equation}\lim_{n\rightarrow \infty}x=0\end{equation}
Notice how it is formatted somewhat differently in
the \textbf{displaymath}
environment.  Now, we'll enter an unnumbered equation:
\begin{displaymath}\sum_{i=0}^{\infty} x + 1\end{displaymath}
and follow it with another numbered equation:
\begin{equation}\sum_{i=0}^{\infty}x_i=\int_{0}^{\pi+2} f\end{equation}
just to demonstrate \LaTeX's able handling of numbering.

\subsection{Citations}
Citations to articles \cite{bowman:reasoning,
clark:pct, braams:babel, herlihy:methodology},
conference proceedings \cite{clark:pct} or
books \cite{salas:calculus, Lamport:LaTeX} listed
in the Bibliography section of your
article will occur throughout the text of your article.
You should use BibTeX to automatically produce this bibliography;
you simply need to insert one of several citation commands with
a key of the item cited in the proper location in
the \texttt{.tex} file \cite{Lamport:LaTeX}.
The key is a short reference you invent to uniquely
identify each work; in this sample document, the key is
the first author's surname and a
word from the title.  This identifying key is included
with each item in the \texttt{.bib} file for your article.

The details of the construction of the \texttt{.bib} file
are beyond the scope of this sample document, but more
information can be found in the \textit{Author's Guide},
and exhaustive details in the \textit{\LaTeX\ User's
Guide}\cite{Lamport:LaTeX}.

This article shows only the plainest form
of the citation command, using \texttt{{\char'134}cite}.
This is what is stipulated in the SIGS style specifications.
No other citation format is endorsed or supported.

\subsection{Tables}
Because tables cannot be split across pages, the best
placement for them is typically the top of the page
nearest their initial cite.  To
ensure this proper ``floating'' placement of tables, use the
environment \textbf{table} to enclose the table's contents and
the table caption.  The contents of the table itself must go
in the \textbf{tabular} environment, to
be aligned properly in rows and columns, with the desired
horizontal and vertical rules.  Again, detailed instructions
on \textbf{tabular} material
is found in the \textit{\LaTeX\ User's Guide}.

Immediately following this sentence is the point at which
Table 1 is included in the input file; compare the
placement of the table here with the table in the printed
dvi output of this document.

\begin{table}
\centering
\caption{Frequency of Special Characters}
\begin{tabular}{|c|c|l|} \hline
Non-English or Math&Frequency&Comments\\ \hline
\O & 1 in 1,000& For Swedish names\\ \hline
$\pi$ & 1 in 5& Common in math\\ \hline
\$ & 4 in 5 & Used in business\\ \hline
$\Psi^2_1$ & 1 in 40,000& Unexplained usage\\
\hline\end{tabular}
\end{table}

To set a wider table, which takes up the whole width of
the page's live area, use the environment
\textbf{table*} to enclose the table's contents and
the table caption.  As with a single-column table, this wide
table will ``float" to a location deemed more desirable.
Immediately following this sentence is the point at which
Table 2 is included in the input file; again, it is
instructive to compare the placement of the
table here with the table in the printed dvi
output of this document.

\begin{table*}
\centering
\caption{Some Typical Commands}
\begin{tabular}{|c|c|l|} \hline
Command&A Number&Comments\\ \hline
\texttt{{\char'134}alignauthor} & 100& Author alignment\\ \hline
\texttt{{\char'134}numberofauthors}& 200& Author enumeration\\ \hline
\texttt{{\char'134}table}& 300 & For tables\\ \hline
\texttt{{\char'134}table*}& 400& For wider tables\\ \hline\end{tabular}
\end{table*}
% end the environment with {table*}, NOTE not {table}!

\subsection{Figures}
Like tables, figures cannot be split across pages; the
best placement for them
is typically the top or the bottom of the page nearest
their initial cite.  To ensure this proper ``floating'' placement
of figures, use the environment
\textbf{figure} to enclose the figure and its caption.

This sample document contains examples of \textbf{.eps}
and \textbf{.ps} files to be displayable with \LaTeX.  More
details on each of these is found in the \textit{Author's Guide}.

\begin{figure}
\centering
\epsfig{file=fly.eps}
\caption{A sample black and white graphic (.eps format).}
\end{figure}

\begin{figure}
\centering
\epsfig{file=fly.eps, height=1in, width=1in}
\caption{A sample black and white graphic (.eps format)
that has been resized with the \texttt{epsfig} command.}
\end{figure}

As was the case with tables, you may want a figure
that spans two columns.  To do this, and still to
ensure proper ``floating'' placement of tables, use the environment
\textbf{figure*} to enclose the figure and its caption.
and don't forget to end the environment with
{figure*}, not {figure}!

\begin{figure*}
\centering
\epsfig{file=flies.eps}
\caption{A sample black and white graphic (.eps format)
that needs to span two columns of text.}
\end{figure*}

Note that either {\textbf{.ps}} or {\textbf{.eps}} formats are
used; use
the \texttt{{\char'134}epsfig} or \texttt{{\char'134}psfig}
commands as appropriate for the different file types.

\begin{figure}
\centering
\psfig{file=rosette.ps, height=1in, width=1in,}
\caption{A sample black and white graphic (.ps format) that has
been resized with the \texttt{psfig} command.}
\vskip -6pt
\end{figure}

\subsection{Theorem-like Constructs}
Other common constructs that may occur in your article are
the forms for logical constructs like theorems, axioms,
corollaries and proofs.  There are
two forms, one produced by the
command \texttt{{\char'134}newtheorem} and the
other by the command \texttt{{\char'134}newdef}; perhaps
the clearest and easiest way to distinguish them is
to compare the two in the output of this sample document:

This uses the \textbf{theorem} environment, created by
the\linebreak\texttt{{\char'134}newtheorem} command:
\newtheorem{theorem}{Theorem}
\begin{theorem}
Let $f$ be continuous on $[a,b]$.  If $G$ is
an antiderivative for $f$ on $[a,b]$, then
\begin{displaymath}\int^b_af(t)dt = G(b) - G(a).\end{displaymath}
\end{theorem}

The other uses the \textbf{definition} environment, created
by the \texttt{{\char'134}newdef} command:
\newdef{definition}{Definition}
\begin{definition}
If $z$ is irrational, then by $e^z$ we mean the
unique number which has
logarithm $z$: \begin{displaymath}{\log e^z = z}\end{displaymath}
\end{definition}

Two lists of constructs that use one of these
forms is given in the
\textit{Author's  Guidelines}.

There is one other similar construct environment, which is
already set up
for you; i.e. you must \textit{not} use
a \texttt{{\char'134}newdef} command to
create it: the \textbf{proof} environment.  Here
is a example of its use:
\begin{proof}
Suppose on the contrary there exists a real number $L$ such that
\begin{displaymath}
\lim_{x\rightarrow\infty} \frac{f(x)}{g(x)} = L.
\end{displaymath}
Then
\begin{displaymath}
l=\lim_{x\rightarrow c} f(x)
= \lim_{x\rightarrow c}
\left[ g{x} \cdot \frac{f(x)}{g(x)} \right ]
= \lim_{x\rightarrow c} g(x) \cdot \lim_{x\rightarrow c}
\frac{f(x)}{g(x)} = 0\cdot L = 0,
\end{displaymath}
which contradicts our assumption that $l\neq 0$.
\end{proof}

Complete rules about using these environments and using the
two different creation commands are in the
\textit{Author's Guide}; please consult it for more
detailed instructions.  If you need to use another construct,
not listed therein, which you want to have the same
formatting as the Theorem
or the Definition\cite{salas:calculus} shown above,
use the \texttt{{\char'134}newtheorem} or the
\texttt{{\char'134}newdef} command,
respectively, to create it.

\subsection*{A {\secit Caveat} for the \TeX\ Expert}
Because you have just been given permission to
use the \texttt{{\char'134}newdef} command to create a
new form, you might think you can
use \TeX's \texttt{{\char'134}def} to create a
new command: \textit{Please refrain from doing this!}
Remember that your \LaTeX\ source code is primarily intended
to create camera-ready copy, but may be converted
to other forms -- e.g. HTML. If you inadvertently omit
some or all of the \texttt{{\char'134}def}s recompilation will
be, to say the least, problematic.

\section{Conclusions}
This paragraph will end the body of this sample document.
Remember that you might still have Acknowledgments or
Appendices; brief samples of these
follow.  There is still the Bibliography to deal with; and
we will make a disclaimer about that here: with the exception
of the reference to the \LaTeX\ book, the citations in
this paper are to articles which have nothing to
do with the present subject and are used as
examples only.
%\end{document}  % This is where a 'short' article might terminate

%ACKNOWLEDGMENTS are optional
\section{Acknowledgments}
This section is optional; it is a location for you
to acknowledge grants, funding, editing assistance and
what have you.  In the present case, for example, the
authors would like to thank Gerald Murray of ACM for
his help in codifying this \textit{Author's Guide}
and the \textbf{.cls} and \textbf{.tex} files that it describes.

%
% The following two commands are all you need in the
% initial runs of your .tex file to
% produce the bibliography for the citations in your paper.
\bibliographystyle{abbrv}
\bibliography{sigproc}  % sigproc.bib is the name of the Bibliography in this case
% You must have a proper ".bib" file
%  and remember to run:
% latex bibtex latex latex
% to resolve all references
%
% ACM needs 'a single self-contained file'!
%
%APPENDICES are optional
%\balancecolumns
\appendix
%Appendix A
\section{Headings in Appendices}
The rules about hierarchical headings discussed above for
the body of the article are different in the appendices.
In the \textbf{appendix} environment, the command
\textbf{section} is used to
indicate the start of each Appendix, with alphabetic order
designation (i.e. the first is A, the second B, etc.) and
a title (if you include one).  So, if you need
hierarchical structure
\textit{within} an Appendix, start with \textbf{subsection} as the
highest level. Here is an outline of the body of this
document in Appendix-appropriate form:
\subsection{Introduction}
\subsection{The Body of the Paper}
\subsubsection{Type Changes and  Special Characters}
\subsubsection{Math Equations}
\paragraph{Inline (In-text) Equations}
\paragraph{Display Equations}
\subsubsection{Citations}
\subsubsection{Tables}
\subsubsection{Figures}
\subsubsection{Theorem-like Constructs}
\subsubsection*{A Caveat for the \TeX\ Expert}
\subsection{Conclusions}
\subsection{Acknowledgments}
\subsection{References}
Generated by bibtex from your ~.bib file.  Run latex,
then bibtex, then latex twice (to resolve references)
to create the ~.bbl file.  Insert that ~.bbl file into
the .tex source file and comment out
the command \texttt{{\char'134}thebibliography}.
% This next section command marks the start of
% Appendix B, and does not continue the present hierarchy
\section{More Help for the Hardy}
The sig-alternate.cls file itself is chock-full of succinct
and helpful comments.  If you consider yourself a moderately
experienced to expert user of \LaTeX, you may find reading
it useful but please remember not to change it.
%\balancecolumns % GM June 2007
% That's all folks!
\end{document}

### 2013-08-29 paper_id1.tex
Added bibliography

@@
 \section{Introduction}
+%update
+Deploying these techniques requires communicating and motivating developers to use them.
+Murphy-Hill and colleagues  \cite{wbsnipes:Hill2011Peer} show that developers can effectively learn about new tools from their peers by watching them work and through recommendations, yet this happens very rarely, in part because developers rarely work on the same computer at the same time.
+This personal connection limits the knowledge transfer to the co-located circle of colleagues connected with the early adopters.
+%update
+Finding a way to automate the communication of technique recommendations between developers enhances deployment of best practices in a globally distributed software development organization.   
+%PF the following sounds redundant
+To best match the peer recommendation scenario described above, the method should mimic the personal recommendation model while rewarding successful adopters.
 
@@
 Developing the study with the available tooling means that we follow steps as follows.
-1. define the practice we wish to measure compliance with
+1. Define the practice we wish to measure compliance with
 2. Develop a mapping of the practice to actions that are captured by the blaze tool from the IDE
@@
 10. Execute the study and report results
+
+\subsection{Setup the practice}
+Selecting a practice to study involves defining specific characteristics of the practice and how capturing actions in Visual Studio relates to the practice.
+In this study we selected the structured navigation practice shown by Robillard et.
+al.
+as correlated with effective program maintenance \cite{wbsnipes:Robillard2004How} .
 
@@
 \bibliographystyle{abbrv}
-\bibliography{sigproc}  % sigproc.bib is the name of the Bibliography in this case
+\balance
+\bibliography{paper_id1}  % sigproc.bib is the name of the Bibliography in this case
 % You must have a proper ".bib" file
@@
 the .tex source file and comment out
-the command \texttt{{\char'134}thebibliography}.
+the command% \texttt{{\char'134}thebibliography}.
 % This next section command marks the start of

### 2013-08-29 paper_id1.tex
making progress

@@
 
-\terms{Theory}
+\terms{gamification}
 
-\keywords{ACM proceedings, \LaTeX, text tagging}
+\keywords{empirical, training, learning, usability}
 
@@
 as correlated with effective program maintenance \cite{wbsnipes:Robillard2004How} .
+Studying structured navigation is a natural fit for capturing commands issued in the IDE.
+The first step is to create the classification of commands that are involved in navigation.
+Then differentiate which commands we consider as structured navigation from unstructured navigation.
 
-\subsection{Type Changes and {\subsecit Special} Characters}
-We have already seen several typeface changes in this sample.  You
-can indicate italicized words or phrases in your text with
-the command \texttt{{\char'134}textit}; emboldening with the
-command \texttt{{\char'134}textbf}
-and typewriter-style (for instance, for computer code) with
-\texttt{{\char'134}texttt}.  But remember, you do not
-have to indicate typestyle changes when such changes are
-part of the \textit{structural} elements of your
-article; for instance, the heading of this subsection will
-be in a sans serif\footnote{A third footnote, here.
-Let's make this a rather short one to
-see how it looks.} typeface, but that is handled by the
-document class file. Take care with the use
-of\footnote{A fourth, and last, footnote.}
-the curly braces in typeface changes; they mark
-the beginning and end of
-the text that is to be in the different typeface.
+The Blaze tool provides an XML configuration file that allows defining multiple levels of categories for commands thus we can categorize commands that are navigation then further classify them into structured and unstructured navigation.
+We classified commands as structured navigation when they allow the developer to locate a code element following the program structure from another element.
+For example, Navigate To \textit{Ctrl+,}, Go To Definition \textit{F12}, and Find All References \textit{Ctrl+K,R} are commands that follow structure of the code that we classify as structured navigation.
+Commands that we considered as unstructured navigation include scrolling through a file using keys or the scroll-bar, opening files from the Solution Explorer, Find commands (e.g.
+Find in Files), Go-to line number, and children of these commands (e.g.
+Find Next).
 
-You can use whatever symbols, accented characters, or
-non-English characters you need anywhere in your document;
-you can find a complete list of what is
-available in the \textit{\LaTeX\
-User's Guide}\cite{Lamport:LaTeX}.
+After classifying the commands we assigned points to the commands in the same configuration file.
+Structured navigation commands received 1-10 points each while unstructured navigation commands received 0 points.
+Assigning negative points while possible, is discouraged as this may demotivate folks who do not have detailed knowledge of how the study is designed.
+When assigning points, make sure you recognize the relationship between points and levels such that "leveling up" happens at a frequent enough pace to make that motivating for users during the study.
+Ideally users would level up after a working day or two worth of activity initially extending to a week or even a month following an exponential curve \cite{gamification}.
+Designing the points and levels such that above average developers can complete all levels during the study period is ideal.
 
-\subsection{Math Equations}
-You may want to display math equations in three distinct styles:
-inline, numbered or non-numbered display.  Each of
-the three are discussed in the next sections.
+\subsection{Study Intervention and Participants}
+Part of the study can include a communication mechanism that informs the users about the study or even provides training relevant to the objective of the study via a web site.
+We discuss this as the "intervention" that expresses the practice that we wish developers to adopt.
+For this study we chose to make a nice looking graphic that users click on to see information about navigation commands, install the Sando search tool, and learn about the Blaze game.
+The site uses a combination of video and blog-post style information easily accessible to the developer.
+The communication site is rated as part of our study completion survey.
 
-\subsubsection{Inline (In-text) Equations}
-A formula that appears in the running text is called an
-inline or in-text formula.  It is produced by the
-\textbf{math} environment, which can be
-invoked with the usual \texttt{{\char'134}begin. . .{\char'134}end}
-construction or with the short form \texttt{\$. . .\$}. You
-can use any of the symbols and structures,
-from $\alpha$ to $\omega$, available in
-\LaTeX\cite{Lamport:LaTeX}; this section will simply show a
-few examples of in-text equations in context. Notice how
-this equation: \begin{math}\lim_{n\rightarrow \infty}x=0\end{math},
-set here in in-line math style, looks slightly different when
-set in display style.  (See next section).
+Timing the study communication is part of the consideration for a good study design.
+We chose to deploy the Blaze tool without communicating the study purpose for a baseline period, then the tool would pop-up a display with the information on the communication site asking the developers to consume the information and begin using the practices.
+This allows us to measure changes in the developer's practices over time during the study.
+%Other intervention approaches may be used, however, you should take care to understand how the method and timing may affect the measurement of the practices.
 
-\subsubsection{Display Equations}
-A numbered display equation -- one set off by vertical space
-from the text and centered horizontally -- is produced
-by the \textbf{equation} environment. An unnumbered display
-equation is produced by the \textbf{displaymath} environment.
+We conducted this study with developers working in an R\&D facility in India on a large industrial software system.
+Developers were asked to volunteer for the pilot by their management, however, managers did not know who participated in the study.
+Keeping managers out of the loop is intended to reduce concerns among the developers that the data may be used for purposes other then the intended study.
+By selecting an intact team, we hope to leverage aspects of motivation such as the Leaderboard to spur feelings of competition among teammates.
 
-Again, in either environment, you can use any of the symbols
-and structures available in \LaTeX; this section will just
-give a couple of examples of display equations in context.
-First, consider the equation, shown as an inline equation above:
-\begin{equation}\lim_{n\rightarrow \infty}x=0\end{equation}
-Notice how it is formatted somewhat differently in
-the \textbf{displaymath}
-environment.  Now, we'll enter an unnumbered equation:
-\begin{displaymath}\sum_{i=0}^{\infty} x + 1\end{displaymath}
-and follow it with another numbered equation:
-\begin{equation}\sum_{i=0}^{\infty}x_i=\int_{0}^{\pi+2} f\end{equation}
-just to demonstrate \LaTeX's able handling of numbering.
+\section{Related Work}
+%update and expand
+Researchers have developed systems that aim to answer specific questions using data from command logs.
+Zorro \cite{V:Johnson2007Automated} leverages the HackyStat data to determine if developers are adhering to test-driven development.
+The tool employs a similar approach dividing the work of the developer into episodes delineated by when test cases pass.
+The researchers define multiple types of episodes then determine whether the developer uses Test-Driven-Development or not in each episode.
 
-\subsection{Citations}
-Citations to articles \cite{bowman:reasoning,
-clark:pct, braams:babel, herlihy:methodology},
-conference proceedings \cite{clark:pct} or
-books \cite{salas:calculus, Lamport:LaTeX} listed
-in the Bibliography section of your
-article will occur throughout the text of your article.
-You should use BibTeX to automatically produce this bibliography;
-you simply need to insert one of several citation commands with
-a key of the item cited in the proper location in
-the \texttt{.tex} file \cite{Lamport:LaTeX}.
-The key is a short reference you invent to uniquely
-identify each work; in this sample document, the key is
-the first author's surname and a
-word from the title.  This identifying key is included
-with each item in the \texttt{.bib} file for your article.
+Murphy-Hill, Parnin, and Black \cite{V:MurphyHill2012How} use the Mylyn Monitor to explore whether or not developers use the automated refactoring tools present in Eclipse.
+They look for specific refactoring commands in Eclipse and determine the amount of time developers use tools versus hand refactoring the code.
 
-The details of the construction of the \texttt{.bib} file
-are beyond the scope of this sample document, but more
-information can be found in the \textit{Author's Guide},
-and exhaustive details in the \textit{\LaTeX\ User's
-Guide}\cite{Lamport:LaTeX}.
+Robillard, Coelho and Murphy explore hypotheses around how developers can be more effective at performing a maintenance task \cite{wbsnipes:Robillard2004How}.
+Key conclusions are that developers are more more successful finding and fixing bugs when they create a detailed plan for implementing a change, use structured navigation with keyword search or cross-reference search, and only review methods once during their search.
 
-This article shows only the plainest form
-of the citation command, using \texttt{{\char'134}cite}.
-This is what is stipulated in the SIGS style specifications.
-No other citation format is endorsed or supported.
-
-\subsection{Tables}
-Because tables cannot be split across pages, the best
-placement for them is typically the top of the page
-nearest their initial cite.  To
-ensure this proper ``floating'' placement of tables, use the
-environment \textbf{table} to enclose the table's contents and
-the table caption.  The contents of the table itself must go
-in the \textbf{tabular} environment, to
-be aligned properly in rows and columns, with the desired
-horizontal and vertical rules.  Again, detailed instructions
-on \textbf{tabular} material
-is found in the \textit{\LaTeX\ User's Guide}.
-
-Immediately following this sentence is the point at which
-Table 1 is included in the input file; compare the
-placement of the table here with the table in the printed
-dvi output of this document.
-
-\begin{table}
-\centering
-\caption{Frequency of Special Characters}
-\begin{tabular}{|c|c|l|} \hline
-Non-English or Math&Frequency&Comments\\ \hline
-\O & 1 in 1,000& For Swedish names\\ \hline
-$\pi$ & 1 in 5& Common in math\\ \hline
-\$ & 4 in 5 & Used in business\\ \hline
-$\Psi^2_1$ & 1 in 40,000& Unexplained usage\\
-\hline\end{tabular}
-\end{table}
-
-To set a wider table, which takes up the whole width of
-the page's live area, use the environment
-\textbf{table*} to enclose the table's contents and
-the table caption.  As with a single-column table, this wide
-table will ``float" to a location deemed more desirable.
-Immediately following this sentence is the point at which
-Table 2 is included in the input file; again, it is
-instructive to compare the placement of the
-table here with the table in the printed dvi
-output of this document.
-
-\begin{table*}
-\centering
-\caption{Some Typical Commands}
-\begin{tabular}{|c|c|l|} \hline
-Command&A Number&Comments\\ \hline
-\texttt{{\char'134}alignauthor} & 100& Author alignment\\ \hline
-\texttt{{\char'134}numberofauthors}& 200& Author enumeration\\ \hline
-\texttt{{\char'134}table}& 300 & For tables\\ \hline
-\texttt{{\char'134}table*}& 400& For wider tables\\ \hline\end{tabular}
-\end{table*}
-% end the environment with {table*}, NOTE not {table}!
-
-\subsection{Figures}
-Like tables, figures cannot be split across pages; the
-best placement for them
-is typically the top or the bottom of the page nearest
-their initial cite.  To ensure this proper ``floating'' placement
-of figures, use the environment
-\textbf{figure} to enclose the figure and its caption.
-
-This sample document contains examples of \textbf{.eps}
-and \textbf{.ps} files to be displayable with \LaTeX.  More
-details on each of these is found in the \textit{Author's Guide}.
-
-\begin{figure}
-\centering
-\epsfig{file=fly.eps}
-\caption{A sample black and white graphic (.eps format).}
-\end{figure}
-
-\begin{figure}
-\centering
-\epsfig{file=fly.eps, height=1in, width=1in}
-\caption{A sample black and white graphic (.eps format)
-that has been resized with the \texttt{epsfig} command.}
-\end{figure}
-
-As was the case with tables, you may want a figure
-that spans two columns.  To do this, and still to
-ensure proper ``floating'' placement of tables, use the environment
-\textbf{figure*} to enclose the figure and its caption.
-and don't forget to end the environment with
-{figure*}, not {figure}!
-
-\begin{figure*}
-\centering
-\epsfig{file=flies.eps}
-\caption{A sample black and white graphic (.eps format)
-that needs to span two columns of text.}
-\end{figure*}
-
-Note that either {\textbf{.ps}} or {\textbf{.eps}} formats are
-used; use
-the \texttt{{\char'134}epsfig} or \texttt{{\char'134}psfig}
-commands as appropriate for the different file types.
-
-\begin{figure}
-\centering
-\psfig{file=rosette.ps, height=1in, width=1in,}
-\caption{A sample black and white graphic (.ps format) that has
-been resized with the \texttt{psfig} command.}
-\vskip -6pt
-\end{figure}
-
-\subsection{Theorem-like Constructs}
-Other common constructs that may occur in your article are
-the forms for logical constructs like theorems, axioms,
-corollaries and proofs.  There are
-two forms, one produced by the
-command \texttt{{\char'134}newtheorem} and the
-other by the command \texttt{{\char'134}newdef}; perhaps
-the clearest and easiest way to distinguish them is
-to compare the two in the output of this sample document:
-
-This uses the \textbf{theorem} environment, created by
-the\linebreak\texttt{{\char'134}newtheorem} command:
-\newtheorem{theorem}{Theorem}
-\begin{theorem}
-Let $f$ be continuous on $[a,b]$.  If $G$ is
-an antiderivative for $f$ on $[a,b]$, then
-\begin{displaymath}\int^b_af(t)dt = G(b) - G(a).\end{displaymath}
-\end{theorem}
-
-The other uses the \textbf{definition} environment, created
-by the \texttt{{\char'134}newdef} command:
-\newdef{definition}{Definition}
-\begin{definition}
-If $z$ is irrational, then by $e^z$ we mean the
-unique number which has
-logarithm $z$: \begin{displaymath}{\log e^z = z}\end{displaymath}
-\end{definition}
-
-Two lists of constructs that use one of these
-forms is given in the
-\textit{Author's  Guidelines}.
-
-There is one other similar construct environment, which is
-already set up
-for you; i.e. you must \textit{not} use
-a \texttt{{\char'134}newdef} command to
-create it: the \textbf{proof} environment.  Here
-is a example of its use:
-\begin{proof}
-Suppose on the contrary there exists a real number $L$ such that
-\begin{displaymath}
-\lim_{x\rightarrow\infty} \frac{f(x)}{g(x)} = L.
-\end{displaymath}
-Then
-\begin{displaymath}
-l=\lim_{x\rightarrow c} f(x)
-= \lim_{x\rightarrow c}
-\left[ g{x} \cdot \frac{f(x)}{g(x)} \right ]
-= \lim_{x\rightarrow c} g(x) \cdot \lim_{x\rightarrow c}
-\frac{f(x)}{g(x)} = 0\cdot L = 0,
-\end{displaymath}
-which contradicts our assumption that $l\neq 0$.
-\end{proof}
-
-Complete rules about using these environments and using the
-two different creation commands are in the
-\textit{Author's Guide}; please consult it for more
-detailed instructions.  If you need to use another construct,
-not listed therein, which you want to have the same
-formatting as the Theorem
-or the Definition\cite{salas:calculus} shown above,
-use the \texttt{{\char'134}newtheorem} or the
-\texttt{{\char'134}newdef} command,
-respectively, to create it.
-
-\subsection*{A {\secit Caveat} for the \TeX\ Expert}
-Because you have just been given permission to
-use the \texttt{{\char'134}newdef} command to create a
-new form, you might think you can
-use \TeX's \texttt{{\char'134}def} to create a
-new command: \textit{Please refrain from doing this!}
-Remember that your \LaTeX\ source code is primarily intended
-to create camera-ready copy, but may be converted
-to other forms -- e.g. HTML. If you inadvertently omit
-some or all of the \texttt{{\char'134}def}s recompilation will
-be, to say the least, problematic.
+Building on this success, this work proposes a generalized method to automate the identification of the techniques described in the related work and others, and determine when the techniques would help a developer improve their work patterns.
+Then we intend to create a game-like virtual environment where developers learn from each other and are motivated to adopt the best practices developed by these and other researchers.
 
@@
 %
-%APPENDICES are optional
-%\balancecolumns
-\appendix
-%Appendix A
-\section{Headings in Appendices}
-The rules about hierarchical headings discussed above for
-the body of the article are different in the appendices.
-In the \textbf{appendix} environment, the command
-\textbf{section} is used to
-indicate the start of each Appendix, with alphabetic order
-designation (i.e. the first is A, the second B, etc.) and
-a title (if you include one).  So, if you need
-hierarchical structure
-\textit{within} an Appendix, start with \textbf{subsection} as the
-highest level. Here is an outline of the body of this
-document in Appendix-appropriate form:
-\subsection{Introduction}
-\subsection{The Body of the Paper}
-\subsubsection{Type Changes and  Special Characters}
-\subsubsection{Math Equations}
-\paragraph{Inline (In-text) Equations}
-\paragraph{Display Equations}
-\subsubsection{Citations}
-\subsubsection{Tables}
-\subsubsection{Figures}
-\subsubsection{Theorem-like Constructs}
-\subsubsection*{A Caveat for the \TeX\ Expert}
-\subsection{Conclusions}
-\subsection{Acknowledgments}
-\subsection{References}
-Generated by bibtex from your ~.bib file.  Run latex,
-then bibtex, then latex twice (to resolve references)
-to create the ~.bbl file.  Insert that ~.bbl file into
-the .tex source file and comment out
-the command% \texttt{{\char'134}thebibliography}.
-% This next section command marks the start of
-% Appendix B, and does not continue the present hierarchy
-\section{More Help for the Hardy}
-The sig-alternate.cls file itself is chock-full of succinct
-and helpful comments.  If you consider yourself a moderately
-experienced to expert user of \LaTeX, you may find reading
-it useful but please remember not to change it.
-%\balancecolumns % GM June 2007
-% That's all folks!
+
 \end{document}

### 2013-09-01 paper_id1.tex
interim updates

@@
 When assigning points, make sure you recognize the relationship between points and levels such that "leveling up" happens at a frequent enough pace to make that motivating for users during the study.
-Ideally users would level up after a working day or two worth of activity initially extending to a week or even a month following an exponential curve \cite{gamification}.
+Ideally users would level up after a working day or two worth of activity initially extending to a week or even a month following an exponential curve \cite{}.
 Designing the points and levels such that above average developers can complete all levels during the study period is ideal.
@@
 \subsection{Study Intervention and Participants}
-Part of the study can include a communication mechanism that informs the users about the study or even provides training relevant to the objective of the study via a web site.
-We discuss this as the "intervention" that expresses the practice that we wish developers to adopt.
-For this study we chose to make a nice looking graphic that users click on to see information about navigation commands, install the Sando search tool, and learn about the Blaze game.
+
+After selecting the practice to focus on, we needed to select an intervention method.
+The term intervention, indicates that we are seeking to influence developers to perform a practice previously defined by software engineering community.
+Intervention involves communication about the practice in a way that motivates adoption \cite{}.
+Interventions can include presentations, one-on-one discussions, distance education mechanisms such as webinars, and word of mouth recommendations.
+Part of the research objective is to determine whether we can acheive adoption of a practice with alternative interventions that are more deliberate than word of mouth and perhaps more scalable than webinars or presentations.
+Utilizing the automation capabilities of Blaze, we chose to automatically pop-up a window with training information on structured navigation practices and tools.
+For this study we chose to make a nice looking graphic that users click on to see information about navigation commands, obtain and install the Sando search tool, and learn about the Blaze game.
 The site uses a combination of video and blog-post style information easily accessible to the developer.
-The communication site is rated as part of our study completion survey.
+The usefulness of the communication site is rated as part of our study completion survey.
 
-Timing the study communication is part of the consideration for a good study design.
-We chose to deploy the Blaze tool without communicating the study purpose for a baseline period, then the tool would pop-up a display with the information on the communication site asking the developers to consume the information and begin using the practices.
-This allows us to measure changes in the developer's practices over time during the study.
-%Other intervention approaches may be used, however, you should take care to understand how the method and timing may affect the measurement of the practices.
+The study established three time periods, a prelinary period to establish a baseline, a period where information is communicated but no immediate feedback or points are displayed, and a period where points are displayed and all information is available.
+We chose to deploy the Blaze tool without communicating the study purpose for the baseline period so that an accurate picture of the baseline could be captured by the tool.
+Then the tool would pop-up a display with the information on the communication web site asking the developers to consume the information and begin using the practices.
+Finally the display of points fo accumulated by the developer provides the instant feedback mechanism.
 
-We conducted this study with developers working in an R\&D facility in India on a large industrial software system.
+We conducted this study with an intact team of developers working in an R\&D facility in India on a large industrial software system.
 Developers were asked to volunteer for the pilot by their management, however, managers did not know who participated in the study.
-Keeping managers out of the loop is intended to reduce concerns among the developers that the data may be used for purposes other then the intended study.
+We intended to keep managers out of the loop to reduce concerns among the developers that the data may be used for purposes other then the intended study.
 By selecting an intact team, we hope to leverage aspects of motivation such as the Leaderboard to spur feelings of competition among teammates.
@@
 They look for specific refactoring commands in Eclipse and determine the amount of time developers use tools versus hand refactoring the code.
+For this study we build a tool that captures the use of all commands within the Visual Studio environment including refactoring, navigation, and edit commands.
 
@@
 Key conclusions are that developers are more more successful finding and fixing bugs when they create a detailed plan for implementing a change, use structured navigation with keyword search or cross-reference search, and only review methods once during their search.
+This builds on this work by testing methods for increasing the use of structured navigation in developers everyday practice.
+
+Singer and Schneider demonstrate the use of a message board and points for encouraging students to increase their frequency of commits to the source code repository.\cite{Singer2012It}  The communication mechanism enabled students to see each others' progress and resulted in more frequent commits than baseline.
+Participants valued the communication and collaboration aspect and some valued the competition enough to change their opinion on optimum commit frequency.
+The subsequent thesis by Singer \cite{Singer2013a} describes results of an experiment conducted across two iterations of a class where active feedback on commits was deployed to one course and another course served as the control group where commit frequency was simply monitored.
+Results show an increase in the frequency of commits at a statistically significant level.
+This study uses finer grained instrumentation to identify practices being used between commits such as structured navigation.
+
+The Pro Metrics (PROM) tool provides a framework for collecting data for further analysis from tools used by developers.\cite{Coman2009Casestudy}  It provides a flexible data model and a plug-in architecture to facilitate collection from different data sources.
+Studies conducted using prom include a series of studies by it's creators on trends in time spent Pair Programming \cite{Coman2008Investigating}, benefits of refactoring on productivity \cite{Moser2008Case}, impact of refactoring on re-usability \cite{Moser2006Does}, and prediction of effort \cite{Abrahamsson2007Effort}.
+The PROM tool and studies that use it track the time spent editing files and methods from the IDE and code metrics obtained trough source code analysis.
+Applying these data to specific research questions such as whether refectoring improves productivity \cite{Moser2008Case} shows the utility of combining automated effort data with code metric data.
+Taking a different approach to focus on techniques, our study used more detailed events captured from the IDE  to detect how the user was navigating through the code in addition to knowing how long the editor is open for a particular file.
 
@@
 \balance
-\bibliography{paper_id1}  % sigproc.bib is the name of the Bibliography in this case
+\bibliography{paper_id1} 
+ % sigproc.bib is the name of the Bibliography in this case
 % You must have a proper ".bib" file

### 2013-09-02 paper_id1.tex
monday nearly done with related work and pre-study

@@
 
+Beecham and colleagues provide a thorough analysis of existing literature on studies of motivation in Software Engineering\cite{wbsnipes:Beecham2008Motivation}.
+It has some valuable lists of factors that motivate software engineers with the most common being "the work" motivates us.
+The list of de-motivators is also useful with common job satisfaction items like stress and inequity in recognition, plus poor quality software (low accomplishment) and lack of influence in decision making.
+The paper also lists characteristics of people in the professions including need for independence (autonomy), desire to learn new skills/tackle challenges.
+
+\section{Assess the Viability of Gamification for Software Engineering}
+
+\subsection{Assessment Background}
+When discussing gamification, folks raise a concern about replacing an activity that is intrinsically movitated with extrinsic motivation provided by points and achievements.
+To discover what is behind this concern, several studies are useful to understand what should be done.
+
+Early studies attempt to evaluate achievement motivation in students and relate that to their performance in school.
+The study by Hermans describes results from applying questionnaire to evaluate achievement motivation in students \cite{wbsnipes:Hermans1970Questionnaire}.
+Each question is ranked for its correlation to achievement motivation.
+Results indicate that some key questions about the diligence with which students approach their work are correlated with achievement motivation.
+The questions in this questionnaire and their ranking may be useful for evaluating achievement motivation in further studies.
+Take care when choosing appropriate questions that for the context of the survey.
+
+ii.	Determine through a questionnaire survey how developers are motivated to use training, practices, and tools and whether they are open to having that information shared with their peers and managers with recognition by badges and points.
+The developer study conducted with over 130 software developers at ABB challenged whether achievement motivation would help software engineers and whether they would accept sharing data on tools and practices they use to help others.
+
+\subsection{Assessment Results}
+
+Using a question from the Saachi and Saachi survey\cite{wbsnipes:SaatchiGameification} as directly as possible, we asked developers at ABB "How interested they would be in working for a company that incorporated some aspects of games into software engineering tools as a way to increase productivity in the workplace"?
+We segmented resuults by country and found that gamification of software engineering is more interesting to developers in India, Poland, and Finland than in other major countries.
+The responses indicating that developers were at least somewhat interested is 75-88\% In India, Poland, and Finland.
+For the US, Sweden, and Switzerland the responses indicate 60-65\% are at least somewhat interested.
+The overall ABB response is comparable to the Saachi and Saaschi survey\cite{wbsnipes:SaatchiGameification} upon which this question was based that found 75\% of 18-45 year olds were at least somewhat interested.
+
+95\% of the developers surveyed are likely to try tools and practices recommended to them through an automated usage tracking system.
+This indicates that a recommender system would positively impact the deployment of software engineering tools and practices at ABB.
+
+One area we questioned was whether sharing this information would be a concern for developers.
+A graduated scale was used for the community that would be shared with and distinction between whether that sharing was anonymous or the information was personally identifiable.
+Over 90\% of respondents are comfortable with sharing anonymous/non-anonymous information with their team members.
+The figure below shows how this comfort level decreases as the scope of who the information is shared with increases.
+People were less comfortable sharing with anyone (the category for people outside ABB) particularly if the data were not anonymous.
+Two conclusions from this feedback are that people are very willing to share information that could help their team.
+Second, sharing within the company is acceptable if we take care to make the data anonymous.
+
+Developers were asked what would motivate them to use tools and practices and given 4 choices to rank.
+The top motivator for using tools and practices is coworker recommendation with 75\% of respondents ranking in 1st or 2nd place.
+The lowest was having badges posted on their You@ABB profile where 33\% ranked this in 1st or 2nd place.
+Badges also ranked low in motivations for taking more software engineering training.
+We conclude that a facility to share coworker recommendations on tools and practices has the best opportunity to increase their adoption.
+Since a competition among teams ranks slightly higher than management mandate, we have the choice to apply one or the other or both as motivation for using key practices.
+
+Developers were asked to rank awards for different actions by which they would find the most desirable.
+Overall 68\% of developers ranked receiving an award for consistently performing quality practices as either 1st or 2nd place.
+This supports a conclusion that focusing on awarding quality practices would motivate developers to do them consistently.
+
+Based on the survey results we draw the following conclusions for this project going forward:
+•	95\% of developer responding would try tools and practices suggested by an automated recommender system particularly if they were provided in context with the current workflow pattern the developer follows.
+•	Responses indicate a preference for team competition over individual awards (badges) and awards for quality practices ranked highest.
+So, the tool will incorporate team competition ideas and recognize quality practices.
+•	Over 90\% of respondents approved of sharing data with their team members
+
 \section{Study Design}
@@
 Designing the study with continuous measurement but no control group presents a traditional study design with some challenges.
-We overcame this by dividing the study in to phases where the participants themselves form a control group.
+We overcame this by dividing the study into phases where the participants themselves form a control group.
 The first phase baselines their activities determining the current state of their practices.
@@
 \subsection{Setup the practice}
+
+Gamification motivates people to achieve goals through incremental scoring and discrete awards that are typically shared with peers in an on line community.
+Companies are gameifying their business processes and even public facing web sites to encourage people to complete a task such as their profile, stay loyal by earning rewards (air miles), or participate frequently by being part of a community.
+Aspects of Gamification that should be considered when applying this popular technique to a process or system include motivational alignment, whether to include the competitive dimension, and making achievements encouraging.
+
+First it is important to consider the existing motivation people may have for performing the activity.
+Applying gamefiication to activities that people are motivated intrinsically to do may decrease their motivation.
+For example applying gamification to education is very difficult because learners are typically intrinsically motivated more than extrinsically motivated to achieve in their education \cite{wbsnipes:Vansteenkiste2008Does}.
+In other words they are mostly interested in learning to better themselves rather than earn the highest GPA.
+If an activity is primarily extrinsically motivated then gamification is an easy fit to directly apply to the activity.
+
+Several studies support the use of games as a learning aid or as the core part of a training program.
+Naicker, Vincent and Adams \cite{wbsnipes:Amory1999Use} studied which type of computer game had the best response for enjoyment of play among undergraduate students.
+The authors concluded that adventure games provide the best foundation for development of teaching resources.
+Students rated sound, graphics, story line, and use of technology as important game elements.
+
 Selecting a practice to study involves defining specific characteristics of the practice and how capturing actions in Visual Studio relates to the practice.
@@
 
-After classifying the commands we assigned points to the commands in the same configuration file.
-Structured navigation commands received 1-10 points each while unstructured navigation commands received 0 points.
+Another factor to consider is the design of achievement awards and points.
+The design of achievement awards should gradually encourage the participant to reach higher and higher levels of success in the activity.
+Achievements must feel earned to the participant so that internally they recognize the effort to get the award was significant and feel satisfied  \cite{wbsnipes:Hamari2011Framework}.
+Hamari and Eranti describe a general relationship between achievements in games and regular game play  \cite{wbsnipes:Hamari2011Framework} .
+How those achievements are awarded to players, and the concept of levels of achievements.
+The article contains a warning that assigning achievements to tasks reduces intrinsic motivation because autonomy is reduced by the achievements.
+In games achievements are typically a parallel scoring system to the main game play.
+They make the game more engaging providing multiple ways to increase your score and multiple challenges in one interface.
+
+Using these guidelines we assigned points to structured navigation commands which received 1-10 points each while unstructured navigation commands received 0 points.
 Assigning negative points while possible, is discouraged as this may demotivate folks who do not have detailed knowledge of how the study is designed.
 When assigning points, make sure you recognize the relationship between points and levels such that "leveling up" happens at a frequent enough pace to make that motivating for users during the study.
-Ideally users would level up after a working day or two worth of activity initially extending to a week or even a month following an exponential curve \cite{}.
+Ideally users would level up after a working day or two worth of activity initially extending to a week or even a month following an exponential curve \cite{wbsnipes:Hamari2011Framework} making the game challenging for beginners and more experienced developers.
 Designing the points and levels such that above average developers can complete all levels during the study period is ideal.
@@
 %update and expand
-Researchers have developed systems that aim to answer specific questions using data from command logs.
-Zorro \cite{V:Johnson2007Automated} leverages the HackyStat data to determine if developers are adhering to test-driven development.
-The tool employs a similar approach dividing the work of the developer into episodes delineated by when test cases pass.
-The researchers define multiple types of episodes then determine whether the developer uses Test-Driven-Development or not in each episode.
+
+Johnson and Kou defined Zorro\cite{V:Johnson2007Automated}, a system for detecting whether developers use Test Driven Development techniques based on data from Hackystat.
+Zorro divides development activities into episodes delimited by events such as configuration management code check-in, start of a unit test run, or start of a build.
+Using the distinct events that developers follow within these episodes, Zorro determines whether the episode followed Test-Driven Development practices per their classic definition of test first - code - test pass, or a different scenario.
+In two student-based studies comparing Zorro classifications with a simultaneous observational screen video, Zorro achieved between 70\% \cite{Kou2010Operational} and 89\% \cite{V:Johnson2007Automated} accuracy in classifying episodes into their proper TDD scenarios.
+The studies did not, however, attempt to use or evaluate the influence of instant feedback aimed at encouraging participants to follow the classic definition of Test Driven Development.
+In our study, we attempt this leap to present data collected directly to developers attempting to influcence them to change their practices.
 
@@
 Applying these data to specific research questions such as whether refectoring improves productivity \cite{Moser2008Case} shows the utility of combining automated effort data with code metric data.
-Taking a different approach to focus on techniques, our study used more detailed events captured from the IDE  to detect how the user was navigating through the code in addition to knowing how long the editor is open for a particular file.
+Taking a different focus on techniques, our study used more detailed events captured from the IDE  to detect how the user was navigating through the code in addition to knowing how long the editor is open for a particular file.
+
+Murphy-Hill et.
+al.
+study a large usage history data set and apply several different algorithms to accurately suggest to users new commands in their environment to use.\cite{MurphyHill2012Improving} Algorithms based on command history performed well particularly when synchronized chronologically with the recipient's usage history.
+Novice users were better served by recommender algorithms that did not require a long history, where more expert users benefited from more sophisticated algorithms that included more history.
+The "Most Widely Used" algorithm that recommends commands based on the collective usage profile of the team performed nearly as well as the more sophisticated algorithms.
+The paper presents the methodology for making recommendations and the evaluation of the accuracy of the recommendations.
+Our work goes the next step of evaluating whether a specific recommended practice gets adopted by developers when influenced through gamification methods.
 

### 2013-09-02 paper_id1.tex
Roughly 90% complete on assessment

@@
 This personal connection limits the knowledge transfer to the co-located circle of colleagues connected with the early adopters.
-%update
-Finding a way to automate the communication of technique recommendations between developers enhances deployment of best practices in a globally distributed software development organization.   
-%PF the following sounds redundant
-To best match the peer recommendation scenario described above, the method should mimic the personal recommendation model while rewarding successful adopters.
-
-Beecham and colleagues provide a thorough analysis of existing literature on studies of motivation in Software Engineering\cite{wbsnipes:Beecham2008Motivation}.
-It has some valuable lists of factors that motivate software engineers with the most common being "the work" motivates us.
-The list of de-motivators is also useful with common job satisfaction items like stress and inequity in recognition, plus poor quality software (low accomplishment) and lack of influence in decision making.
-The paper also lists characteristics of people in the professions including need for independence (autonomy), desire to learn new skills/tackle challenges.
 
@@
 \subsection{Assessment Background}
-When discussing gamification, folks raise a concern about replacing an activity that is intrinsically movitated with extrinsic motivation provided by points and achievements.
+When discussing gamification, folks raise a concern about replacing an activity that is intrinsically motivated with extrinsic motivation provided by points and achievements.
 To discover what is behind this concern, several studies are useful to understand what should be done.
@@
 
-ii.	Determine through a questionnaire survey how developers are motivated to use training, practices, and tools and whether they are open to having that information shared with their peers and managers with recognition by badges and points.
+Maehr proposes an affirmative theory that individuals achieve as a member of a social group choosing the behavior that will meet the expectations and values of the group that is significant to them.
+\cite{wbsnipes:MaehrCulture}  So the presence of a social group where the person is motivated to belong and excel within is important for creating achievement motivation.
+Maehr states that since achievement behaviors can be triggered by circumstance, that they are present in most people regardless of background.
+
+Beecham and colleagues provide a thorough analysis of existing literature on studies of motivation in Software Engineering\cite{wbsnipes:Beecham2008Motivation}.
+It has some valuable lists of factors that motivate software engineers with the most common being "the work" motivates us.
+The list of de-motivators is also useful with common job satisfaction items like stress and inequity in recognition, plus poor quality software (low accomplishment) and lack of influence in decision making.
+The paper also lists characteristics of people in the professions including need for independence (autonomy), desire to learn new skills/tackle challenges.
+
+Understanding the concerns with maintaining intrinsic motivation while adding elements of achievement motivation, we developed an assessment survey questionnaire to determine whether the community would be receptive to a gamification approach to software development.
+We designed  the survey to assess how developers are motivated to use training, practices, and tools and whether they are open to having that information shared with their peers and managers, and being recognized with badges and points.
 The developer study conducted with over 130 software developers at ABB challenged whether achievement motivation would help software engineers and whether they would accept sharing data on tools and practices they use to help others.
@@
 Using a question from the Saachi and Saachi survey\cite{wbsnipes:SaatchiGameification} as directly as possible, we asked developers at ABB "How interested they would be in working for a company that incorporated some aspects of games into software engineering tools as a way to increase productivity in the workplace"?
-We segmented resuults by country and found that gamification of software engineering is more interesting to developers in India, Poland, and Finland than in other major countries.
+We segmented results by country in Figure \ref{fig:gamification} and found that gamification of software engineering is more interesting to developers in India, Poland, and Finland than in other major countries.
 The responses indicating that developers were at least somewhat interested is 75-88\% In India, Poland, and Finland.
@@
 
-95\% of the developers surveyed are likely to try tools and practices recommended to them through an automated usage tracking system.
-This indicates that a recommender system would positively impact the deployment of software engineering tools and practices at ABB.
+\begin{figure}\begin{mdframed}[linecolor=white]
+	\includegraphics[width=3in]{gamificationquestion.pdf}
+	\caption{Acceptance of gamification}
+	\label{fig:gamification}
+\end{mdframed}\end{figure}
 
-One area we questioned was whether sharing this information would be a concern for developers.
+We asked developers how likely they would be to try tools and practices recommended to them through an automated usage tracking system.
+Answers showed that 95\% of the developers surveyed are likely to try tools and practices recommended to them.
+This indicates that a recommendation system would positively impact the deployment of software engineering tools and practices at ABB.
+
+One area we questioned was whether sharing detailed usage information with colleagues and the company as a whole would be a concern for developers.
 A graduated scale was used for the community that would be shared with and distinction between whether that sharing was anonymous or the information was personally identifiable.
-Over 90\% of respondents are comfortable with sharing anonymous/non-anonymous information with their team members.
-The figure below shows how this comfort level decreases as the scope of who the information is shared with increases.
+Figure \ref{fig:comfortwithsharing}  shows that over 90\% of respondents are comfortable with sharing either anonymous or non-anonymous information with their team members.
+Their comfort level decreases as the scope of who the information is shared with increases.
 People were less comfortable sharing with anyone (the category for people outside ABB) particularly if the data were not anonymous.
@@
 
+\begin{figure}\begin{mdframed}[linecolor=white]
+	\includegraphics[width=3in]{ComfortWithSharing.pdf}
+	\caption{Comfortable with sharing usage information}
+	\label{fig:comfortwithsharing}
+\end{mdframed}\end{figure}
+
 Developers were asked what would motivate them to use tools and practices and given 4 choices to rank.
-The top motivator for using tools and practices is coworker recommendation with 75\% of respondents ranking in 1st or 2nd place.
-The lowest was having badges posted on their You@ABB profile where 33\% ranked this in 1st or 2nd place.
+The top motivator for using tools and practices is coworker recommendation with 75\% of respondents in Figure \ref{fig:toolandpracticemotivators}   ranking in 1st or 2nd place.
+The lowest was having badges posted on their social profile where 33\% ranked this in 1st or 2nd place.
 Badges also ranked low in motivations for taking more software engineering training.
@@
 Since a competition among teams ranks slightly higher than management mandate, we have the choice to apply one or the other or both as motivation for using key practices.
+
+\begin{figure}\begin{mdframed}[linecolor=white]
+	\includegraphics[width=3in]{ToolAndPracticeMotivators.pdf}
+	\caption{Motivation for using tools and practices}
+	\label{fig:toolandpracticemotivators}
+\end{mdframed}\end{figure}
 
@@
 
-Based on the survey results we draw the following conclusions for this project going forward:
-•	95\% of developer responding would try tools and practices suggested by an automated recommender system particularly if they were provided in context with the current workflow pattern the developer follows.
+Based on the survey results we draw the following conclusions:
+•	95\% of developer responding would try tools and practices suggested by an automated recommendation system particularly if they were provided in context with the current workflow pattern the developer follows.
 •	Responses indicate a preference for team competition over individual awards (badges) and awards for quality practices ranked highest.
@@
 Gamification motivates people to achieve goals through incremental scoring and discrete awards that are typically shared with peers in an on line community.
-Companies are gameifying their business processes and even public facing web sites to encourage people to complete a task such as their profile, stay loyal by earning rewards (air miles), or participate frequently by being part of a community.
+Companies are gamifying their business processes and even public facing web sites to encourage people to complete a task such as their profile, stay loyal by earning rewards (air miles), or participate frequently by being part of a community.
 Aspects of Gamification that should be considered when applying this popular technique to a process or system include motivational alignment, whether to include the competitive dimension, and making achievements encouraging.
@@
 First it is important to consider the existing motivation people may have for performing the activity.
-Applying gamefiication to activities that people are motivated intrinsically to do may decrease their motivation.
+Applying gamification to activities that people are motivated intrinsically to do may decrease their motivation.
 For example applying gamification to education is very difficult because learners are typically intrinsically motivated more than extrinsically motivated to achieve in their education \cite{wbsnipes:Vansteenkiste2008Does}.
@@
 Using these guidelines we assigned points to structured navigation commands which received 1-10 points each while unstructured navigation commands received 0 points.
-Assigning negative points while possible, is discouraged as this may demotivate folks who do not have detailed knowledge of how the study is designed.
-When assigning points, make sure you recognize the relationship between points and levels such that "leveling up" happens at a frequent enough pace to make that motivating for users during the study.
-Ideally users would level up after a working day or two worth of activity initially extending to a week or even a month following an exponential curve \cite{wbsnipes:Hamari2011Framework} making the game challenging for beginners and more experienced developers.
-Designing the points and levels such that above average developers can complete all levels during the study period is ideal.
+Assigning negative points is discouraged as this may demotivate folks who do not have detailed knowledge of how the points are assigned.
+In creating levels, following the guidance from Hamari \cite{wbsnipes:Hamari2011Framework}, we built levels following an exponential curve.
+At first users level up after a working day or two worth of activity initially extending to a week or even a month following an exponential curve  making the game motivating for beginners and challenging for more experienced developers.
+We designed the points and levels such that above average developers can complete all levels during the study period.
 
@@
 Interventions can include presentations, one-on-one discussions, distance education mechanisms such as webinars, and word of mouth recommendations.
-Part of the research objective is to determine whether we can acheive adoption of a practice with alternative interventions that are more deliberate than word of mouth and perhaps more scalable than webinars or presentations.
+Part of the research objective is to determine whether we can achieve adoption of a practice with alternative interventions that are more deliberate than word of mouth and perhaps more scalable than webinars or presentations.
 Utilizing the automation capabilities of Blaze, we chose to automatically pop-up a window with training information on structured navigation practices and tools.
@@
 
-The study established three time periods, a prelinary period to establish a baseline, a period where information is communicated but no immediate feedback or points are displayed, and a period where points are displayed and all information is available.
+The study established three time periods, a preliminary period to establish a baseline, a period where information is communicated but no immediate feedback or points are displayed, and a period where points are displayed and all information is available.
 We chose to deploy the Blaze tool without communicating the study purpose for the baseline period so that an accurate picture of the baseline could be captured by the tool.
@@
 The studies did not, however, attempt to use or evaluate the influence of instant feedback aimed at encouraging participants to follow the classic definition of Test Driven Development.
-In our study, we attempt this leap to present data collected directly to developers attempting to influcence them to change their practices.
+In our study, we attempt this leap to present data collected directly to developers attempting to influence them to change their practices.
 
@@
 The PROM tool and studies that use it track the time spent editing files and methods from the IDE and code metrics obtained trough source code analysis.
-Applying these data to specific research questions such as whether refectoring improves productivity \cite{Moser2008Case} shows the utility of combining automated effort data with code metric data.
+Applying these data to specific research questions such as whether refactoring improves productivity \cite{Moser2008Case} shows the utility of combining automated effort data with code metric data.
 Taking a different focus on techniques, our study used more detailed events captured from the IDE  to detect how the user was navigating through the code in addition to knowing how long the editor is open for a particular file.
@@
 study a large usage history data set and apply several different algorithms to accurately suggest to users new commands in their environment to use.\cite{MurphyHill2012Improving} Algorithms based on command history performed well particularly when synchronized chronologically with the recipient's usage history.
-Novice users were better served by recommender algorithms that did not require a long history, where more expert users benefited from more sophisticated algorithms that included more history.
+Novice users were better served by recommendation algorithms that did not require a long history, where more expert users benefited from more sophisticated algorithms that included more history.
 The "Most Widely Used" algorithm that recommends commands based on the collective usage profile of the team performed nearly as well as the more sophisticated algorithms.

### 2013-09-04 paper_id1.tex
progress is slow but steady

@@
 
-\title{Effectiveness of Instant Feedback on Developer Navigation Practices}
-\subtitle{Does Instant Feedback Make Us Better Developers?}
+\title{Applying gamification to developer practices}
+\subtitle{Why developer habits die hard}
 % 
@@
 
-Based on the survey results we draw the following conclusions:
+Based on the survey results we made the following conclusions:
 •	95\% of developer responding would try tools and practices suggested by an automated recommendation system particularly if they were provided in context with the current workflow pattern the developer follows.
@@
 \section{Study Design}
-Creating a longitudinal study involves defining a long-term goal that is measurable in increments, determining the discrete steps that the study will use to communicate the goal to the participants, distribute interventions to participants, and track participants progress over time.
-Many software engineering studies are conducted at discrete points in time via surveys of sampled participants.
-Perhaps prior to an intervention, and multiple points following an intervention.
-This study demonstrates a method to perform an intervention, continuously monitor the progress of participant response to that intervention, and provide continuous feedback to the participant and researchers about the progress.
-Continuous meaning that no information gaps exist in the data collection, there are no periods of unmonitored activity by the participant.
-Continuous also applies to the intervention being evaluated, where the participant and researchers receive constant feedback about the conformance to the intervention's purpose.
-Studying an intervention in a continuous context determines how sticky the intervention is in the mind of the developer.
-It can remove bias that might be introduced by periodic surveys or audits and enhance the habits formed in the participant receiving the intervention.
 
-Designing the study with continuous measurement but no control group presents a traditional study design with some challenges.
-We overcame this by dividing the study into phases where the participants themselves form a control group.
-The first phase baselines their activities determining the current state of their practices.
-The second phase presents the intervention to the participants.
-This phase the participants do not receive continuous feedback on their compliance with the intervention.
-This phase simulates the traditional approach of communicating an intervention then returning with an audit to determine whether participants are complaint with it.
-The last phases turns on continuous feedback to the participants on their compliance with the intervention.
-As they receive the feedback, the participants can determine for themselves how much they are implementing the intervention.
-In the last phase they possess complete knowledge of the intervention and their compliance with it making it plausible that they can continuously control their compliance.
+Utilizing the unobtrusive continuous monitoring capabilities of Blaze, we designed the pilot as a longtiudinal study in three waves.
+Longtiudinal stuides allow us to design an experiement and determine the effects over two or more sample periods using standard MANOVA (Multiple ANalysis of VAriance) models.
+\cite{longitudinalbook}  We employed the three wave model defining the first wave as a baseline capturing the current practices of the pilot group.
+The second wave employed the intervention that included an informative web site on structured navigation commands in Visual Studio, information about the sando search tool, and information about how points would be awarded during the pilot.
+The third wave provided the participants with continuous feedback in the from of points accumulated and an indicator arrow \ref{fig:blazegui} that points up when their navigation ratio is higher than their historical average.
 
@@
 \subsection{Setup the practice}
-
-Gamification motivates people to achieve goals through incremental scoring and discrete awards that are typically shared with peers in an on line community.
-Companies are gamifying their business processes and even public facing web sites to encourage people to complete a task such as their profile, stay loyal by earning rewards (air miles), or participate frequently by being part of a community.
-Aspects of Gamification that should be considered when applying this popular technique to a process or system include motivational alignment, whether to include the competitive dimension, and making achievements encouraging.
-
-First it is important to consider the existing motivation people may have for performing the activity.
-Applying gamification to activities that people are motivated intrinsically to do may decrease their motivation.
-For example applying gamification to education is very difficult because learners are typically intrinsically motivated more than extrinsically motivated to achieve in their education \cite{wbsnipes:Vansteenkiste2008Does}.
-In other words they are mostly interested in learning to better themselves rather than earn the highest GPA.
-If an activity is primarily extrinsically motivated then gamification is an easy fit to directly apply to the activity.
-
-Several studies support the use of games as a learning aid or as the core part of a training program.
-Naicker, Vincent and Adams \cite{wbsnipes:Amory1999Use} studied which type of computer game had the best response for enjoyment of play among undergraduate students.
-The authors concluded that adventure games provide the best foundation for development of teaching resources.
-Students rated sound, graphics, story line, and use of technology as important game elements.
 
@@
 Using these guidelines we assigned points to structured navigation commands which received 1-10 points each while unstructured navigation commands received 0 points.
-Assigning negative points is discouraged as this may demotivate folks who do not have detailed knowledge of how the points are assigned.
-In creating levels, following the guidance from Hamari \cite{wbsnipes:Hamari2011Framework}, we built levels following an exponential curve.
-At first users level up after a working day or two worth of activity initially extending to a week or even a month following an exponential curve  making the game motivating for beginners and challenging for more experienced developers.
+Assigning negative points is discouraged as this may demotivate folks who do not have detailed knowledge of how the points are assigned.\cite{}   In creating levels, following the guidance from Hamari \cite{wbsnipes:Hamari2011Framework}, we built levels following an exponential curve.
+Users initially level up after a working day or two extending to achieving the next level after a week or even a month following an exponential curve.
+The leveling up points make the game motivating for beginners and challenging for more experienced developers.
 We designed the points and levels such that above average developers can complete all levels during the study period.
@@
 After selecting the practice to focus on, we needed to select an intervention method.
-The term intervention, indicates that we are seeking to influence developers to perform a practice previously defined by software engineering community.
+The term intervention, indicates that we are applying a technique to influence a change in the measured behavior of developers.
 Intervention involves communication about the practice in a way that motivates adoption \cite{}.
@@
 Part of the research objective is to determine whether we can achieve adoption of a practice with alternative interventions that are more deliberate than word of mouth and perhaps more scalable than webinars or presentations.
-Utilizing the automation capabilities of Blaze, we chose to automatically pop-up a window with training information on structured navigation practices and tools.
-For this study we chose to make a nice looking graphic that users click on to see information about navigation commands, obtain and install the Sando search tool, and learn about the Blaze game.
-The site uses a combination of video and blog-post style information easily accessible to the developer.
+
+In the initial wave, participants were informed of the purpose of the study what to expect from the Blaze tool.
+The participants installed the tool in their Visual Studio environment.
+The tool did not present any window for them to see and simply collected data in the background.
+
+After a week's data collection, Blaze started to automatically pop-up a window containing a button for a web page with information.
+The page included a link to download another tool for code search called Sando \ref{sando} and information about using that tool in a video.
+The sub-page on navigation contained a page of tutorial on the built-in structured navigation commands in Visual Studio.
+The About Blaze page contained details of the point scoring system and background on Blaze's purpose.
 The usefulness of the communication site is rated as part of our study completion survey.
-
-The study established three time periods, a preliminary period to establish a baseline, a period where information is communicated but no immediate feedback or points are displayed, and a period where points are displayed and all information is available.
-We chose to deploy the Blaze tool without communicating the study purpose for the baseline period so that an accurate picture of the baseline could be captured by the tool.
-Then the tool would pop-up a display with the information on the communication web site asking the developers to consume the information and begin using the practices.
-Finally the display of points fo accumulated by the developer provides the instant feedback mechanism.
 
@@
 Results show an increase in the frequency of commits at a statistically significant level.
-This study uses finer grained instrumentation to identify practices being used between commits such as structured navigation.
+This study uses fine grained instrumentation to identify practices used between commits such as structured navigation.
 
@@
 
-Building on this success, this work proposes a generalized method to automate the identification of the techniques described in the related work and others, and determine when the techniques would help a developer improve their work patterns.
-Then we intend to create a game-like virtual environment where developers learn from each other and are motivated to adopt the best practices developed by these and other researchers.
+Dubois and Tamburrelli discuss theoretical concepts behind applying gamification to software development.
+\cite{Dubois2013Understanding} Their approach is to consider how gamification could be applied throughout the software development life-cycle to motivate many practices in different phases.
+They provide guidance on assessing gamification in the software development domain that includes analyzing the actors and game mechanics, defining integration with existing tools and processes, and designing the evaluation of results.
+Dbois and Tamburrelli demonstrate this guidance using a study of students working on a class project.
+Defining a control group and subject group of students, both groups receive the same feedback about their code comments, test coverage, code size, and static analysis rule violations.
+The subject group can see each others' rankings while the control group only sees their personal rankings.
+Results though inconclusive indicate that competition increases the desired results.
+In our study we apply both individual feedback and competitive based feedback (anonymized) to participants and seek to observe a change in their behavior after an intervention.
+
+\section{Results}
+The following hypotheses interest us in the current study.  
+1. Developers using blaze use more structured navigtion when they receive instant feedback about their practices
+2. Developers find the competitive aspect of blaze encourages them to pay more attention to navigation practices
+
+To evaluate whether developers use more structured navigation when they receive instant feedback, we monitor the change in navigation ratio over the pilot study period.
+Expecting that as developers learn about navigation commands and tools that count towards the score in the second week they will begin using the practices more.
+Also expecting that noticing the leaderboard competition will encourage some developers to use the navigation practices more.
+What we observed is shown in Figure \ref{fig:navigationaverage} that the structured navigation practice did not increase significantly from the beginning to the end of the study or between weeks when the additional information was revealed to them.
+
+\begin{figure}\begin{mdframed}[linecolor=white]
+	\includegraphics[width=3in]{navigationaverage.pdf}
+	\caption{Average Navigation Ratio by Week by Developer}
+	\label{fig:navigationaverage}
+\end{mdframed}\end{figure}
+
+\section{Threats to Validity}
+There are several threats to validity for this result.
+By making the purpose of our measurement and contest known although without details we may have triggered the participants to think about navigation commands they use before establishing a baseline.
+We attempt to control for this by comparing the pilot participants' data to data from Blaze users who are knowledgeable Visual Studio developers but did not receive the interventions.
+Another threat is the intact team represents people with different backgrounds and experience levels who may have established habits for navigating through source code.
+We attempt to control for this by asking post survey questions of the participants about their familiarity with the code they work in and their experience levels.
+The threat of non- generalizable of results remains particularly because the pilot was conducted within an intact team in a single industrial site in India.
+As the pre-study questionnaire shows, the developer's desirability for a gamification type of intervention vary by country.
 

### 2013-09-05 paper_id1.tex
study design looking good

@@
 \alignauthor
-Will Snipes
+Will Snipes\\
 \affaddr{ABB Corporate Research}\\
@@
 \alignauthor
-Anil R. Nair
+Anil R. Nair\\
 \affaddr{ABB Corporate Research}\\
@@
 Utilizing the unobtrusive continuous monitoring capabilities of Blaze, we designed the pilot as a longtiudinal study in three waves.
-Longtiudinal stuides allow us to design an experiement and determine the effects over two or more sample periods using standard MANOVA (Multiple ANalysis of VAriance) models.
+Longtiudinal studies allow us to design an experiment and determine the effects over two or more sample periods using standard MANOVA (Multiple ANalysis of VAriance) models.
 \cite{longitudinalbook}  We employed the three wave model defining the first wave as a baseline capturing the current practices of the pilot group.
-The second wave employed the intervention that included an informative web site on structured navigation commands in Visual Studio, information about the sando search tool, and information about how points would be awarded during the pilot.
+The second wave employed the intervention that included an informative web site on structured navigation commands in Visual Studio, information about the Sando search tool, and information about how points would be awarded during the pilot.
 The third wave provided the participants with continuous feedback in the from of points accumulated and an indicator arrow \ref{fig:blazegui} that points up when their navigation ratio is higher than their historical average.
@@
 
-\subsection{Setup the practice}
+\subsection{Select a Practice}
+In their paper on game design patterns, Hamari et.
+al.
+issue a warning that assigning achievements to required tasks reduces intrinsic motivation because player autonomy is reduced by the achievements \cite{wbsnipes:Hamari2011Framework} .
+This warning indicates that we should avoid giving points for things that are already an expected deliverable such as finding or fixing bugs, completing development tasks, or meeting required process criteria.
+Practices to consider are those practices with room to go beyond a checklist or minimum criteria where developers have the opportunity to excel.
 
-Selecting a practice to study involves defining specific characteristics of the practice and how capturing actions in Visual Studio relates to the practice.
-In this study we selected the structured navigation practice shown by Robillard et.
+Selecting a practice to study involves identifying a practice, defining specific characteristics of the practice to measure, and identifying the actions in Visual Studio that relate to the practice.
+We currently limit ourselves to practices that exist entirely in the IDE thus can be monitored using Blaze.
+We identified potential practices of test-driven development, debugging practices, navigation practices,  eliminating static analysis bugs, and check-in frequency.
+Practices of check-in frequency, static analysis bug elimination are highlighted in this paper's related work section with prior gamification studies in classroom environments.
+Johnson and Kou achieved automated monitoring of test-driven development practices with Zorro thus may provide a candidate for future gamification application.
+Debugging practices have developer community opinions on optimal methods, however, detailed debug practices with modern IDEs have not received research community attention, thus are also reserved for future work.
+Based on studies of structured navigation by Robillard et.
 al.
-as correlated with effective program maintenance \cite{wbsnipes:Robillard2004How} .
-Studying structured navigation is a natural fit for capturing commands issued in the IDE.
-The first step is to create the classification of commands that are involved in navigation.
-Then differentiate which commands we consider as structured navigation from unstructured navigation.
+that identify structured navigation as a practice leading to effective program maintenance  \cite{wbsnipes:Robillard2004How}, we identified structured navigation as a key practice.
+Structured navigation fits the requirement of a practice that is contained in the IDE, is something that has proven benefits, and lends itself to measurement and scoring, is not an assigned task, and leaves room for developers to excel.
 
-The Blaze tool provides an XML configuration file that allows defining multiple levels of categories for commands thus we can categorize commands that are navigation then further classify them into structured and unstructured navigation.
-We classified commands as structured navigation when they allow the developer to locate a code element following the program structure from another element.
-For example, Navigate To \textit{Ctrl+,}, Go To Definition \textit{F12}, and Find All References \textit{Ctrl+K,R} are commands that follow structure of the code that we classify as structured navigation.
+\subsection{Design the Game}
+
+To setup the game for a practice, the Blaze tool provides an XML configuration file where the researcher can configure Blaze to categorize and assign points to the commands that are part of a software enginering practice.
+Blaze allows the researcher to define multiple command category levels.
+Thus we can categorize commands that are navigation then further classify them into structured and unstructured navigation.
+We categorize commands as structured navigation when they allow the developer to locate a code element following the program structure from another element.
+For example, Navigate To \textit{Ctrl+,}, Go To Definition \textit{F12}, and Find All References \textit{Ctrl+K,R} are commands that follow structure of the code, which we classify as structured navigation.
 Commands that we considered as unstructured navigation include scrolling through a file using keys or the scroll-bar, opening files from the Solution Explorer, Find commands (e.g.
@@
 
-Another factor to consider is the design of achievement awards and points.
+Gamification practices consider the feedback of achievements and points as critical to a successful outcome.
 The design of achievement awards should gradually encourage the participant to reach higher and higher levels of success in the activity.
 Achievements must feel earned to the participant so that internally they recognize the effort to get the award was significant and feel satisfied  \cite{wbsnipes:Hamari2011Framework}.
-Hamari and Eranti describe a general relationship between achievements in games and regular game play  \cite{wbsnipes:Hamari2011Framework} .
-How those achievements are awarded to players, and the concept of levels of achievements.
-The article contains a warning that assigning achievements to tasks reduces intrinsic motivation because autonomy is reduced by the achievements.
-In games achievements are typically a parallel scoring system to the main game play.
+Hamari and Eranti describe a general relationship between achievements in games and regular game play      In games achievements are typically a parallel scoring system to the main game play.
 They make the game more engaging providing multiple ways to increase your score and multiple challenges in one interface.
@@
 Users initially level up after a working day or two extending to achieving the next level after a week or even a month following an exponential curve.
-The leveling up points make the game motivating for beginners and challenging for more experienced developers.
+The levels make the game motivating for beginners and challenging for more experienced developers.
 We designed the points and levels such that above average developers can complete all levels during the study period.
 
-\subsection{Study Intervention and Participants}
+\subsection{Intervention Staging}
 
@@
 
-After a week's data collection, Blaze started to automatically pop-up a window containing a button for a web page with information.
+After a week's data collection marking the end of the first wave, Blaze started to automatically pop-up a window containing a button for a web page with information.
 The page included a link to download another tool for code search called Sando \ref{sando} and information about using that tool in a video.
@@
 
+In the third wave developers received instant feedback from Blaze on the points they accumulated during the development session.
+The up/down arrow provided a simple indicator on whether they were increasing their use of structured navigation or were using it less than their personal historical average.
+This final configuration continued until the end of the pilot.
+
+\subsection{Participants}
+
 We conducted this study with an intact team of developers working in an R\&D facility in India on a large industrial software system.
 Developers were asked to volunteer for the pilot by their management, however, managers did not know who participated in the study.
-We intended to keep managers out of the loop to reduce concerns among the developers that the data may be used for purposes other then the intended study.
+Mechanisms in our data collection to identify individuals were under the control of each developer, they could chose to share or keep confidential the association of a generated unique id with researchers or each other.
+
+Management wished to have no knowledge of who was participating and took great care to remain detached from the pilot proceedings.
+Data at this level of detail has not been collected in ABB development organizations except for a few minutes of video study that might be conducted.
+Researchers were careful to create the system so that developers controlled their identification with the data and fortunately, management bought into the necessary anonymity of the participants.
+
+\subsection{Competitive Elements}
 By selecting an intact team, we hope to leverage aspects of motivation such as the Leaderboard to spur feelings of competition among teammates.
+The leaderboard was presented so that developers could see their own score in relation to the top 5 people on the leaderboard.
+However, the identifier of the other participants was generated as 3 capital letters assigned based on their position on the leaderboard.
+Therefore the assigned letters could even change when an individual's ranking changed.
+Not knowing who of your colleagues was ahead of you could be a factor in motivation, but also avoids one potential source of conflict in an intact team.
+
+Competitive dynamics may be more effective when teams from different areas can compete as a team against other teams..
+This tried and true method both builds esprit de corps and enhances achievement towards the intended outcome.
+This study design is potential future work.
 

### 2013-09-07 paper_id1.tex
not sure what changed today

@@
 
-\section{Study Design}
+\section{Study and Game Design}
 
@@
 
-To setup the game for a practice, the Blaze tool provides an XML configuration file where the researcher can configure Blaze to categorize and assign points to the commands that are part of a software enginering practice.
+To setup the game for a practice, the Blaze tool provides an XML configuration file where the researcher can configure Blaze to categorize and assign points to the commands that are part of a software engineering practice.
 Blaze allows the researcher to define multiple command category levels.
@@
 \section{Results}
+
+Although we translated navigation commands to points for developers, we defined the metric Navigation ratio based on Jones and Jones \cite[] as the number of structured navigation events in a day over the number of unstructured navigation events.
+This metric provides a continuous variable allowing us to use statistical data of continuous data to check our hypotheses.
+To create the evaluation for each wave, we took the navigation ratio for each developer in each wave.
+
 The following hypotheses interest us in the current study.  
-1. Developers using blaze use more structured navigtion when they receive instant feedback about their practices
+1. Developers using blaze use more structured navigation when they receive instant feedback about their practices
 2. Developers find the competitive aspect of blaze encourages them to pay more attention to navigation practices
@@
 Also expecting that noticing the leaderboard competition will encourage some developers to use the navigation practices more.
-What we observed is shown in Figure \ref{fig:navigationaverage} that the structured navigation practice did not increase significantly from the beginning to the end of the study or between weeks when the additional information was revealed to them.
+What we observed is shown in Figure \ref{fig:navigationaverage} that the structured navigation practice did not increase significantly from the beginning to the end of the study.
+Overall, we did not see a significant effect towards increasing navigation ratio for the interventions in wave 2 or wave 3.
 
@@
 \end{mdframed}\end{figure}
+
+Is this a spectacular fail from a gamification effectiveness perspective?
+How could this be when pre-study results indicated very positive attitudes towards the approach?
+Our post pilot survey attempts to shed some light on the reasons.
+One question in the post survey is how knowledgeable the developer was about the code they worked in with 3 possibilities,  they were authors of the code, they regularly maintained the code, they were unfamiliar with the code.
+In this team, x\% of developers cited that they ...the code.
+Another factor to consider is experience level which we also asked in the post-pilot survey.
+More experienced developers may be more convinced of their effective habits and perhaps react almost with muscle memory to use structured or unstructured navigation to locate code of interest.
+Even developers whose navigation ratio  was lower than average, used structured navigation occasionally as much as their more frequently using colleagues.
+The variation for each individual can be seen in Figure \ref{} .
 

### 2013-09-09 paper_id1.tex
from the office

@@
 
-\subsection{Participants}
+\subsection{Pilot Participants}
 
-We conducted this study with an intact team of developers working in an R\&D facility in India on a large industrial software system.
+We conducted this study with an intact team of 6 developers working in an R\&D facility in India on a large industrial software system.
+We approached a team which develops and maintains an industrial software product.
+Initially, we met the whole team along with the management staff to explain the research project and the Blaze tool we developed.
+We  explained the aim of the study, what data is collected, and how the data will be processed and managed.
+
+It was important to gain the confidence of the participants that their data were kept confidential so their activities would be close as possible to real-world.
+if the participants felt like they are being tracked and the data is going to use for some study that would identify them to their peers or managers, then they may deviate from the natural development style.
+
+We  assured the team about the confidentiality of the data and the management also informed the team that the data collected by the research group will not be used for any other purpose other than the study.
 Developers were asked to volunteer for the pilot by their management, however, managers did not know who participated in the study.
+Their name and data were kept confidential to ensure that no one apart from the research team knows that he/she is participating in the pilot.
 Mechanisms in our data collection to identify individuals were under the control of each developer, they could chose to share or keep confidential the association of a generated unique id with researchers or each other.
@@
 Researchers were careful to create the system so that developers controlled their identification with the data and fortunately, management bought into the necessary anonymity of the participants.
+%Through this approach, we ensured that the developers are performing the task on a very natural environment with out bothering that they are being tracked.
 
@@
 This study design is potential future work.
+
+\subsection{Pilot Survey}
+
+The goal for the pilot follow-up survey was to collect feedback on the tool and the effect it had on the developers navigation practices and knowledge.
+The survey questions were segmented into portions that evaluate prior and post knowledge and practices for navigation, key steps in the pilot itself, the influence of game elements on their navigation practices, typical tasks and demographic information.
+
+The key questions about this pilot were did we have an effect on knowledge of the developers regarding structured navigation commands.
+We asked them to rate their knowledge of structured navigation commands prior to the pilot on a 1-10 scale to understand their baseline.
+We asked them to contrast their knowledge after the pilot to assess what they learned from the pilot.
+
+The other key question is did we have an effect on their navigation practices.
+Here too we asked them to rate how much they used structured navigation prior to the pilot using a 4 choice rating of not at all, some, many times, and as much as possible.
+Then we asked whether the used structured navigation commands about the same amount, more,  or a lot more during the pilot.
+These two questions assessed their intent in terms of using structured navigation commands and we checked that with the data.
+
+A key set of questions focused on whether using Blaze increased their attention to navigation commands by asking how much specific game elements influenced their practices.
+The scale started with not at all, a little, some, to a lot.
+
+The other questions included whether they reviewed the training materials on structured navigation commands, installed the Sando code search tool, and what types of development tasks they would consider using structured navigation for in general.
+A key demographic question is how well they know the code they work in rated from "I work in code that I wrote", "I maintain code that I am familiar with", to "I recently learned the code"
+
+Survey results are discussed in the results section \ref{Results}
 
@@
 
+\subsection{Pilot Survey Results}
+
+On the assessment of structured navigation knowledge, half of the participants replied that they had very good prior knowledge about structured navigation and they use structured navigation as much as possible.
+Other half of the participants have some understanding on structured navigation use it some of the time.
+While analyzing the points they scored with blaze tool, it is observed that the participants who mentioned that they have good knowledge in structured navigation are on top of the points table.
+%the following statement is not supported with data unless you have additional data?
+Incidentally, while cross verifying their performance on the coding, they are better productive compared to the participants on the bottom of points table.
+
+Assessing the game elements of Blaze that influenced them, four out of six said the leaderboard influenced them a lot and all rated the leaderboard as having some influence.
+Three out of six  said their individual point score influenced them a lot and all said the score had some influence.
+The element with mixed ratings was a graphical indicator on whether they were improving over past use of structured navigation commands.
+The indicator received the a lot rating from one participant and a not at all rating from one participant with the rest having some influence.
+
+The participants requested that the system to provide more feedback on tips to them to increase their score. This feature of the tool is planned for the next release.
+
+Comparing the survey demographic results with game scores reveals that years of experience did not have a strong influence over scores or structured navigation knowledge.
+All developers in the study had 7 or more years experience and one person with 8 years experience used structured navigation enough to obtain second place matching a 10 year experienced colleague.
+
+Those who rated their knowledge of structured navigation commands highly, did better in the points and generally had a higher average navigation ratio than those who rated their prior knowledge lower.
+This could indicate that using structure navigation is somewhat of a habit that takes longer to change.
+There was also feedback that people wanted more information presented to them by the tool on what commands to use, so they may not have closed the knowledge gap during the pilot.
+
+Those who worked in code that they wrote tended to use less structured navigation having a lower navigation ratio and point score.
+Those who had recently learned the code they were working in had more consistent use of structured navigation and one of them used Sando the most of the pilot participants.
+
 \section{Threats to Validity}

### 2013-09-09 paper_id1.tex
adjusted pivot table on nav ratio

@@
 The other questions included whether they reviewed the training materials on structured navigation commands, installed the Sando code search tool, and what types of development tasks they would consider using structured navigation for in general.
-A key demographic question is how well they know the code they work in rated from "I work in code that I wrote", "I maintain code that I am familiar with", to "I recently learned the code"
+Another demographic question is how well they know the code they work in rated from "I work in code that I wrote", "I maintain code that I am familiar with", to "I recently learned the code"
 
@@
 
-Those who rated their knowledge of structured navigation commands highly, did better in the points and generally had a higher average navigation ratio than those who rated their prior knowledge lower.
-This could indicate that using structure navigation is somewhat of a habit that takes longer to change.
+Those who rated their knowledge of structured navigation commands highly, did better in the points than those who rated their prior knowledge lower.
+This could indicate that using structured navigation is somewhat of a habit that takes longer to change.
 There was also feedback that people wanted more information presented to them by the tool on what commands to use, so they may not have closed the knowledge gap during the pilot.
 
-Those who worked in code that they wrote tended to use less structured navigation having a lower navigation ratio and point score.
+Those who worked in code that they wrote tended to use less structured navigation having a lower navigation ratio.
 Those who had recently learned the code they were working in had more consistent use of structured navigation and one of them used Sando the most of the pilot participants.

### 2013-09-09 paper_id1.tex
Completed draft of introduction and conclusion.  Worked towards refining results sections for pilot survey and pilot data.

@@
 In addition to training, discussing, and presenting software engineering practices and tools, we seek to communicate them through active two-way information embedded in the Integrated Development Environment (IDE).
-While communicating, we also collect data on new and current practices and tools used by each developer by recording actions in the IDE.
-With a rich data source, we evaluate what practices are used, define metrics that give individual feedback, and create an environment where developers can see how they are doing compared with their teammates.
 
-This approach helps understand practice and tool use over the long term and judges the lasting change a communication effort makes in everyday habits.
-Tool support provides a mechanism to embed into Visual Studio an extension that provides the capture of usage data while giving feedback to users on their practices both in general and specific to the ones being promoted.
-Through a points score system that assigns points to specific commands and tools we encourage everyone to use the new practices.
+We collect data on current practices and tools used by the developer while giving them feedback on how they can improve.
+
+ With a rich data source, we evaluate developer practices, define metrics that give individual feedback, and create an environment for developers to see how they are doing compared with their teammates.
+
+Blaze helps understand developer practice and tool use over the long term and judges the lasting change a communication effort makes in everyday habits.
+Embeding Blaze n the developers work environment as a Visual Studio extension provides the capture of usage data while giving feedback to users on their practices.
+Blaze encourages everyone to use the new practices by assigning points to specific commands and tools .
 Providing a light score ranking view should help us stay motivated as we compare our command usage to our peers.
-To demonstrate how this works, we implement a longitudinal study that demonstrates the results from this approach as we engage with developers over a month-long period.
+A month-long longitudinal study conducted with pilot users shows how develipers react to the presented information.
 
-Results from this demonstrative study show that in general the approach is <blank> with a few developers <blank> of the study.
+Results from this demonstrative study show the approach is <blank> with a few developers <blank> of the study.
 The effectiveness of this approach for communicating about practices is demonstrated through the continued use of the commands as we communicate them.
@@
 %update
-Deploying these techniques requires communicating and motivating developers to use them.
-Murphy-Hill and colleagues  \cite{wbsnipes:Hill2011Peer} show that developers can effectively learn about new tools from their peers by watching them work and through recommendations, yet this happens very rarely, in part because developers rarely work on the same computer at the same time.
-This personal connection limits the knowledge transfer to the co-located circle of colleagues connected with the early adopters.
+
+As  a developer I'm focused on the problem I'm currently solving and driven by productivity and deadlines to the point that I don't have time to sharpen the saw identifying better practices.
+
+Developers with years of experience develop habits and ways of doing things that are difficult to change even when they might recognize better methods.
+For instance, the commands they use and ways they use to navigate the code base in the development environment make them less efficient than using newer practices.
+Robillard et.
+al.
+show that developers who think of code as a graph and use commands that allow them to follow code structure are more effective and efficient than developers who browse through code.
+\cite{wbsnipes:Robillard2004How}
+
+Perhaps the persistence of their habit is why developers are hard to turn from their old practices.
+They may need some coaching to continue the new practice following a given lesson.
+To continuously coach the developer, we conceived of a tool in our prior work to provide constant feedback to retain the attention and interest of the developer.
+\cite{Snipes2013Towards} This tool creates a competitive game that rewards the developer continuously for adopting the improved practice.
+The more they change their habits, the higher they score in the game.
+
+Taking a step back, gamification is a popular movement that people apply to customer oriented web sites and social games.
+Why would software developers writing serious software be interested in this?
+Our first step was to challenge this question through a survey that queried both whether developers would be interested in gamifying their workspace and what aspects of that space would be most interesting to them.
+
+With these questions answered, we identified a set of practices and defined a game meeting expectations from the developers feedback.
+The game, implemented in a Visual Studio add-in called Blaze, integrates into the Visual Studio workspace and establishes a light anonymous competition between developers to improve their navigation practices.
+
+During the pilot of Blaze we observed the variation in developer navigation practices and identified some potential reasons for this variation.
+We evaluated developer's acceptance of game mechanics in the tool and their desires for additional functionality.
+While further data is required to assess the impact of the game, the initial developer acceptance encourages an expanded study.
 
@@
 Each question is ranked for its correlation to achievement motivation.
-Results indicate that some key questions about the diligence with which students approach their work are correlated with achievement motivation.
-The questions in this questionnaire and their ranking may be useful for evaluating achievement motivation in further studies.
-Take care when choosing appropriate questions that for the context of the survey.
+Some key questions about the diligence with which students approach their work are correlated with achievement motivation.
 
-Maehr proposes an affirmative theory that individuals achieve as a member of a social group choosing the behavior that will meet the expectations and values of the group that is significant to them.
-\cite{wbsnipes:MaehrCulture}  So the presence of a social group where the person is motivated to belong and excel within is important for creating achievement motivation.
+Maehr proposes an affirmative theory whereby  individuals achieve as a member of a social group choosing the behavior that will meet the expectations and values of the group that is significant to them.
+\cite{wbsnipes:MaehrCulture}  The presence of a social group where the person is motivated to belong and excel within is important for creating achievement motivation.
 Maehr states that since achievement behaviors can be triggered by circumstance, that they are present in most people regardless of background.
@@
 For the US, Sweden, and Switzerland the responses indicate 60-65\% are at least somewhat interested.
-The overall ABB response is comparable to the Saachi and Saaschi survey\cite{wbsnipes:SaatchiGameification} upon which this question was based that found 75\% of 18-45 year olds were at least somewhat interested.
+The overall response from our survey is comparable to the Saachi and Saaschi survey\cite{wbsnipes:SaatchiGameification} upon which this question was based that found 75\% of 18-45 year olds were at least somewhat interested.
 
@@
 We identified potential practices of test-driven development, debugging practices, navigation practices,  eliminating static analysis bugs, and check-in frequency.
+
 Practices of check-in frequency, static analysis bug elimination are highlighted in this paper's related work section with prior gamification studies in classroom environments.
@@
 Debugging practices have developer community opinions on optimal methods, however, detailed debug practices with modern IDEs have not received research community attention, thus are also reserved for future work.
+
 Based on studies of structured navigation by Robillard et.
@@
 Thus we can categorize commands that are navigation then further classify them into structured and unstructured navigation.
+
 We categorize commands as structured navigation when they allow the developer to locate a code element following the program structure from another element.
 For example, Navigate To \textit{Ctrl+,}, Go To Definition \textit{F12}, and Find All References \textit{Ctrl+K,R} are commands that follow structure of the code, which we classify as structured navigation.
+
 Commands that we considered as unstructured navigation include scrolling through a file using keys or the scroll-bar, opening files from the Solution Explorer, Find commands (e.g.
@@
 
-Although we translated navigation commands to points for developers, we defined the metric Navigation ratio based on Jones and Jones \cite[] as the number of structured navigation events in a day over the number of unstructured navigation events.
+\subsection{Pilot Data Observations}
+
+\subsubsection{Observations about navigation ratio improvement}
+
+To evaluate whether developers improved their practices for structured navigation we established a metric Navigation ratio as the number as the number of structured navigation events in a day over the number of unstructured navigation events.
+
+Structured navigation events included:
+
+Unstructured navigation events included:
+
 This metric provides a continuous variable allowing us to use statistical data of continuous data to check our hypotheses.
@@
 
-The following hypotheses interest us in the current study.  
-1. Developers using blaze use more structured navigation when they receive instant feedback about their practices
-2. Developers find the competitive aspect of blaze encourages them to pay more attention to navigation practices
+%The following hypotheses interest us in the current study.  
+%1. Developers using blaze use more structured navigation when they receive instant feedback about their practices
+%2. Developers find the competitive aspect of blaze encourages them to pay more attention to navigation practices
 
-To evaluate whether developers use more structured navigation when they receive instant feedback, we monitor the change in navigation ratio over the pilot study period.
+%To evaluate whether developers use more structured navigation when they receive instant feedback, we monitor the change in navigation ratio over the pilot study period.
 Expecting that as developers learn about navigation commands and tools that count towards the score in the second week they will begin using the practices more.
@@
 
-Is this a spectacular fail from a gamification effectiveness perspective?
-How could this be when pre-study results indicated very positive attitudes towards the approach?
-Our post pilot survey attempts to shed some light on the reasons.
-One question in the post survey is how knowledgeable the developer was about the code they worked in with 3 possibilities,  they were authors of the code, they regularly maintained the code, they were unfamiliar with the code.
-In this team, x\% of developers cited that they ...the code.
-Another factor to consider is experience level which we also asked in the post-pilot survey.
-More experienced developers may be more convinced of their effective habits and perhaps react almost with muscle memory to use structured or unstructured navigation to locate code of interest.
+\subsubsection{Observations about efficiency of structured navigation}
+
+Structured navigation sequences are less frequent between edits than unstructured navigation.
+To study navigation practices in relationship to edits, we grouped the data into sequences delimited by edits meaning each sequence ends with one or more edits to the source code.
+We evaluated the number of navigation commands between edits including clicks on locations in the file, navigating between files, and scrolling the file in the window.
+
+We found that in x\% of the edit sequences structured navigation was sued.
+We compared the time between edits where the sequences involved structured navigation to those that involved unstructured navigation.
+The data supports the developer's perception that unstructured navigation is faster than structured navigation particularly when search is used.
+We did find that when navigating between class files, developers who use structured navigation reached their destination X\% faster/slower than when using unstructured navigation.
+
+Search was instrumental for developers who used the search tool where occurrences including search were completed x\% faster or slower than navigating by unstructured commands.
+
+%  Our post pilot survey attempts to shed some light on the reasons behind the inconsistent improvement in navigation ratio.
 Even developers whose navigation ratio  was lower than average, used structured navigation occasionally as much as their more frequently using colleagues.
@@
 Other half of the participants have some understanding on structured navigation use it some of the time.
-While analyzing the points they scored with blaze tool, it is observed that the participants who mentioned that they have good knowledge in structured navigation are on top of the points table.
+While analyzing the points they scored with Blaze tool, we observed that the participants who mentioned that they have good knowledge in structured navigation are on top of the points table.
 %the following statement is not supported with data unless you have additional data?
@@
 
-Assessing the game elements of Blaze that influenced them, four out of six said the leaderboard influenced them a lot and all rated the leaderboard as having some influence.
-Three out of six  said their individual point score influenced them a lot and all said the score had some influence.
+Assessing the game elements of Blaze that influenced them, four out of six said the leaderboard influenced them "a lot" and all rated the leaderboard as having some influence.
+Three out of six  said their individual point score influenced them "a lot" and all said the score had some influence.
 The element with mixed ratings was a graphical indicator on whether they were improving over past use of structured navigation commands.
-The indicator received the a lot rating from one participant and a not at all rating from one participant with the rest having some influence.
+The indicator received the influencing "a lot" rating from one participant and a "not at all" influence rating from one participant with the rest having some influence.
+The results indicate the more obvious and more clearly comparative information in the leaderboard was the strongest influence for developers.
 
-The participants requested that the system to provide more feedback on tips to them to increase their score. This feature of the tool is planned for the next release.
+The participants requested that the system to provide more feedback on tips to help them increase their score.
+They indicated that more obvious feedback such as hints that pop-up when they launch the tool would help them more than passively provided information.
+Their desire for personal metrics and historical views of the data encourages us that this detailed level of measurement and analysis is perceived as helpful.
+Features to meet these requests are part of future work.
 
-Comparing the survey demographic results with game scores reveals that years of experience did not have a strong influence over scores or structured navigation knowledge.
-All developers in the study had 7 or more years experience and one person with 8 years experience used structured navigation enough to obtain second place matching a 10 year experienced colleague.
+Comparing the survey demographic results with game scores reveals that years of experience did not have a strong influence over scores.
+All developers in the study had 7 or more years experience and one person with 8 years experience used structured navigation enough to obtain second place nearly matching a 10 year experienced colleague.
 
@@
 Those who had recently learned the code they were working in had more consistent use of structured navigation and one of them used Sando the most of the pilot participants.
+Familiarity with the code thus may influence the use of structured navigation since when you maintain a mental map of the code, it may be more intuitive for you to click your way to it rather than use a search tool.
+This may be an opportunity for future study to determine whether some developers maintain a somewhat abstract map of the code that makes it easier for them to click to it rather than recall keywords for a search.
+%Could this be similar to how people navigate to software commands  in a menu structure more easily than using command line interfaces.
 
@@
 \section{Conclusions}
-This paragraph will end the body of this sample document.
-Remember that you might still have Acknowledgments or
-Appendices; brief samples of these
-follow.  There is still the Bibliography to deal with; and
-we will make a disclaimer about that here: with the exception
-of the reference to the \LaTeX\ book, the citations in
-this paper are to articles which have nothing to
-do with the present subject and are used as
-examples only.
+Developers accept gamification ideas into their environment even when coupled with detailed development activity monitoring.
+We identified specific concerns particularly around maintaining anonymity of the developer as their data are shared with other developers.
+The participants' feedback indicates a desire for more instructional hints on how to succeed in the game similar to how real games provide hints when the user might be stuck.
+Implementing a more comprehensive set of practices in the game would expand the usefulness of Blaze as a training tool.
+Other ideas might include making the game more subtle so that it is not as obvious a competition, but builds the elements of self-improvement that move the experience from an extrinsically motivated one to an intrinsically motivated one.
 %\end{document}  % This is where a 'short' article might terminate
@@
 \section{Acknowledgments}
-This section is optional; it is a location for you
-to acknowledge grants, funding, editing assistance and
-what have you.  In the present case, for example, the
-authors would like to thank Gerald Murray of ACM for
-his help in codifying this \textit{Author's Guide}
-and the \textbf{.cls} and \textbf{.tex} files that it describes.
+The authors wish to thank ABB and the pilot participants for their support of this research.
 

### 2013-09-10 paper_id1.tex
anil's update

--- 2013-09-09 paper_id1.tex
@@
 \usepackage{cite}
+\usepackage{enumerate}
 % cite.sty was written by Donald Arseneau
@@
 
-Maehr proposes an affirmative theory that individuals achieve as a member of a social group choosing the behavior that will meet the expectations and values of the group that is significant to them.
-\cite{wbsnipes:MaehrCulture}  So the presence of a social group where the person is motivated to belong and excel within is important for creating achievement motivation.
-Maehr states that since achievement behaviors can be triggered by circumstance, that they are present in most people regardless of background.
+Maehr \cite{wbsnipes:MaehrCulture} proposes an affirmative theory that individuals achieve as a member of a social group choosing the behavior that will meet the expectations and values of the group that is significant to them.
+So the presence of a social group where the person is motivated to belong and excel within is important for creating achievement motivation.
+Maehr states that since achievement behaviors can be triggered by circumstance, they are present in most people regardless of background.
 
 Beecham and colleagues provide a thorough analysis of existing literature on studies of motivation in Software Engineering\cite{wbsnipes:Beecham2008Motivation}.
-It has some valuable lists of factors that motivate software engineers with the most common being "the work" motivates us.
+It has some valuable lists of factors that motivate software engineers with the most common being ``the work'' motivates us.
 The list of de-motivators is also useful with common job satisfaction items like stress and inequity in recognition, plus poor quality software (low accomplishment) and lack of influence in decision making.
@@
 We segmented results by country in Figure \ref{fig:gamification} and found that gamification of software engineering is more interesting to developers in India, Poland, and Finland than in other major countries.
-The responses indicating that developers were at least somewhat interested is 75-88\% In India, Poland, and Finland.
+The responses indicating that developers were at least somewhat interested is 75-88\% in India, Poland, and Finland.
 For the US, Sweden, and Switzerland the responses indicate 60-65\% are at least somewhat interested.
@@
 Based on the survey results we made the following conclusions:
-•	95\% of developer responding would try tools and practices suggested by an automated recommendation system particularly if they were provided in context with the current workflow pattern the developer follows.
-•	Responses indicate a preference for team competition over individual awards (badges) and awards for quality practices ranked highest.
+
+\begin{itemize}
+ \item 95\% of developer responding would try tools and practices suggested by an automated recommendation system particularly if they were provided in context with the current workflow pattern the developer follows.
+\item Responses indicate a preference for team competition over individual awards (badges) and awards for quality practices ranked highest.
 So, the tool will incorporate team competition ideas and recognize quality practices.
-•	Over 90\% of respondents approved of sharing data with their team members
+\item Over 90\% of respondents approved of sharing data with their team members
+\end{itemize}
 
@@
 
-Utilizing the unobtrusive continuous monitoring capabilities of Blaze, we designed the pilot as a longtiudinal study in three waves.
+Utilizing the unobtrusive continuous monitoring capabilities of Blaze, we designed the pilot as a longitudinal study in three waves.
 Longtiudinal studies allow us to design an experiment and determine the effects over two or more sample periods using standard MANOVA (Multiple ANalysis of VAriance) models.
@@
 Developing the study with the available tooling means that we follow steps as follows.
-1. Define the practice we wish to measure compliance with
-2. Develop a mapping of the practice to actions that are captured by the blaze tool from the IDE
-3. Weight the actions according to their importance to the practice and assign points to the actions in the configuration file
-4. Develop the intervention method
-5. Determine the phases and length of the phase in the study according to baseline, intervention, and continuous feedback
-6. Select participants to deploy
-7. Define a pre-study questionnaire if one is necessary to collect associated information
-8. Define a post-study questionnaire if necessary to collect post study information
-9. Communicate with the participants about the study
-10. Execute the study and report results
+
+\begin{enumerate}
+\item Define the practice we wish to measure compliance with
+\item Develop a mapping of the practice to actions that are captured by the blaze tool from the IDE
+\item Weight the actions according to their importance to the practice and assign points to the actions in the configuration file
+\item Develop the intervention method
+\item Determine the phases and length of the phase in the study according to baseline, intervention, and continuous feedback
+\item Select participants to deploy
+\item Define a pre-study questionnaire if one is necessary to collect associated information
+\item Define a post-study questionnaire if necessary to collect post study information
+\item Communicate with the participants about the study
+\item Execute the study and report results
+\end{enumerate}
 
@@
 
-We conducted this study with an intact team of 6 developers working in an R\&D facility in India on a large industrial software system.
+As mentioned in fig.
+\ref{fig:gamification}, India is one country where as per our survey, the acceptance to gamification is the highest.
+Hence, we conducted this study with an intact team of 6 developers working in an R\&D facility in India on a large industrial software system.
 We approached a team which develops and maintains an industrial software product.
@@
 It was important to gain the confidence of the participants that their data were kept confidential so their activities would be close as possible to real-world.
-if the participants felt like they are being tracked and the data is going to use for some study that would identify them to their peers or managers, then they may deviate from the natural development style.
+If the participants felt like they are being tracked and the data is going to use for some study that would identify them to their peers or managers, then they may deviate from the natural development style.
 
@@
 
-Competitive dynamics may be more effective when teams from different areas can compete as a team against other teams..
+Competitive dynamics may be more effective when teams from different areas can compete as a team against other teams.
 This tried and true method both builds esprit de corps and enhances achievement towards the intended outcome.
@@
 The following hypotheses interest us in the current study.
-1. Developers using blaze use more structured navigation when they receive instant feedback about their practices
-2. Developers find the competitive aspect of blaze encourages them to pay more attention to navigation practices
+
+\begin{enumerate}
+\item Developers using blaze use more structured navigation when they receive instant feedback about their practices
+\item Developers find the competitive aspect of blaze encourages them to pay more attention to navigation practices
+\end{enumerate}
 
@@
 One question in the post survey is how knowledgeable the developer was about the code they worked in with 3 possibilities,  they were authors of the code, they regularly maintained the code, they were unfamiliar with the code.
-In this team, x\% of developers cited that they ...the code.
+In this team,67\% of developers cited that they work with the code they wrote and the balance 33\% works with the code they recently learned.
 Another factor to consider is experience level which we also asked in the post-pilot survey.

### 2013-09-10 paper_id1.tex
here is the paper with updates.  I had to manually merge so I hope I didn't miss any of your updates.

--- 2013-09-09 paper_id1.tex
@@
 Beecham and colleagues provide a thorough analysis of existing literature on studies of motivation in Software Engineering\cite{wbsnipes:Beecham2008Motivation}.
-It has some valuable lists of factors that motivate software engineers with the most common being "the work" motivates us.
+It has some valuable lists of factors that motivate software engineers with the most common being ''the work'' motivates us.
 The list of de-motivators is also useful with common job satisfaction items like stress and inequity in recognition, plus poor quality software (low accomplishment) and lack of influence in decision making.
@@
 This supports a conclusion that focusing on awarding quality practices would motivate developers to do them consistently.
-
-Based on the survey results we made the following conclusions:
-•	95\% of developer responding would try tools and practices suggested by an automated recommendation system particularly if they were provided in context with the current workflow pattern the developer follows.
-•	Responses indicate a preference for team competition over individual awards (badges) and awards for quality practices ranked highest.
+\begin{itemize}
+\item 95\% of developers responding would try tools and practices suggested by an automated recommendation system particularly if they were provided in context with the current workflow pattern the developer follows.
+\item Responses indicate a preference for team competition over individual awards (badges) and awards for quality practices ranked highest.
 So, the tool will incorporate team competition ideas and recognize quality practices.
-•	Over 90\% of respondents approved of sharing data with their team members
+\item Over 90\% of respondents approved of sharing data with their team members
+\end{itemize}
 
@@
 
-Developing the study with the available tooling means that we follow steps as follows.
-1. Define the practice we wish to measure compliance with
-2. Develop a mapping of the practice to actions that are captured by the blaze tool from the IDE
-3. Weight the actions according to their importance to the practice and assign points to the actions in the configuration file
-4. Develop the intervention method
-5. Determine the phases and length of the phase in the study according to baseline, intervention, and continuous feedback
-6. Select participants to deploy
-7. Define a pre-study questionnaire if one is necessary to collect associated information
-8. Define a post-study questionnaire if necessary to collect post study information
-9. Communicate with the participants about the study
-10. Execute the study and report results
+\begin{enumerate}
+\item Define the practice we wish to measure compliance with
+\item Develop a mapping of the practice to actions that are captured by the blaze tool from the IDE
+\item Weight the actions according to their importance to the practice and assign points to the actions in the configuration file
+\item Develop the intervention method
+\item Determine the phases and length of the phase in the study according to baseline, intervention, and continuous feedback
+\item Select participants to deploy
+\item Define a pre-study questionnaire if one is necessary to collect associated information
+\item Define a post-study questionnaire if necessary to collect post study information
+\item Communicate with the participants about the study
+\item Execute the study and report results
+\end{enumerate}
 

### 2013-10-03 paper_id1.tex
updated data
submitted to reading group for review

@@
 % --- Author Metadata here ---
-\conferenceinfo{WOODSTOCK}{'97 El Paso, Texas USA}
+\conferenceinfo{International Conference on Software Engineering}{2014 Hyderbad, India}
 %\CopyrightYear{2007} % Allows default copyright year (20XX) to be over-ridden - IF NEED BE.
@@
 
-\title{Applying gamification to developer practices}
-\subtitle{Why developer habits die hard}
+\title{Experiences Gamifying Code Navigation}
+\subtitle{Are development practices a habit?}
 % 
@@
 As developers, we both like to try new things and like to stick with the familiar.
-In addition to training, discussing, and presenting software engineering practices and tools, we seek to communicate them through active two-way information embedded in the Integrated Development Environment (IDE).
+In addition to training, discussing, and presenting software engineering practices and tools, we propose a method to both motivate and monitor their adoption through a tool embedded in the Integrated Development Environment (IDE).
 
-We collect data on current practices and tools used by the developer while giving them feedback on how they can improve.
-
+Blaze monitors developer practice and tool use over the long term and reports the lasting change a communication effort makes in everyday habits.
+Embeding Blaze in the developer`s work environment as a Visual Studio extension provides the capture of developer work patterns while giving feedback to users on their practices.
+Blaze encourages everyone to use the new practices by assigning points to specific commands and tools.
  With a rich data source, we evaluate developer practices, define metrics that give individual feedback, and create an environment for developers to see how they are doing compared with their teammates.
 
-Blaze helps understand developer practice and tool use over the long term and judges the lasting change a communication effort makes in everyday habits.
-Embeding Blaze n the developers work environment as a Visual Studio extension provides the capture of usage data while giving feedback to users on their practices.
-Blaze encourages everyone to use the new practices by assigning points to specific commands and tools .
-Providing a light score ranking view should help us stay motivated as we compare our command usage to our peers.
-A month-long longitudinal study conducted with pilot users shows how develipers react to the presented information.
-
-Results from this demonstrative study show the approach is <blank> with a few developers <blank> of the study.
-The effectiveness of this approach for communicating about practices is demonstrated through the continued use of the commands as we communicate them.
-Evaluation of metrics for navigation ratio over time show that developer use the commands <blank> than they did during the benchmarking phase.
-The result of the study shows that this method for communicating and promoting software engineering tools and practices <blank>.
+Data collected from the pilot group show a dichotomy of navigation and code understanding practices used by developers even the same developer in different situations.
+We observed that the duration of an edit session is more correlated with categories of activity that benefit from improved practices and tools.
+The Blaze tool was well received by most participants, however, the intended improvements in navigation practices promoted by the pilot did not occur.
+We conclude that more proactive interventions are necessary to move developers to change their code navigation habits.
 \end{abstract}
@@
 
-As  a developer I'm focused on the problem I'm currently solving and driven by productivity and deadlines to the point that I don't have time to sharpen the saw identifying better practices.
+%As  a developer I'm focused on the problem I'm currently solving and driven by productivity and deadlines to the point that I don't have time to sharpen the saw identifying better practices.
 
@@
 Perhaps the persistence of their habit is why developers are hard to turn from their old practices.
-They may need some coaching to continue the new practice following a given lesson.
-To continuously coach the developer, we conceived of a tool in our prior work to provide constant feedback to retain the attention and interest of the developer.
-\cite{Snipes2013Towards} This tool creates a competitive game that rewards the developer continuously for adopting the improved practice.
+Developers may need some coaching to continue a new practice following an introduction.
+To continuously coach the developer, we conceived of a tool in our prior work to provide constant feedback and motivate them to adopt a new practice.
+\cite{Snipes2013Towards} This tool creates a competitive game that gives continous feedback to the developer continuously on adopting the improved practice.
 The more they change their habits, the higher they score in the game.
@@
 During the pilot of Blaze we observed the variation in developer navigation practices and identified some potential reasons for this variation.
-We evaluated developer's acceptance of game mechanics in the tool and their desires for additional functionality.
-While further data is required to assess the impact of the game, the initial developer acceptance encourages an expanded study.
+We evaluated developer's acceptance of game elements in the tool and their desires for additional functionality.
+While further data is required to assess the impact of the tool, the initial developer acceptance encourages an expanded pilot.
+
+The rest of the paper is organized as follows:  Section 2 reviews the preliminary survey-based assement, Section 3 Discusses the design considerations for developing the Blaze tool/game, Section 4 Discusses the design of the study and participant selection, Section 5 is Related work, Section 6 results from the pilot, Section 7 Threats to Validity, and Section 8 has our Conclusions.
 
@@
 
-\section{Study and Game Design}
+\section{Game Design}
 
-Utilizing the unobtrusive continuous monitoring capabilities of Blaze, we designed the pilot as a longtiudinal study in three waves.
-Longtiudinal studies allow us to design an experiment and determine the effects over two or more sample periods using standard MANOVA (Multiple ANalysis of VAriance) models.
-\cite{longitudinalbook}  We employed the three wave model defining the first wave as a baseline capturing the current practices of the pilot group.
-The second wave employed the intervention that included an informative web site on structured navigation commands in Visual Studio, information about the Sando search tool, and information about how points would be awarded during the pilot.
-The third wave provided the participants with continuous feedback in the from of points accumulated and an indicator arrow \ref{fig:blazegui} that points up when their navigation ratio is higher than their historical average.
+%\begin{enumerate}
+%\item Define the practice we wish to measure compliance with
+%\item Develop a mapping of the practice to actions that are captured by the blaze tool from the IDE
+%\item Weight the actions according to their importance to the practice and assign points to the actions in the configuration file
+%\item Develop the intervention method
+%\item Determine the phases and length of the phase in the study according to baseline, intervention, and continuous feedback
+%\item Select participants to deploy
+%\item Define a pre-study questionnaire if one is necessary to collect associated information
+%\item Define a post-study questionnaire if necessary to collect post study information
+%\item Communicate with the participants about the study
+%\item Execute the study and report results
+%\end{enumerate}
 
-\begin{enumerate}
-\item Define the practice we wish to measure compliance with
-\item Develop a mapping of the practice to actions that are captured by the blaze tool from the IDE
-\item Weight the actions according to their importance to the practice and assign points to the actions in the configuration file
-\item Develop the intervention method
-\item Determine the phases and length of the phase in the study according to baseline, intervention, and continuous feedback
-\item Select participants to deploy
-\item Define a pre-study questionnaire if one is necessary to collect associated information
-\item Define a post-study questionnaire if necessary to collect post study information
-\item Communicate with the participants about the study
-\item Execute the study and report results
-\end{enumerate}
+The pilot evaluates the aspects of game implementation as well as the study of the results of their results.
+Here we discuss how the aspects of selecting a practice suitable for gamification, designing the game, and competitive element considerations.
 
@@
 
+\subsection{Competitive Elements}
+By selecting an intact team, we hope to leverage aspects of motivation such as the Leaderboard to spur feelings of competition among teammates.
+The leaderboard was presented so that developers could see their own score in relation to the top 5 people on the leaderboard.
+However, the identifier of the other participants was generated as 3 capital letters assigned based on their position on the leaderboard.
+Therefore the assigned letters could even change when an individual's ranking changed.
+Not knowing who of your colleagues was ahead of you could be a factor in motivation, but also avoids one potential source of conflict in an intact team.
+
+Competitive dynamics may be more effective when teams from different areas can compete as a team against other teams..
+This tried and true method both builds esprit de corps and enhances achievement towards the intended outcome.
+This study design is potential future work.
+
+\section {Study Design}
+
+Utilizing the unobtrusive continuous monitoring capabilities of Blaze, we designed the pilot to reveal feedback to the developer in 3 stages of 1 week duration each.
+% Longtiudinal studies allow us to design an experiment and determine the effects over two or more sample periods using standard MANOVA (Multiple ANalysis of VAriance) models.
+\cite{longitudinalbook}
+We employed the three stage model defining the first stage as a baseline capturing the current practices of the pilot group.
+The second stage employed the intervention that included an informative web site on structured navigation commands in Visual Studio, information about the Sando search tool, and information about how points would be awarded during the pilot.
+The third stage provided the participants with continuous feedback in the from of points accumulated and an indicator arrow \ref{fig:blazegui} that points up when their navigation ratio is higher than their historical average.
+
 \subsection{Intervention Staging}
 
-After selecting the practice to focus on, we needed to select an intervention method.
-The term intervention, indicates that we are applying a technique to influence a change in the measured behavior of developers.
-Intervention involves communication about the practice in a way that motivates adoption \cite{}.
-Interventions can include presentations, one-on-one discussions, distance education mechanisms such as webinars, and word of mouth recommendations.
-Part of the research objective is to determine whether we can achieve adoption of a practice with alternative interventions that are more deliberate than word of mouth and perhaps more scalable than webinars or presentations.
+Part of the research objective was to determine whether we can achieve adoption of a practice with alternative interventions that are more deliberate than word of mouth and perhaps more scalable than webinars or presentations.
+As previously discussed we sought to provide Blaze as a means to influence developers to use structured navigation practices.
+We constructed the interventions during the pilot to roll out in stages each week that they use the tool.
+This allowed us to establish a baseline and determine the changes as the pilot continued.
 
-In the initial wave, participants were informed of the purpose of the study what to expect from the Blaze tool.
+In the initial stage, participants were informed of the purpose of the study what to expect from the Blaze tool.
 The participants installed the tool in their Visual Studio environment.
@@
 
-After a week's data collection marking the end of the first wave, Blaze started to automatically pop-up a window containing a button for a web page with information.
-The page included a link to download another tool for code search called Sando \ref{sando} and information about using that tool in a video.
+After a week's data collection marking the end of the first stage, Blaze started to automatically pop-up a window containing a button for a web page with information.
+The page included a link to download another tool for code search called Sando \cite{Shepherd2012Sando} and information about using that tool in a video.
 The sub-page on navigation contained a page of tutorial on the built-in structured navigation commands in Visual Studio.
@@
 
-In the third wave developers received instant feedback from Blaze on the points they accumulated during the development session.
+In the third stage developers received instant feedback from Blaze on the points they accumulated during the development session.
 The up/down arrow provided a simple indicator on whether they were increasing their use of structured navigation or were using it less than their personal historical average.
@@
 %Through this approach, we ensured that the developers are performing the task on a very natural environment with out bothering that they are being tracked.
-
-\subsection{Competitive Elements}
-By selecting an intact team, we hope to leverage aspects of motivation such as the Leaderboard to spur feelings of competition among teammates.
-The leaderboard was presented so that developers could see their own score in relation to the top 5 people on the leaderboard.
-However, the identifier of the other participants was generated as 3 capital letters assigned based on their position on the leaderboard.
-Therefore the assigned letters could even change when an individual's ranking changed.
-Not knowing who of your colleagues was ahead of you could be a factor in motivation, but also avoids one potential source of conflict in an intact team.
-
-Competitive dynamics may be more effective when teams from different areas can compete as a team against other teams..
-This tried and true method both builds esprit de corps and enhances achievement towards the intended outcome.
-This study design is potential future work.
-
 \subsection{Pilot Survey}
 
-The goal for the pilot follow-up survey was to collect feedback on the tool and the effect it had on the developers navigation practices and knowledge.
+The goal for the pilot follow-up survey was to collect feedback on the tool and the effect it had on the developers` navigation practices and knowledge.
 The survey questions were segmented into portions that evaluate prior and post knowledge and practices for navigation, key steps in the pilot itself, the influence of game elements on their navigation practices, typical tasks and demographic information.
@@
 %update and expand
-
+\subsection{Monitoring Practice Studies }
 Johnson and Kou defined Zorro\cite{V:Johnson2007Automated}, a system for detecting whether developers use Test Driven Development techniques based on data from Hackystat.
@@
 Key conclusions are that developers are more more successful finding and fixing bugs when they create a detailed plan for implementing a change, use structured navigation with keyword search or cross-reference search, and only review methods once during their search.
-This builds on this work by testing methods for increasing the use of structured navigation in developers everyday practice.
-
-Singer and Schneider demonstrate the use of a message board and points for encouraging students to increase their frequency of commits to the source code repository.\cite{Singer2012It}  The communication mechanism enabled students to see each others' progress and resulted in more frequent commits than baseline.
-Participants valued the communication and collaboration aspect and some valued the competition enough to change their opinion on optimum commit frequency.
-The subsequent thesis by Singer \cite{Singer2013a} describes results of an experiment conducted across two iterations of a class where active feedback on commits was deployed to one course and another course served as the control group where commit frequency was simply monitored.
-Results show an increase in the frequency of commits at a statistically significant level.
-This study uses fine grained instrumentation to identify practices used between commits such as structured navigation.
+This builds on this work by testing methods for increasing the use of structured navigation in developers` everyday practice.
 
@@
 
+\subsection{Gamification}
+Singer and Schneider demonstrate the use of a message board and points for encouraging students to increase their frequency of commits to the source code repository.\cite{Singer2012It}  The communication mechanism enabled students to see each others' progress and resulted in more frequent commits than baseline.
+Participants valued the communication and collaboration aspect and some valued the competition enough to change their opinion on optimum commit frequency.
+The subsequent thesis by Singer \cite{Singer2013a} describes results of an experiment conducted across two iterations of a class where active feedback on commits was deployed to one course and another course served as the control group where commit frequency was simply monitored.
+Results show an increase in the frequency of commits at a statistically significant level.
+This study uses fine grained instrumentation to identify practices used between commits such as structured navigation.
+
 Dubois and Tamburrelli discuss theoretical concepts behind applying gamification to software development.
@@
 
-\subsubsection{Observations about navigation ratio improvement}
+Blaze logs events from actions the developer takes or actions from Visual Studio itself.
+ The event is basically a GUI event that the program manages with event handlers that are registered with the application to listen for the event.
+Blaze becomes a global event handler listening for all events by registering for them in Visual Studio.
 
-To evaluate whether developers improved their practices for structured navigation we established a metric Navigation ratio as the number as the number of structured navigation events in a day over the number of unstructured navigation events.
+Blaze records key attributes along with each event in visual studio.
+The log captures a time-stamp for each event normalized to GMT.
+The name of the event provides a low-level classification.
+The event type reflects an internal classification in Blaze.
+An optional field for Artifact Reference captures data such as the file-name being edited and the currently selected line number.
+Finally an anonymized unique identifier for the user of blaze allows us to investigate differences between developers.
+
+Analysis of the log extracts several key information bits from this raw source.
+First we classify the event names at a higher level into categories related activities such as Navigating, Reading, Editing, Building, Debugging,  Testing, and using a  known Tool.
+
+The second information we extract is aggregating the events along the time-line into sessions.
+A session is an abstract concept that groups multiple events into a sequence with a beginning and an end.
+Of the many possible ways to define sessions, we chose to define sequences as all the events that transpire before and during a set of edits to a specific file.
+When the developer edits a new file, all the events after the last edit to the previous file are part of the new sequence for the new file.
+Although not at the method level, this is similar to the say Robillard et.
+al.
+generated sequence based on located methods for their study.\cite{wbsnipes:Robillard2004How}
+
+\subsubsection{Observations About Developers` Habits}
+
+We are interested in understanding whether developers are consistently using practices that make them inefficient.
+One way to check that is to determine whether our measurements are associated with specific developers in the study, in other words, can we independently classify the edits performed by each developer.
+If this is possible, then we can discuss whether developer have consistent ways of navigating and understanding for each edit.
+Thus the hypothesis is developers have consistent ways of navigating and understanding code while they are making changes.
+
+To test this hypothesis we construct a classification tree using Weka \cite{Hall2009WEKA} with the J48 classifier.
+Slicing the data and training the model using 10-fold cross-validation, the model identifies 53\% of the 647 edit sessions with the correct developer among 6 participants.
+The tree uses nearly all available fields including time for all categories and time of day for edits in its attempt to associate sessions with their developer generating a Kappa statistic of 0.403  and square root of the relative squared error at 101\%.
+
+While there is some consistency to how developers work, we cannot say that individual developers are always consistent in their work patterns.
+Thus developers are somewhat flexible in the commands they use when navigating and making changes to code and may be more open to adopting new tools and techniques.\
+
+\subsubsection{Observations About Drivers of Edit Session Duration}
+
+We selected navigation as the subject of the pilot game because it was identified by Robillard et.
+al\cite{wbsnipes:Robillard2004How} as correlated with developers who effectively completed a maintenance task.
+Are we focusing on the right area of developers` activity in Visual Studio?
+
+In order to test whether navigation is an important area to address, we construct a hypothesis that use of unstructured navigation is an important factor in developer productivity.
+Edit session duration, we assert by opinion, is a factor related to developer productivity though there are many other factors we could consider.
+With the data available from Blaze, we can explore the categories of events that correlate with the duration of edit sessions over the period of the pilot.
+
+Using Weka\cite{Hall2009WEKA}, we performed an attribute selection process to identify the most significant attributes related to edit session duration.
+The attribute search method we chose was Greedy Step-wise search that selects attributes based on their correlation with the target and stops adding attributes when the next attribute decreases the correlation.
+The process identified the top 5 out of 10 possible categories as follows:
+\begin{enumerate}
+\item Unstructured Navigation
+\item Read
+\item Build
+\item Debug
+\item Edit
+\end{enumerate}
+
+Using R\cite{Rcitation}, we constructed a linear regression model for Unstructured Navigation to determine the correlation with edit duration.
+The linear model found a positive correlation with unstructured navigation with a p-value <0.001 on 646 degrees of freedom.
+The model with only unstructured navigation had an R-squared of 0.52 showing a good portion of the variation is explained by that category.
+In order to have a more acceptable explanation of edit duration, we conducted a multiple linear regression model with both Unstructured Navigation and Read categories.
+The Read category of events captures the activities the developer is doing to look through the code in the editor.
+They are navigation commands that move incrementally like pressing the up/down arrow keys, or moving the scroll-bar an incremental amount.
+Combining these two categories results in a linear model with a p-value <0.001 and an R-squared of 0.76, more acceptable for drawing conclusions.
+
+Conducting an ANOVA analysis shown in Table \ref{editDurationANOVA} for this regression model, we can explore the contribution of Unstructured Navigation and Reading to the model.
+The contribution of each variable can be determined by calculating the portion of the total variation that each variable explains.
+Unstructured Navigation explains 53\% of the variation.
+Out of the remaining variation found by subtracting the amount explained by Unstructured Navigation from the total, Read explains 49\%.
+
+\begin{table}[!t]
+
+\begin{mdframed}[linecolor=white]
+\begin{tabular}{|l|r|r|r|r|r|}
+\hline
+                          &  Df  & SSQ & MeanSq &Fval.   & Pr(>F)    \\
+\hline
+Unstr. Nav.& 1 & 0.039& 0.03920 & 1383 & <0.001\\
+\hline
+Read                 & 1 & 0.017 & 0.01741 & 614 & <0.001\\
+\hline
+Residuals                & 645 & 0.018 & 0.00003   &  &  \\
+\hline
+Total		& 647 & 0.074 & 0.05664 & & \\
+\hline
+\end{tabular}
+\caption{ANOVA for Edit Duration}
+\label{editDurationANOVA}
+\end{mdframed}\end{table}
+
+After this analysis we cannot disprove that unstructured navigation is correlated with edit duration, thus it is an important factor in developer productivity.
+Combining unstructured navigation with reading describes a larger portion of the overall edit duration and motivates us towards improvements in both navigation methods and code comprehension.
+
+\subsubsection{Observations About navigation ratio improvement}
+
+To evaluate whether developers improved their practices for structured navigation we established a metric Navigation ratio as the number as the number of structured navigation events in a session or period of time over the number of unstructured navigation events.
 
 Structured navigation events included: 
+\begin{itemize}
+\item Navigate To (Ctrl+,) is a fuzzy search interface that locates symbols matching your search terms.
+\item Go To Definition (F12) brings up the code where the selected symbol is defined. 
+\item View Call Hierarchy (Ctrl+K Ctrl+T) provides a two way analysis of a symbol's dependencies and uses. 
+\item Class View (Ctrl+W, C) priovides a searchable browser of classes and class hierarchy
+\item Find All References provides a list of refernces for an object
+\item Navigate to Event Handler brings up the event handler from the XAML editor
+\item View Class Diagram generates a class diagram for the project
+\item View Obeject Browser is a search tool for objects in the solution
+\end{itemize}
 
-Unstructured navigation events included:
+Unstructured navigation events included selecting a file in an explorer window or selecting the tab for a file, using arrow and page up/down keys to go up/down through a file, scrolling, clicking on a file element, and using any of the built-in find commands such as Find In Files or Quick Find.
 
-This metric provides a continuous variable allowing us to use statistical data of continuous data to check our hypotheses.
-To create the evaluation for each wave, we took the navigation ratio for each developer in each wave.
+During the pilot we staged the deployment of information feedback to developers so they would gradually improve their structured navigation practices.
+Our hypothesis was that this instant feedback and information would result in an increase in the use of structured navigation commands in Visual Studio.
 
-%The following hypotheses interest us in the current study.  
-%1. Developers using blaze use more structured navigation when they receive instant feedback about their practices
-%2. Developers find the competitive aspect of blaze encourages them to pay more attention to navigation practices
-
-%To evaluate whether developers use more structured navigation when they receive instant feedback, we monitor the change in navigation ratio over the pilot study period.
+To evaluate whether developers use more structured navigation when they receive instant feedback, we monitor the change in navigation ratio over the pilot study period.
 Expecting that as developers learn about navigation commands and tools that count towards the score in the second week they will begin using the practices more.
 Also expecting that noticing the leaderboard competition will encourage some developers to use the navigation practices more.
-What we observed is shown in Figure \ref{fig:navigationaverage} that the structured navigation practice did not increase significantly from the beginning to the end of the study.
-Overall, we did not see a significant effect towards increasing navigation ratio for the interventions in wave 2 or wave 3.
 
 \begin{figure}\begin{mdframed}[linecolor=white]
-	\includegraphics[width=3in]{navigationaverage.pdf}
-	\caption{Average Navigation Ratio by Week by Developer}
+	\includegraphics[width=3in]{navratioboxplot.pdf}
+	\caption{Navigation Ratio by Week}
 	\label{fig:navigationaverage}
@@
 
-\subsubsection{Observations about efficiency of structured navigation}
-
-Structured navigation sequences are less frequent between edits than unstructured navigation.
-To study navigation practices in relationship to edits, we grouped the data into sequences delimited by edits meaning each sequence ends with one or more edits to the source code.
-We evaluated the number of navigation commands between edits including clicks on locations in the file, navigating between files, and scrolling the file in the window.
-
-We found that in x\% of the edit sequences structured navigation was sued.
-We compared the time between edits where the sequences involved structured navigation to those that involved unstructured navigation.
-The data supports the developer's perception that unstructured navigation is faster than structured navigation particularly when search is used.
-We did find that when navigating between class files, developers who use structured navigation reached their destination X\% faster/slower than when using unstructured navigation.
-
-Search was instrumental for developers who used the search tool where occurrences including search were completed x\% faster or slower than navigating by unstructured commands.
-
-%  Our post pilot survey attempts to shed some light on the reasons behind the inconsistent improvement in navigation ratio.
-Even developers whose navigation ratio  was lower than average, used structured navigation occasionally as much as their more frequently using colleagues.
-The variation for each individual can be seen in Figure \ref{} .
+What we observed visually in Figure \ref{fig:navigationaverage} is that the structured navigation practice did not increase significantly from the beginning to the end of the study.
+Overall, we did not see a significant effect towards increasing navigation ratio for the interventions in week 2 or week 3.
+Weeks 4 and 5 that follow the formal period of study show the navigation ratio continues in a similar distribution pattern.
 
@@
 
+Our post pilot survey attempts to shed some light on the reasons behind the consistency in navigation ratio over the pilot.
+
 On the assessment of structured navigation knowledge, half of the participants replied that they had very good prior knowledge about structured navigation and they use structured navigation as much as possible.
 Other half of the participants have some understanding on structured navigation use it some of the time.
+Even developers whose navigation ratio  was lower than average, used structured navigation occasionally as much as their more frequently using colleagues.
 While analyzing the points they scored with Blaze tool, we observed that the participants who mentioned that they have good knowledge in structured navigation are on top of the points table.
@@
 Incidentally, while cross verifying their performance on the coding, they are better productive compared to the participants on the bottom of points table.
+\begin{figure}\begin{mdframed}[linecolor=white]
+	\includegraphics[width=3in]{ElementInfluenceChart.pdf}
+	\caption{Influence of Feedback Elements}
+	\label{fig:elementInfluence}
+\end{mdframed}\end{figure}
 
-Assessing the game elements of Blaze that influenced them, four out of six said the leaderboard influenced them "a lot" and all rated the leaderboard as having some influence.
+The chart in figure \ref{fig:elementInfluence} shows how the individual feedback features influenced developers to think about using structured navigation.
+Four out of six developers said the leaderboard influenced them "a lot" and all rated the leaderboard as having some influence.
 Three out of six  said their individual point score influenced them "a lot" and all said the score had some influence.
@@
 
-Comparing the survey demographic results with game scores reveals that years of experience did not have a strong influence over scores.
-All developers in the study had 7 or more years experience and one person with 8 years experience used structured navigation enough to obtain second place nearly matching a 10 year experienced colleague.
-
-Those who rated their knowledge of structured navigation commands highly, did better in the points than those who rated their prior knowledge lower.
-This could indicate that using structured navigation is somewhat of a habit that takes longer to change.
-There was also feedback that people wanted more information presented to them by the tool on what commands to use, so they may not have closed the knowledge gap during the pilot.
-
-Those who worked in code that they wrote tended to use less structured navigation having a lower navigation ratio.
-Those who had recently learned the code they were working in had more consistent use of structured navigation and one of them used Sando the most of the pilot participants.
-Familiarity with the code thus may influence the use of structured navigation since when you maintain a mental map of the code, it may be more intuitive for you to click your way to it rather than use a search tool.
-This may be an opportunity for future study to determine whether some developers maintain a somewhat abstract map of the code that makes it easier for them to click to it rather than recall keywords for a search.
-%Could this be similar to how people navigate to software commands  in a menu structure more easily than using command line interfaces.
+%Though not significant, there is a loose correlation between those who rated their prior knowledge of structured navigation commands highly and their point scores.   
+There was also feedback that people wanted more information presented to them by the tool on what structured navigation commands to use, so they may not have closed the knowledge gap during the pilot.
 
@@
 \balance
-\bibliography{paper_id1} 
+\bibliography{../mousetap} 
  % sigproc.bib is the name of the Bibliography in this case

### 2013-10-09 paper_id1.tex
revised in the process of looking at comments.

@@
 \begin{abstract}
-As software development practices evolve, we face the continuous challenge of communicating new practices and tools to the community.
-As developers, we both like to try new things and like to stick with the familiar.
+As software development practices evolve, toolsmiths face the continuous challenge of communicating new practices and tools to the community.
+Developers both like to try new things and like to stick with the familiar.
 In addition to training, discussing, and presenting software engineering practices and tools, we propose a method to both motivate and monitor their adoption through a tool embedded in the Integrated Development Environment (IDE).
@@
 Blaze monitors developer practice and tool use over the long term and reports the lasting change a communication effort makes in everyday habits.
-Embeding Blaze in the developer`s work environment as a Visual Studio extension provides the capture of developer work patterns while giving feedback to users on their practices.
+Embeding Blaze in the developer`s work environment as a Visual Studio extension captures developer work patterns while giving feedback to users on their practices.
 Blaze encourages everyone to use the new practices by assigning points to specific commands and tools.
@@
 
-\terms{gamification}
+\terms{Human Factors, Management}
 
@@
 Developers with years of experience develop habits and ways of doing things that are difficult to change even when they might recognize better methods.
-For instance, the commands they use and ways they use to navigate the code base in the development environment make them less efficient than using newer practices.
+For instance, the commands they use and ways they use to navigate the code base in the development environment may make them less efficient than using newer practices.
 Robillard et.
 al.
-show that developers who think of code as a graph and use commands that allow them to follow code structure are more effective and efficient than developers who browse through code.
-\cite{wbsnipes:Robillard2004How}
+show that developers who use commands that allow them to follow code structure are more effective and efficient than developers who browse through code \cite{wbsnipes:Robillard2004How}.
+Can we make developers more productivity by training them to use structured navigation commands and tools?
 
@@
 The more they change their habits, the higher they score in the game.
-
+%clunky
 Taking a step back, gamification is a popular movement that people apply to customer oriented web sites and social games.
@@
 While further data is required to assess the impact of the tool, the initial developer acceptance encourages an expanded pilot.
+%state the contribution
 
@@
 
-\section{Assess the Viability of Gamification for Software Engineering}
+\section{Related Work}
+%update and expand
+\subsection{Monitoring Practice Studies }
+Johnson and Kou defined Zorro\cite{V:Johnson2007Automated}, a system for detecting whether developers use Test Driven Development techniques based on data from Hackystat.
+Zorro divides development activities into episodes delimited by events such as configuration management code check-in, start of a unit test run, or start of a build.
+Using the distinct events that developers follow within these episodes, Zorro determines whether the episode followed Test-Driven Development practices per their classic definition of test first - code - test pass, or a different scenario.
+In two student-based studies comparing Zorro classifications with a simultaneous observational screen video, Zorro achieved between 70\% \cite{Kou2010Operational} and 89\% \cite{V:Johnson2007Automated} accuracy in classifying episodes into their proper TDD scenarios.
+The studies did not, however, attempt to use or evaluate the influence of instant feedback aimed at encouraging participants to follow the classic definition of Test Driven Development.
+In our study, we attempt this leap to present data collected directly to developers attempting to influence them to change their practices.
 
-\subsection{Assessment Background}
+Murphy-Hill, Parnin, and Black \cite{V:MurphyHill2012How} use the Mylyn Monitor to explore whether or not developers use the automated refactoring tools present in Eclipse.
+They look for specific refactoring commands in Eclipse and determine the amount of time developers use tools versus hand refactoring the code.
+For this study we build a tool that captures the use of all commands within the Visual Studio environment including refactoring, navigation, and edit commands.
+
+Robillard, Coelho and Murphy explore hypotheses around how developers can be more effective at performing a maintenance task \cite{wbsnipes:Robillard2004How}.
+Key conclusions are that developers are more more successful finding and fixing bugs when they create a detailed plan for implementing a change, use structured navigation with keyword search or cross-reference search, and only review methods once during their search.
+This builds on this work by testing methods for increasing the use of structured navigation in developers` everyday practice.
+
+The Pro Metrics (PROM) tool provides a framework for collecting data for further analysis from tools used by developers.\cite{Coman2009Casestudy}  It provides a flexible data model and a plug-in architecture to facilitate collection from different data sources.
+Studies conducted using prom include a series of studies by it's creators on trends in time spent Pair Programming \cite{Coman2008Investigating}, benefits of refactoring on productivity \cite{Moser2008Case}, impact of refactoring on re-usability \cite{Moser2006Does}, and prediction of effort \cite{Abrahamsson2007Effort}.
+The PROM tool and studies that use it track the time spent editing files and methods from the IDE and code metrics obtained trough source code analysis.
+Applying these data to specific research questions such as whether refactoring improves productivity \cite{Moser2008Case} shows the utility of combining automated effort data with code metric data.
+Taking a different focus on techniques, our study used more detailed events captured from the IDE  to detect how the user was navigating through the code in addition to knowing how long the editor is open for a particular file.
+
+Murphy-Hill et.
+al.
+study a large usage history data set and apply several different algorithms to accurately suggest to users new commands in their environment to use.\cite{MurphyHill2012Improving} Algorithms based on command history performed well particularly when synchronized chronologically with the recipient's usage history.
+Novice users were better served by recommendation algorithms that did not require a long history, where more expert users benefited from more sophisticated algorithms that included more history.
+The "Most Widely Used" algorithm that recommends commands based on the collective usage profile of the team performed nearly as well as the more sophisticated algorithms.
+The paper presents the methodology for making recommendations and the evaluation of the accuracy of the recommendations.
+Our work goes the next step of evaluating whether a specific recommended practice gets adopted by developers when influenced through gamification methods.
+
+\subsection{Gamification}
+Singer and Schneider demonstrate the use of a message board and points for encouraging students to increase their frequency of commits to the source code repository \cite{Singer2012It}.
+The communication mechanism enabled students to see each others' progress and resulted in more frequent commits than baseline.
+Participants valued the communication and collaboration aspect and some valued the competition enough to change their opinion on optimum commit frequency.
+The subsequent thesis by Singer \cite{Singer2013a} describes results of an experiment conducted across two iterations of a class where active feedback on commits was deployed to one course and another course served as the control group where commit frequency was simply monitored.
+Results show an increase in the frequency of commits at a statistically significant level.
+This study uses fine grained instrumentation to identify practices used between commits such as structured navigation.
+
+Dubois and Tamburrelli discuss theoretical concepts behind applying gamification to software development \cite{Dubois2013Understanding}.
+Their approach is to consider how gamification could be applied throughout the software development life-cycle to motivate many practices in different phases.
+They provide guidance on assessing gamification in the software development domain that includes analyzing the actors and game mechanics, defining integration with existing tools and processes, and designing the evaluation of results.
+Dbois and Tamburrelli demonstrate this guidance using a study of students working on a class project.
+Defining a control group and subject group of students, both groups receive the same feedback about their code comments, test coverage, code size, and static analysis rule violations.
+The subject group can see each others' rankings while the control group only sees their personal rankings.
+Results though inconclusive indicate that competition increases the desired results.
+In our study we apply both individual feedback and competitive based feedback (anonymized) to participants and seek to observe a change in their behavior after an intervention.
+
+\section{Pre-study Sruvery on Viability of Applying Gamification for Software Engineering}
+
+\subsection{Pre-study Background}
 When discussing gamification, folks raise a concern about replacing an activity that is intrinsically motivated with extrinsic motivation provided by points and achievements.
@@
 
-Maehr proposes an affirmative theory whereby  individuals achieve as a member of a social group choosing the behavior that will meet the expectations and values of the group that is significant to them.
-\cite{wbsnipes:MaehrCulture}  The presence of a social group where the person is motivated to belong and excel within is important for creating achievement motivation.
+Maehr proposes an affirmative theory whereby  individuals achieve as a member of a social group choosing the behavior that will meet the expectations and values of the group that is significant to them \cite{wbsnipes:MaehrCulture}.
+The presence of a social group where the person is motivated to belong and excel within is important for creating achievement motivation.
 Maehr states that since achievement behaviors can be triggered by circumstance, that they are present in most people regardless of background.
 
-Beecham and colleagues provide a thorough analysis of existing literature on studies of motivation in Software Engineering\cite{wbsnipes:Beecham2008Motivation}.
+Beecham and colleagues provide a thorough analysis of existing literature on studies of motivation in Software Engineering \cite{wbsnipes:Beecham2008Motivation}.
 It has some valuable lists of factors that motivate software engineers with the most common being ''the work'' motivates us.
@@
 
-\subsection{Assessment Results}
+\subsection{Pre-study Survey Results}
 
-Using a question from the Saachi and Saachi survey\cite{wbsnipes:SaatchiGameification} as directly as possible, we asked developers at ABB "How interested they would be in working for a company that incorporated some aspects of games into software engineering tools as a way to increase productivity in the workplace"?
+Using a question from the Saachi and Saachi survey \cite{wbsnipes:SaatchiGameification} as directly as possible, we asked developers at ABB "How interested they would be in working for a company that incorporated some aspects of games into software engineering tools as a way to increase productivity in the workplace"?
 We segmented results by country in Figure \ref{fig:gamification} and found that gamification of software engineering is more interesting to developers in India, Poland, and Finland than in other major countries.
@@
 The overall response from our survey is comparable to the Saachi and Saaschi survey\cite{wbsnipes:SaatchiGameification} upon which this question was based that found 75\% of 18-45 year olds were at least somewhat interested.
-
-\begin{figure}\begin{mdframed}[linecolor=white]
-	\includegraphics[width=3in]{gamificationquestion.pdf}
+%EMH why is our survey result comparable to Saatchi?
+\begin{figure}
+	\includegraphics[width=3.4in]{gamificationquestion.pdf}
 	\caption{Acceptance of gamification}
 	\label{fig:gamification}
-\end{mdframed}\end{figure}
+\end{figure}
 
 We asked developers how likely they would be to try tools and practices recommended to them through an automated usage tracking system.
-Answers showed that 95\% of the developers surveyed are likely to try tools and practices recommended to them.
+Answers showed that 95\% of the developers surveyed are likely to try the recommended tools and practices.
 This indicates that a recommendation system would positively impact the deployment of software engineering tools and practices at ABB.
 
-One area we questioned was whether sharing detailed usage information with colleagues and the company as a whole would be a concern for developers.
+Another question was whether sharing detailed usage information with colleagues and the company as a whole would be a concern for developers.
 A graduated scale was used for the community that would be shared with and distinction between whether that sharing was anonymous or the information was personally identifiable.
@@
 
-\begin{figure}\begin{mdframed}[linecolor=white]
-	\includegraphics[width=3in]{ComfortWithSharing.pdf}
+\begin{figure}
+	\includegraphics[width=3.4in]{ComfortWithSharing.pdf}
 	\caption{Comfortable with sharing usage information}
 	\label{fig:comfortwithsharing}
-\end{mdframed}\end{figure}
+\end{figure}
 
@@
 
-\begin{figure}\begin{mdframed}[linecolor=white]
-	\includegraphics[width=3in]{ToolAndPracticeMotivators.pdf}
+\begin{figure}
+	\includegraphics[width=3.4in]{ToolAndPracticeMotivators.pdf}
 	\caption{Motivation for using tools and practices}
 	\label{fig:toolandpracticemotivators}
-\end{mdframed}\end{figure}
+\end{figure}
 
@@
 This supports a conclusion that focusing on awarding quality practices would motivate developers to do them consistently.
+%EMH connection?
 \begin{itemize}
@@
 
+%distinguish pilot from pre-study.  How does gamification apply?
 The pilot evaluates the aspects of game implementation as well as the study of the results of their results.
@@
 \subsection{Select a Practice}
+%transition
 In their paper on game design patterns, Hamari et.
@@
 This warning indicates that we should avoid giving points for things that are already an expected deliverable such as finding or fixing bugs, completing development tasks, or meeting required process criteria.
+%passive
 Practices to consider are those practices with room to go beyond a checklist or minimum criteria where developers have the opportunity to excel.
-
+%passive
 Selecting a practice to study involves identifying a practice, defining specific characteristics of the practice to measure, and identifying the actions in Visual Studio that relate to the practice.
@@
 We identified potential practices of test-driven development, debugging practices, navigation practices,  eliminating static analysis bugs, and check-in frequency.
-
+%EMH is check-in frequency a practice?
 Practices of check-in frequency, static analysis bug elimination are highlighted in this paper's related work section with prior gamification studies in classroom environments.
 Johnson and Kou achieved automated monitoring of test-driven development practices with Zorro thus may provide a candidate for future gamification application.
+%reference
 Debugging practices have developer community opinions on optimal methods, however, detailed debug practices with modern IDEs have not received research community attention, thus are also reserved for future work.
@@
 Using these guidelines we assigned points to structured navigation commands which received 1-10 points each while unstructured navigation commands received 0 points.
-Assigning negative points is discouraged as this may demotivate folks who do not have detailed knowledge of how the points are assigned.\cite{}   In creating levels, following the guidance from Hamari \cite{wbsnipes:Hamari2011Framework}, we built levels following an exponential curve.
+In creating levels, following the guidance from Hamari \cite{wbsnipes:Hamari2011Framework}, we built levels following an exponential curve.
 Users initially level up after a working day or two extending to achieving the next level after a week or even a month following an exponential curve.
@@
 The second stage employed the intervention that included an informative web site on structured navigation commands in Visual Studio, information about the Sando search tool, and information about how points would be awarded during the pilot.
-The third stage provided the participants with continuous feedback in the from of points accumulated and an indicator arrow \ref{fig:blazegui} that points up when their navigation ratio is higher than their historical average.
+The third stage provided the participants with continuous feedback in the from of points accumulated and an indicator arrow that points up when their navigation ratio is higher than their historical average.
 
@@
 
-As mentioned in fig.
-\ref{fig:gamification}, India is one country where as per our survey, the acceptance to gamification is the highest.
+As mentioned in Figure \ref{fig:gamification}, India is one country where as per our survey, the acceptance to gamification is the highest.
 Hence, we conducted this study with an intact team of 6 developers working in an R\&D facility in India on a large industrial software system.
@@
 Another demographic question is how well they know the code they work in rated from "I work in code that I wrote", "I maintain code that I am familiar with", to "I recently learned the code"
-
-Survey results are discussed in the results section \ref{Results}
-
-\section{Related Work}
-%update and expand
-\subsection{Monitoring Practice Studies }
-Johnson and Kou defined Zorro\cite{V:Johnson2007Automated}, a system for detecting whether developers use Test Driven Development techniques based on data from Hackystat.
-Zorro divides development activities into episodes delimited by events such as configuration management code check-in, start of a unit test run, or start of a build.
-Using the distinct events that developers follow within these episodes, Zorro determines whether the episode followed Test-Driven Development practices per their classic definition of test first - code - test pass, or a different scenario.
-In two student-based studies comparing Zorro classifications with a simultaneous observational screen video, Zorro achieved between 70\% \cite{Kou2010Operational} and 89\% \cite{V:Johnson2007Automated} accuracy in classifying episodes into their proper TDD scenarios.
-The studies did not, however, attempt to use or evaluate the influence of instant feedback aimed at encouraging participants to follow the classic definition of Test Driven Development.
-In our study, we attempt this leap to present data collected directly to developers attempting to influence them to change their practices.
-
-Murphy-Hill, Parnin, and Black \cite{V:MurphyHill2012How} use the Mylyn Monitor to explore whether or not developers use the automated refactoring tools present in Eclipse.
-They look for specific refactoring commands in Eclipse and determine the amount of time developers use tools versus hand refactoring the code.
-For this study we build a tool that captures the use of all commands within the Visual Studio environment including refactoring, navigation, and edit commands.
-
-Robillard, Coelho and Murphy explore hypotheses around how developers can be more effective at performing a maintenance task \cite{wbsnipes:Robillard2004How}.
-Key conclusions are that developers are more more successful finding and fixing bugs when they create a detailed plan for implementing a change, use structured navigation with keyword search or cross-reference search, and only review methods once during their search.
-This builds on this work by testing methods for increasing the use of structured navigation in developers` everyday practice.
-
-The Pro Metrics (PROM) tool provides a framework for collecting data for further analysis from tools used by developers.\cite{Coman2009Casestudy}  It provides a flexible data model and a plug-in architecture to facilitate collection from different data sources.
-Studies conducted using prom include a series of studies by it's creators on trends in time spent Pair Programming \cite{Coman2008Investigating}, benefits of refactoring on productivity \cite{Moser2008Case}, impact of refactoring on re-usability \cite{Moser2006Does}, and prediction of effort \cite{Abrahamsson2007Effort}.
-The PROM tool and studies that use it track the time spent editing files and methods from the IDE and code metrics obtained trough source code analysis.
-Applying these data to specific research questions such as whether refactoring improves productivity \cite{Moser2008Case} shows the utility of combining automated effort data with code metric data.
-Taking a different focus on techniques, our study used more detailed events captured from the IDE  to detect how the user was navigating through the code in addition to knowing how long the editor is open for a particular file.
-
-Murphy-Hill et.
-al.
-study a large usage history data set and apply several different algorithms to accurately suggest to users new commands in their environment to use.\cite{MurphyHill2012Improving} Algorithms based on command history performed well particularly when synchronized chronologically with the recipient's usage history.
-Novice users were better served by recommendation algorithms that did not require a long history, where more expert users benefited from more sophisticated algorithms that included more history.
-The "Most Widely Used" algorithm that recommends commands based on the collective usage profile of the team performed nearly as well as the more sophisticated algorithms.
-The paper presents the methodology for making recommendations and the evaluation of the accuracy of the recommendations.
-Our work goes the next step of evaluating whether a specific recommended practice gets adopted by developers when influenced through gamification methods.
-
-\subsection{Gamification}
-Singer and Schneider demonstrate the use of a message board and points for encouraging students to increase their frequency of commits to the source code repository.\cite{Singer2012It}  The communication mechanism enabled students to see each others' progress and resulted in more frequent commits than baseline.
-Participants valued the communication and collaboration aspect and some valued the competition enough to change their opinion on optimum commit frequency.
-The subsequent thesis by Singer \cite{Singer2013a} describes results of an experiment conducted across two iterations of a class where active feedback on commits was deployed to one course and another course served as the control group where commit frequency was simply monitored.
-Results show an increase in the frequency of commits at a statistically significant level.
-This study uses fine grained instrumentation to identify practices used between commits such as structured navigation.
-
-Dubois and Tamburrelli discuss theoretical concepts behind applying gamification to software development.
-\cite{Dubois2013Understanding} Their approach is to consider how gamification could be applied throughout the software development life-cycle to motivate many practices in different phases.
-They provide guidance on assessing gamification in the software development domain that includes analyzing the actors and game mechanics, defining integration with existing tools and processes, and designing the evaluation of results.
-Dbois and Tamburrelli demonstrate this guidance using a study of students working on a class project.
-Defining a control group and subject group of students, both groups receive the same feedback about their code comments, test coverage, code size, and static analysis rule violations.
-The subject group can see each others' rankings while the control group only sees their personal rankings.
-Results though inconclusive indicate that competition increases the desired results.
-In our study we apply both individual feedback and competitive based feedback (anonymized) to participants and seek to observe a change in their behavior after an intervention.
 
@@
 
-\subsubsection{Observations About navigation ratio improvement}
+\subsubsection{Observations About navigation ratio}
 
@@
 
-\begin{figure}\begin{mdframed}[linecolor=white]
-	\includegraphics[width=3in]{navratioboxplot.pdf}
+\begin{figure}
+	\includegraphics[width=3.3in]{navratioboxplot.pdf}
 	\caption{Navigation Ratio by Week}
 	\label{fig:navigationaverage}
-\end{mdframed}\end{figure}
+\end{figure}
 
@@
 Weeks 4 and 5 that follow the formal period of study show the navigation ratio continues in a similar distribution pattern.
+
+\subsubsection{Direct Feedback Discussions}
+
+Anil this is one place  you can discuss the findings from the post-interviews with the feedback.  It would be great if you can provide some quotes or anecdotes here.
+
+\begin{figure}
+	\includegraphics[width=3.3in]{developerEmedian.pdf}
+	\caption{Developer Analysis Against Median}
+	\label{fig:developercomparison}
+\end{figure}
+
+Following a team oriented analysis of data, we conducted one on one feedback sessions with the pilot participants.
+The participants received the feedback in the form of charts and written analysis of the charts display.
+The key analysis chart compared their distribution of time in each category to the median for the group.
+Figure \ref{fig:developercomparison} shows the chart analyzed for one developer comparing their time to the median time.
+
+We discovered during these interviews that developers did not spend the entire time in their desktop environment because they had a virtual machine environment to use for testing.
+Thus categories like testing, debug, and reading were affected by whether or not they used the virtual environment or their desktop for these activities.
+Individual differences in those categories are thus not relevant for the feedback given.
+Because our points analysis focused on navigation commands excluding reading, the data from the alternative environment does not impact the results.
+None of the developers said they used Sando in the alternative environment, thus results for points driven by Sando usage remain relevant.
+
+Defelopers also reiterated that they wanted different feedback from the tool and experience such as immediate prompts to suggest navigation commands within the context of what they are working on.
+Though commonly held opinion of Microsoft`s Clippy experience suggests this more invasive form of feedback is not always welcome, we nonetheless shall list as future work providing more immeidate suggestions on how developers can improve their navigation practices.
+
+Moving towards more immediate feedback, we shared a set of achivement badges tied to specific navigation commands  with the developers.
+Developers thought the bades were interesting and helpful in their specificity.
 
@@
 Incidentally, while cross verifying their performance on the coding, they are better productive compared to the participants on the bottom of points table.
-\begin{figure}\begin{mdframed}[linecolor=white]
-	\includegraphics[width=3in]{ElementInfluenceChart.pdf}
+\begin{figure}
+	\includegraphics[width=3.4in]{ElementInfluenceChart.pdf}
 	\caption{Influence of Feedback Elements}
 	\label{fig:elementInfluence}
-\end{mdframed}\end{figure}
+\end{figure}
 

### 2013-10-11 paper_id1.tex
Data Complete

@@
 
-\title{Experiences Gamifying Code Navigation}
-\subtitle{Are development practices a habit?}
+\title{Experiences Gamifying Developer Practices}
+%\subtitle{}
 % 
@@
 As software development practices evolve, toolsmiths face the continuous challenge of communicating new practices and tools to the community.
-Developers both like to try new things and like to stick with the familiar.
 In addition to training, discussing, and presenting software engineering practices and tools, we propose a method to both motivate and monitor their adoption through a tool embedded in the Integrated Development Environment (IDE).
@@
 Blaze monitors developer practice and tool use over the long term and reports the lasting change a communication effort makes in everyday habits.
-Embeding Blaze in the developer`s work environment as a Visual Studio extension captures developer work patterns while giving feedback to users on their practices.
+Embedding Blaze in the developer`s work environment as a Visual Studio extension captures developer work patterns while giving feedback to users on their practices.
 Blaze encourages everyone to use the new practices by assigning points to specific commands and tools.
@@
 We observed that the duration of an edit session is more correlated with categories of activity that benefit from improved practices and tools.
-The Blaze tool was well received by most participants, however, the intended improvements in navigation practices promoted by the pilot did not occur.
-We conclude that more proactive interventions are necessary to move developers to change their code navigation habits.
+The Blaze tool was well received by most participants, however, navigation practices were not significantly affected by the Blaze feedback mechanism.
+We conclude that more proactive interventions are necessary to encourage developers to change their code navigation habits.
 \end{abstract}
@@
 \section{Introduction}
-%update
 
-%As  a developer I'm focused on the problem I'm currently solving and driven by productivity and deadlines to the point that I don't have time to sharpen the saw identifying better practices.
-
-Developers with years of experience develop habits and ways of doing things that are difficult to change even when they might recognize better methods.
-For instance, the commands they use and ways they use to navigate the code base in the development environment may make them less efficient than using newer practices.
+Software developers who maintain code face additional challenges of navigating an existing code base to incorporate changes as they build new features or fix bugs.
+Over the years, developers learn techniques to analyze code and locate areas that are relevant to an enhancment.
+As new techniques and tools are introduced to the development environment, developer may not learn about them unless they interact with a peer developer who uses them\cite{wbsnipes:Hill2011Peer}.
+For instance, the commands they use to navigate the code base in the development environment may make them less efficient than using newer practices or tools.
 Robillard et.
@@
 show that developers who use commands that allow them to follow code structure are more effective and efficient than developers who browse through code \cite{wbsnipes:Robillard2004How}.
-Can we make developers more productivity by training them to use structured navigation commands and tools?
 
-Perhaps the persistence of their habit is why developers are hard to turn from their old practices.
-Developers may need some coaching to continue a new practice following an introduction.
-To continuously coach the developer, we conceived of a tool in our prior work to provide constant feedback and motivate them to adopt a new practice.
-\cite{Snipes2013Towards} This tool creates a competitive game that gives continous feedback to the developer continuously on adopting the improved practice.
-The more they change their habits, the higher they score in the game.
-%clunky
-Taking a step back, gamification is a popular movement that people apply to customer oriented web sites and social games.
+Enter the toolsmith,  who responds to challenges faced by software developers by creating tools such as code search tools that make the job easier and more efficient.
+The toolsmith needs to spread the word about advanced tools they develop and determine their effect on developer practices over the long term.
+Is there something we can do that encourages developers to adopt new practices and tools that also provides feedback to developers and toolsmiths on improvements that result?
+
+We envisioned a tool in our prior work that would motivate developers to adopt new practices or tools by providing constant feedback and allowing them to compare their improvements to how their peers are doing  \cite{Snipes2013Towards}.
+This tool creates a competitive game that gives continuous feedback to the developer on adopting the improved practice.
+The more they change their practices to adopt new techniques or tools, the higher they score in the game.
+
+Yes gamification, the incorporation of game elements into an activity that people do not typcially consider a game\cite{2013Oxford}.
 Why would software developers writing serious software be interested in this?
@@
 
-The rest of the paper is organized as follows:  Section 2 reviews the preliminary survey-based assement, Section 3 Discusses the design considerations for developing the Blaze tool/game, Section 4 Discusses the design of the study and participant selection, Section 5 is Related work, Section 6 results from the pilot, Section 7 Threats to Validity, and Section 8 has our Conclusions.
+The rest of the paper is organized as follows:  Section 2 reviews the preliminary survey-based assessment, Section 3 Discusses the design considerations for developing the Blaze tool/game, Section 4 Discusses the design of the study and participant selection, Section 5 is Related work, Section 6 results from the pilot, Section 7 Threats to Validity, and Section 8 has our Conclusions.
 
@@
 \begin{itemize}
-\item 95\% of developers responding would try tools and practices suggested by an automated recommendation system particularly if they were provided in context with the current workflow pattern the developer follows.
+\item 95\% of developers responding would try tools and practices suggested by an automated recommendation system particularly if they were provided in context with the current work-flow pattern the developer follows.
 \item Responses indicate a preference for team competition over individual awards (badges) and awards for quality practices ranked highest.
@@
 
+To evaluate developers using recommended tools in Visual Studio, we categorized them by their tool name.
+In the pilot we were interested in developers learning about and using the Sando search tool.
+Sando is part of the recommended developer tool suite for ABB and supports structured navigation using code search \cite{Shepherd2012Sando}.
+
 Gamification practices consider the feedback of achievements and points as critical to a successful outcome.
@@
 
-Using these guidelines we assigned points to structured navigation commands which received 1-10 points each while unstructured navigation commands received 0 points.
+Using these guidelines we assigned points to structured navigation commands which received 1 point each while unstructured navigation commands received 0 points.
+We gave additional emphasis to using the Sando search tool by assigning it 10 points.
 In creating levels, following the guidance from Hamari \cite{wbsnipes:Hamari2011Framework}, we built levels following an exponential curve.
@@
 After a week's data collection marking the end of the first stage, Blaze started to automatically pop-up a window containing a button for a web page with information.
-The page included a link to download another tool for code search called Sando \cite{Shepherd2012Sando} and information about using that tool in a video.
+The page included a link to download Sando  and information about using that tool in a video.
 The sub-page on navigation contained a page of tutorial on the built-in structured navigation commands in Visual Studio.
@@
 
-\subsubsection{Observations About Developers` Habits}
-
-We are interested in understanding whether developers are consistently using practices that make them inefficient.
+%\subsubsection{Observations About Developers` Habits}
+%
+%We are interested in understanding whether developers are consistently using practices that make them inefficient.
 One way to check that is to determine whether our measurements are associated with specific developers in the study, in other words, can we independently classify the edits performed by each developer.
@@
 Thus the hypothesis is developers have consistent ways of navigating and understanding code while they are making changes.
-
-To test this hypothesis we construct a classification tree using Weka \cite{Hall2009WEKA} with the J48 classifier.
+%
+%To test this hypothesis we construct a classification tree using Weka \cite{Hall2009WEKA} with the J48 classifier.
 Slicing the data and training the model using 10-fold cross-validation, the model identifies 53\% of the 647 edit sessions with the correct developer among 6 participants.
 The tree uses nearly all available fields including time for all categories and time of day for edits in its attempt to associate sessions with their developer generating a Kappa statistic of 0.403  and square root of the relative squared error at 101\%.
-
-While there is some consistency to how developers work, we cannot say that individual developers are always consistent in their work patterns.
+%
+%While there is some consistency to how developers work, we cannot say that individual developers are always consistent in their work patterns.
 Thus developers are somewhat flexible in the commands they use when navigating and making changes to code and may be more open to adopting new tools and techniques.\
@@
 The attribute search method we chose was Greedy Step-wise search that selects attributes based on their correlation with the target and stops adding attributes when the next attribute decreases the correlation.
-The process identified the top 5 out of 10 possible categories as follows:
+The process identified the top 5 out of 15 possible categories as follows:
 \begin{enumerate}
 \item Unstructured Navigation
-\item Read
-\item Build
-\item Debug
-\item Edit
+\item Debug - events from running the debugger
+\item Edit - events that modify the code
+\item Other Actions - other events in visual studio that we do not sub-classify
+\item Build - events from running a build
 \end{enumerate}
@@
 Using R\cite{Rcitation}, we constructed a linear regression model for Unstructured Navigation to determine the correlation with edit duration.
-The linear model found a positive correlation with unstructured navigation with a p-value <0.001 on 646 degrees of freedom.
-The model with only unstructured navigation had an R-squared of 0.52 showing a good portion of the variation is explained by that category.
-In order to have a more acceptable explanation of edit duration, we conducted a multiple linear regression model with both Unstructured Navigation and Read categories.
-The Read category of events captures the activities the developer is doing to look through the code in the editor.
-They are navigation commands that move incrementally like pressing the up/down arrow keys, or moving the scroll-bar an incremental amount.
-Combining these two categories results in a linear model with a p-value <0.001 and an R-squared of 0.76, more acceptable for drawing conclusions.
+The linear model found a positive correlation with unstructured navigation with a p-value <0.001 on 1694 degrees of freedom.
+The model had an Adjusted R-squared of 0.64 showing a good portion of the variation in edit duration is explained by that category.
 
-Conducting an ANOVA analysis shown in Table \ref{editDurationANOVA} for this regression model, we can explore the contribution of Unstructured Navigation and Reading to the model.
-The contribution of each variable can be determined by calculating the portion of the total variation that each variable explains.
-Unstructured Navigation explains 53\% of the variation.
-Out of the remaining variation found by subtracting the amount explained by Unstructured Navigation from the total, Read explains 49\%.
+This analysis shows that we cannot disprove that unstructured navigation is correlated with edit duration, thus it is an important factor in developer productivity.
+Combining unstructured navigation with reading describes a larger portion of the overall edit duration and motivates us towards making tools that improve both navigation methods and code comprehension.
 
-\begin{table}[!t]
-
-\begin{mdframed}[linecolor=white]
-\begin{tabular}{|l|r|r|r|r|r|}
-\hline
-                          &  Df  & SSQ & MeanSq &Fval.   & Pr(>F)    \\
-\hline
-Unstr. Nav.& 1 & 0.039& 0.03920 & 1383 & <0.001\\
-\hline
-Read                 & 1 & 0.017 & 0.01741 & 614 & <0.001\\
-\hline
-Residuals                & 645 & 0.018 & 0.00003   &  &  \\
-\hline
-Total		& 647 & 0.074 & 0.05664 & & \\
-\hline
-\end{tabular}
-\caption{ANOVA for Edit Duration}
-\label{editDurationANOVA}
-\end{mdframed}\end{table}
-
-After this analysis we cannot disprove that unstructured navigation is correlated with edit duration, thus it is an important factor in developer productivity.
-Combining unstructured navigation with reading describes a larger portion of the overall edit duration and motivates us towards improvements in both navigation methods and code comprehension.
-
-\subsubsection{Observations About navigation ratio}
+\subsubsection{Observations on Navigation Ratio}
 
@@
 \item View Call Hierarchy (Ctrl+K Ctrl+T) provides a two way analysis of a symbol's dependencies and uses. 
-\item Class View (Ctrl+W, C) priovides a searchable browser of classes and class hierarchy
-\item Find All References provides a list of refernces for an object
+\item Class View (Ctrl+W, C) provides a browser and search function for classes and class hierarchy
+\item Find All References provides a list of references for an object
 \item Navigate to Event Handler brings up the event handler from the XAML editor
 \item View Class Diagram generates a class diagram for the project
-\item View Obeject Browser is a search tool for objects in the solution
+\item View Object Browser is a search tool for objects in the solution
 \end{itemize}
@@
 
-During the pilot we staged the deployment of information feedback to developers so they would gradually improve their structured navigation practices.
-Our hypothesis was that this instant feedback and information would result in an increase in the use of structured navigation commands in Visual Studio.
+During the pilot we staged the deployment of information feedback to developers so they would first have access to learn about structured navigation and then get instant feedback in the form of points for using structured navigation commands.
+Our hypothesis was that this instant feedback and information would result in an increase in the number of structured navigation commands used in Visual Studio.
 
@@
 Expecting that as developers learn about navigation commands and tools that count towards the score in the second week they will begin using the practices more.
-Also expecting that noticing the leaderboard competition will encourage some developers to use the navigation practices more.
+In the third week we turn on points feedback and the leaderboard expecting this will encourage developers to use more structured navigation.
 
 \begin{figure}
-	\includegraphics[width=3.3in]{navratioboxplot.pdf}
+	\includegraphics[width=3.3in]{navratioboxplot_ann.pdf}
 	\caption{Navigation Ratio by Week}
@@
 What we observed visually in Figure \ref{fig:navigationaverage} is that the structured navigation practice did not increase significantly from the beginning to the end of the study.
-Overall, we did not see a significant effect towards increasing navigation ratio for the interventions in week 2 or week 3.
-Weeks 4 and 5 that follow the formal period of study show the navigation ratio continues in a similar distribution pattern.
+We evaluated the week over week differences using a Wilcoxon Rank sum test that compares week one to week two and week one to week three.
+In week two when the developers were able to learn about structured navigation commands there was no significant change.
+Also in week two, two developers stopped participating, one was working in another environment and the other dropped of for the rest of the study.
+Week three there was a slight increase, however the increase is not significant.
+Weeks four and five that follow the formal period of study show the navigation ratio continues in a similar range.
+Overall, we did not see a significant effect towards increasing navigation ratio for the interventions Blaze provided to the environment.
+
+\subsubsection{Observations on Points}
+
+The developers using Blaze received feedback in the form of points given for using structured navigation commands and using the Sando search tool\cite{Shepherd2012Sando}.
+Per the pilot design, the points display was disabled until the third week following installation so that a baseline without instant feedback could be established.
+The points display and the navigation arrow were enabled in the third week to test whether the instant feedback was successful in driving increased use of the targeted features.
+Observing the result shows that during the first two weeks, we see in Figure \ref{fig:pointsbyweek} developers were fairly consistent in their use patterns.
+With the introduction of the points feedback, two developers` points accumulation spiked in week three due mostly to increased use of Sando.
+We informed developers at the beginning of the pilot that Sando usage would receive 10 points and using built-in structured navigation commands would receive 1 point.
+
+\begin{figure}
+	\includegraphics[width=3.3in]{pointsbyweek_ann.pdf}
+	\caption{Average Points per Day by Week}
+	\label{fig:pointsbyweek}
+\end{figure}
 
 \subsubsection{Direct Feedback Discussions}
-
-Anil this is one place  you can discuss the findings from the post-interviews with the feedback.  It would be great if you can provide some quotes or anecdotes here.
 
@@
 
-Following a team oriented analysis of data, we conducted one on one feedback sessions with the pilot participants.
-The participants received the feedback in the form of charts and written analysis of the charts display.
-The key analysis chart compared their distribution of time in each category to the median for the group.
-Figure \ref{fig:developercomparison} shows the chart analyzed for one developer comparing their time to the median time.
+Following our data analysis, we conducted one on one feedback sessions with the pilot participants.
+Before initiating the discussion, we assured the individual developer that the data shared with them is confidential and we will not share the same with management.
+The participants received the feedback in the form of quantitative charts and written analysis of the information.
+The key quantitative chart shown in Figure \ref{fig:developercomparison} compared their distribution of time in each category to the median for the group.
 
-We discovered during these interviews that developers did not spend the entire time in their desktop environment because they had a virtual machine environment to use for testing.
+We discovered during these feedback sessions that developers did not spend the entire time in their desktop environment because they had a virtual machine environment available for testing.
 Thus categories like testing, debug, and reading were affected by whether or not they used the virtual environment or their desktop for these activities.
-Individual differences in those categories are thus not relevant for the feedback given.
-Because our points analysis focused on navigation commands excluding reading, the data from the alternative environment does not impact the results.
-None of the developers said they used Sando in the alternative environment, thus results for points driven by Sando usage remain relevant.
+Individual differences in those categories are considered irrelevant and not discussed here.
+Because our game points analysis focused on navigation commands excluding reading, the presence of the virtual environment may explain how the data have missing values on some days.
 
-Defelopers also reiterated that they wanted different feedback from the tool and experience such as immediate prompts to suggest navigation commands within the context of what they are working on.
-Though commonly held opinion of Microsoft`s Clippy experience suggests this more invasive form of feedback is not always welcome, we nonetheless shall list as future work providing more immeidate suggestions on how developers can improve their navigation practices.
+We also discussed with one developer who`s build effort and the underlying build frequency was much higher than median as shown in Figure \ref{fig:developercomparison}.
+We asked the developer to explain how building more frequently helps them.
+The developer mentioned that by doing more frequent build, (s)he is trying to ensure the correctness of the work completed so far.
+The developer feels that if (s)he resolves the bugs then and there, (s)he can do it faster as the code and logic is fresh in her memory.
+If they build less frequently, even though it reduces the build effort, there are more bugs each time and the developer often spends more time in identifying the problem area than fixing it.
+Hence the developer feels that more frequent builds improves their productivity.
 
-Moving towards more immediate feedback, we shared a set of achivement badges tied to specific navigation commands  with the developers.
-Developers thought the bades were interesting and helpful in their specificity.
+In another case, we observed a higher level of unstructured navigation than median and asked the developer about that.
+They explained that even though they agree that structured navigation is better than unstructured navigation, (s)he prefers unstructured navigation in some specific instances.
+The developer is the author of the code and (s)he knows the code completely.
+``I know that the line I need to fix is in XYZ method and it is 4 blocks down from where I am currently.
+It takes much less time for me to reach the fix location using unstructured navigation.
+If the code is not written by me, I will prefer structured navigation.'' This instance cautions us to gather information on aspects such as code familiarity and developer experience with the product when assessing the results of a navigation study.
+
+%Though commonly held opinion of Microsoft`s Clippy experience suggests this more invasive form of feedback is not always welcome, we shall list as future work providing more immediate suggestions on how developers can improve their navigation practices.
+Developers also reiterated that they wanted a different feedback experience from Blaze  such as immediate prompts to suggest navigation commands within the context of what they are working on.
+Moving towards more immediate feedback, we shared a set of achievement badges tied to specific navigation commands  as we gave the more quantitative feedback to the developers.
+Developers said that badges would motivate them to use some specific tools or commands more than points alone.
+This contradicts a result from the pre-study survey where respondents rated badges as less important than team or individual competition in motivating them to improve.
+The important factor in this case is badges serve as both a feedback mechanism and information showing which commands are important to use for improving structured navigation and their score.
 
@@
 
-Our post pilot survey attempts to shed some light on the reasons behind the consistency in navigation ratio over the pilot.
+We conducted an in-person post pilot survey in order to cross check the quantitative data from Blaze with developers` perception of their practices and their demographic information.
+
+The pilot survey queried developers` relevant demographic data to help understand some differences in results.
+We asked developers to provide their years of experience so that we could evaluate any effects experience might have.
+The developers in the study had between 7 and 14 years experience.
+One developer who dropped out of the study said that their lower level of experience compared to their peers influenced their decision to stop participating.
+
+We asked developers to rate their knowledge of the code they maintain.
+Four of the six developers reported that they maintain code that they wrote while the two most experienced developers maintain code that they recently learned.
+One developer reported during our individual analysis feedback that working in code they know well affects their navigation practices leading them to use more unstructured navigation.
 
@@
 Other half of the participants have some understanding on structured navigation use it some of the time.
-Even developers whose navigation ratio  was lower than average, used structured navigation occasionally as much as their more frequently using colleagues.
-While analyzing the points they scored with Blaze tool, we observed that the participants who mentioned that they have good knowledge in structured navigation are on top of the points table.
-%the following statement is not supported with data unless you have additional data?
-Incidentally, while cross verifying their performance on the coding, they are better productive compared to the participants on the bottom of points table.
+Even developers whose navigation ratio  was lower than average, used structured navigation on occasion  as much as their colleagues.
+While analyzing the points they scored with Blaze tool, we observed that the participants who rated their prior knowledge of structured navigation highly also scored higher in the pilot.
+
 \begin{figure}
@@
 Features to meet these requests are part of future work.
-
-%Though not significant, there is a loose correlation between those who rated their prior knowledge of structured navigation commands highly and their point scores.   
-There was also feedback that people wanted more information presented to them by the tool on what structured navigation commands to use, so they may not have closed the knowledge gap during the pilot.
 

### 2013-10-14 paper_id1.tex
Reworked the abstract into much shorter version.  Polished later portion of intro.  Minor wording edits elsewhere.

@@
 \begin{abstract}
+
 As software development practices evolve, toolsmiths face the continuous challenge of communicating new practices and tools to the community.
-In addition to training, discussing, and presenting software engineering practices and tools, we propose a method to both motivate and monitor their adoption through a tool embedded in the Integrated Development Environment (IDE).
-
-Blaze monitors developer practice and tool use over the long term and reports the lasting change a communication effort makes in everyday habits.
-Embedding Blaze in the developer`s work environment as a Visual Studio extension captures developer work patterns while giving feedback to users on their practices.
+We tested an idea with industrial software developers that adding game-like feedback to the development environment focused on improving practices and tool use would improve adoption of tools and practices for code navigation.
+We conducted this study in three stages, the pre-study to validate our ideas, a pilot of a tool called Blaze that implements some game features to motivte better navigation practices and use of a code search tool.
+We found that developers used the tool willingly and their feedback indicated the Blaze tool and game made them think about navigation practices more.
+We also learned that demonstrating significant improvement can be difficult and that developers want more proactive messages and recomendations on ways they can improve.
+%
+% In addition to training, discussing, and presenting software engineering practices and tools, we demonstrate a method to both motivate and monitor their adoption through a tool embedded in the Integrated Development Environment (IDE).
+%
+%We determine through a survey of 130 developers the things we can do to motivate them to adopt 
+%Blaze monitors developer practice and tool use over the long term and reports the lasting change a tool deployment or communication effort makes in everyday habits. 
+%
+% Embedding Blaze in the developer`s work environment as a Visual Studio extension captures developer work patterns while giving feedback to users on their practices.
 Blaze encourages everyone to use the new practices by assigning points to specific commands and tools.
- With a rich data source, we evaluate developer practices, define metrics that give individual feedback, and create an environment for developers to see how they are doing compared with their teammates.
-
-Data collected from the pilot group show a dichotomy of navigation and code understanding practices used by developers even the same developer in different situations.
+% With a rich data source, we evaluate developer practices, define metrics that give individual feedback, and create an environment for developers to see how they are doing compared with their teammates.
+%
+%Data collected from the pilot group show a dichotomy of navigation and code understanding practices used by developers even the same developer in different situations.
 We observed that the duration of an edit session is more correlated with categories of activity that benefit from improved practices and tools.
@@
 
-Yes gamification, the incorporation of game elements into an activity that people do not typcially consider a game\cite{2013Oxford}.
+Gamification is the incorporation of game elements into an activity that people do not typcially consider a game\cite{2013Oxford}.
 Why would software developers writing serious software be interested in this?
-Our first step was to challenge this question through a survey that queried both whether developers would be interested in gamifying their workspace and what aspects of that space would be most interesting to them.
+Our first step was to challenge this question through a pre-study survey that queried both whether developers would be interested in gamifying their workspace and what aspects of that space would be most interesting to them.
 
-With these questions answered, we identified a set of practices and defined a game meeting expectations from the developers feedback.
-The game, implemented in a Visual Studio add-in called Blaze, integrates into the Visual Studio workspace and establishes a light anonymous competition between developers to improve their navigation practices.
+After confirming a good level of interest from the pre-study survey, we identified a set of practices and defined a game meeting expectations from the developers feedback.
+The game, implemented in a Visual Studio add-in called Blaze, integrates into the Visual Studio workspace and establishes an anonymous competition between developers to improve their navigation practices.
 
 During the pilot of Blaze we observed the variation in developer navigation practices and identified some potential reasons for this variation.
-We evaluated developer's acceptance of game elements in the tool and their desires for additional functionality.
-While further data is required to assess the impact of the tool, the initial developer acceptance encourages an expanded pilot.
-%state the contribution
+We conducted post-pilot survey and feedback sessions that evaluated developers` acceptance of game elements in the tool and their desires for additional functionality.
+
+This work contributes an assessment of developers` response to game-like elments in the promotion of new tools and practices.
+We assess their early response through a survey of 130 software developers.
+Then conduct a pilot of a tool, Blaze, that introduces game elements into the developer's IDE focused on improving navigation practices.
+With the pilot feedback we determine the acceptance, influence, and effect of deploying the Blaze tool in the pilot group.
 
@@
 The pilot evaluates the aspects of game implementation as well as the study of the results of their results.
-Here we discuss how the aspects of selecting a practice suitable for gamification, designing the game, and competitive element considerations.
+Here we discuss  relevant aspects of selecting a practice suitable for gamification, designing the game, and competitive element considerations.
 

### 2013-10-14 paper_id1.tex
Edited entire paper.  Worked out threats to validity section.  Added graphic with blaze tool image.

@@
 We tested an idea with industrial software developers that adding game-like feedback to the development environment focused on improving practices and tool use would improve adoption of tools and practices for code navigation.
-We conducted this study in three stages, the pre-study to validate our ideas, a pilot of a tool called Blaze that implements some game features to motivte better navigation practices and use of a code search tool.
+We conducted this study in three stages, the pre-study to validate our ideas, a pilot of a tool called Blaze that implements some game features to motivate better navigation practices and use of a code search tool.
 We found that developers used the tool willingly and their feedback indicated the Blaze tool and game made them think about navigation practices more.
-We also learned that demonstrating significant improvement can be difficult and that developers want more proactive messages and recomendations on ways they can improve.
+We also learned that demonstrating significant improvement can be difficult and that developers want more proactive messages and recommendations on ways they can improve.
 %
@@
 %
-% Embedding Blaze in the developer`s work environment as a Visual Studio extension captures developer work patterns while giving feedback to users on their practices.
+% Embedding Blaze in the developer's work environment as a Visual Studio extension captures developer work patterns while giving feedback to users on their practices.
 Blaze encourages everyone to use the new practices by assigning points to specific commands and tools.
@@
 Software developers who maintain code face additional challenges of navigating an existing code base to incorporate changes as they build new features or fix bugs.
-Over the years, developers learn techniques to analyze code and locate areas that are relevant to an enhancment.
+Over the years, developers learn techniques to analyze code and locate areas that are relevant to an enhancement.
 As new techniques and tools are introduced to the development environment, developer may not learn about them unless they interact with a peer developer who uses them\cite{wbsnipes:Hill2011Peer}.
@@
 
-Gamification is the incorporation of game elements into an activity that people do not typcially consider a game\cite{2013Oxford}.
+Gamification is the incorporation of game elements into an activity that people do not typically consider a game\cite{2013Oxford}.
 Why would software developers writing serious software be interested in this?
@@
 During the pilot of Blaze we observed the variation in developer navigation practices and identified some potential reasons for this variation.
-We conducted post-pilot survey and feedback sessions that evaluated developers` acceptance of game elements in the tool and their desires for additional functionality.
+We conducted post-pilot survey and feedback sessions that evaluated developers' acceptance of game elements in the tool and their desires for additional functionality.
 
-This work contributes an assessment of developers` response to game-like elments in the promotion of new tools and practices.
+This work contributes an assessment of developers' response to game-like elements in the promotion of new tools and practices.
 We assess their early response through a survey of 130 software developers.
@@
 Key conclusions are that developers are more more successful finding and fixing bugs when they create a detailed plan for implementing a change, use structured navigation with keyword search or cross-reference search, and only review methods once during their search.
-This builds on this work by testing methods for increasing the use of structured navigation in developers` everyday practice.
+This builds on this work by testing methods for increasing the use of structured navigation in developers' everyday practice.
 
@@
 \subsection{Gamification}
+
+When discussing gamification, folks raise a concern about replacing an activity that is intrinsically motivated with extrinsic motivation provided by points and achievements.
+To discover what is behind this concern, several studies are useful to understand what should be done.
+
+Early studies attempt to evaluate achievement motivation in students and relate that to their performance in school.
+The study by Hermans describes results from applying questionnaire to evaluate achievement motivation in students \cite{wbsnipes:Hermans1970Questionnaire}.
+Each question is ranked for its correlation to achievement motivation.
+Some key questions about the diligence with which students approach their work are correlated with achievement motivation.
+
+Maehr proposes an affirmative theory whereby  individuals achieve as a member of a social group choosing the behavior that will meet the expectations and values of the group that is significant to them \cite{wbsnipes:MaehrCulture}.
+The presence of a social group where the person is motivated to belong and excel within is important for creating achievement motivation.
+Maehr states that since achievement behaviors can be triggered by circumstance, that they are present in most people regardless of background.
+
+Beecham and colleagues provide a thorough analysis of existing literature on studies of motivation in Software Engineering \cite{wbsnipes:Beecham2008Motivation}.
+They provide a useful inventory of aspects of the job motivate software engineers with the most common being ''the work'' motivates us.
+The list of de-motivators is also useful with common job satisfaction items like stress and inequity in recognition, plus poor quality software (low accomplishment) and lack of influence in decision making.
+The paper also lists characteristics of people in the professions including need for independence (autonomy), desire to learn new skills/tackle challenges.
 Singer and Schneider demonstrate the use of a message board and points for encouraging students to increase their frequency of commits to the source code repository \cite{Singer2012It}.
@@
 
-\section{Pre-study Sruvery on Viability of Applying Gamification for Software Engineering}
+\section{Pre-study Sruvery}
 
 \subsection{Pre-study Background}
-When discussing gamification, folks raise a concern about replacing an activity that is intrinsically motivated with extrinsic motivation provided by points and achievements.
-To discover what is behind this concern, several studies are useful to understand what should be done.
 
-Early studies attempt to evaluate achievement motivation in students and relate that to their performance in school.
-The study by Hermans describes results from applying questionnaire to evaluate achievement motivation in students \cite{wbsnipes:Hermans1970Questionnaire}.
-Each question is ranked for its correlation to achievement motivation.
-Some key questions about the diligence with which students approach their work are correlated with achievement motivation.
-
-Maehr proposes an affirmative theory whereby  individuals achieve as a member of a social group choosing the behavior that will meet the expectations and values of the group that is significant to them \cite{wbsnipes:MaehrCulture}.
-The presence of a social group where the person is motivated to belong and excel within is important for creating achievement motivation.
-Maehr states that since achievement behaviors can be triggered by circumstance, that they are present in most people regardless of background.
-
-Beecham and colleagues provide a thorough analysis of existing literature on studies of motivation in Software Engineering \cite{wbsnipes:Beecham2008Motivation}.
-It has some valuable lists of factors that motivate software engineers with the most common being ''the work'' motivates us.
-The list of de-motivators is also useful with common job satisfaction items like stress and inequity in recognition, plus poor quality software (low accomplishment) and lack of influence in decision making.
-The paper also lists characteristics of people in the professions including need for independence (autonomy), desire to learn new skills/tackle challenges.
-
-Understanding the concerns with maintaining intrinsic motivation while adding elements of achievement motivation, we developed an assessment survey questionnaire to determine whether the community would be receptive to a gamification approach to software development.
+Understanding the concerns with maintaining developers'  intrinsic motivation while adding elements of achievement motivation, we developed an assessment survey questionnaire to determine whether the ABB developer community would be receptive to a gamification approach to software development.
 We designed  the survey to assess how developers are motivated to use training, practices, and tools and whether they are open to having that information shared with their peers and managers, and being recognized with badges and points.
@@
 
-Using a question from the Saachi and Saachi survey \cite{wbsnipes:SaatchiGameification} as directly as possible, we asked developers at ABB "How interested they would be in working for a company that incorporated some aspects of games into software engineering tools as a way to increase productivity in the workplace"?
+Using a question from the Saatchi and Saatchi survey \cite{wbsnipes:SaatchiGameification} as directly as possible, we asked developers at ABB "How interested they would be in working for a company that incorporated some aspects of games into software engineering tools as a way to increase productivity in the workplace"?
 We segmented results by country in Figure \ref{fig:gamification} and found that gamification of software engineering is more interesting to developers in India, Poland, and Finland than in other major countries.
@@
 For the US, Sweden, and Switzerland the responses indicate 60-65\% are at least somewhat interested.
-The overall response from our survey is comparable to the Saachi and Saaschi survey\cite{wbsnipes:SaatchiGameification} upon which this question was based that found 75\% of 18-45 year olds were at least somewhat interested.
-%EMH why is our survey result comparable to Saatchi?
+Because we nearly identically reused their original question, we compare the overall response from our survey to the Saatchi and Saatchi survey\cite{wbsnipes:SaatchiGameification} upon which this question was based.
+Our results land on both sides of the 75\% of 18-45 year olds that Saatchi and Saatchi found were at least somewhat interested.
+
 \begin{figure}
@@
 Another question was whether sharing detailed usage information with colleagues and the company as a whole would be a concern for developers.
-A graduated scale was used for the community that would be shared with and distinction between whether that sharing was anonymous or the information was personally identifiable.
+A graduated scale was used for the scope of the shared community and two questions distinguished between whether that sharing was anonymous or the information was personally identifiable.
 Figure \ref{fig:comfortwithsharing}  shows that over 90\% of respondents are comfortable with sharing either anonymous or non-anonymous information with their team members.
@@
 This supports a conclusion that focusing on awarding quality practices would motivate developers to do them consistently.
-%EMH connection?
+
+ To summarize, the high-level conclusions from each key question are:
 \begin{itemize}
-\item 95\% of developers responding would try tools and practices suggested by an automated recommendation system particularly if they were provided in context with the current work-flow pattern the developer follows.
-\item Responses indicate a preference for team competition over individual awards (badges) and awards for quality practices ranked highest.
-So, the tool will incorporate team competition ideas and recognize quality practices.
+\item 95\% of developers responding would try tools and practices suggested by an automated recommendation system 
+\item Developers favor a team-oriented competition competition over individual awards
+\item Awards for performing good quality development practices ranked highest
 \item Over 90\% of respondents approved of sharing data with their team members
 \end{itemize}
-
+Thus, this pre-study survey provided a strong positive indicator for going forward with a software engineering tool that focused on sharing and rewarding developers for using good practices and tools.
 \section{Game Design}
@@
 %distinguish pilot from pre-study.  How does gamification apply?
-The pilot evaluates the aspects of game implementation as well as the study of the results of their results.
+
+Following the pre-study results, we began implement a game-like tool for developers to use in Visual Studio.
+We defined the goals for the tool and designed the game elements that it would include.
 Here we discuss  relevant aspects of selecting a practice suitable for gamification, designing the game, and competitive element considerations.
@@
 \subsection{Select a Practice}
-%transition
+The first step is to select a development practice to incorporate into the game.  
+Selecting a practice to study involves identifying a practice, defining specific characteristics of the practice to measure, and identifying the actions in Visual Studio that relate to the practice.
 In their paper on game design patterns, Hamari et.
@@
 issue a warning that assigning achievements to required tasks reduces intrinsic motivation because player autonomy is reduced by the achievements \cite{wbsnipes:Hamari2011Framework} .
-This warning indicates that we should avoid giving points for things that are already an expected deliverable such as finding or fixing bugs, completing development tasks, or meeting required process criteria.
-%passive
-Practices to consider are those practices with room to go beyond a checklist or minimum criteria where developers have the opportunity to excel.
-%passive
-Selecting a practice to study involves identifying a practice, defining specific characteristics of the practice to measure, and identifying the actions in Visual Studio that relate to the practice.
-We currently limit ourselves to practices that exist entirely in the IDE thus can be monitored using Blaze.
+Thus we avoided giving points for actions like fixing bugs or completing tasks.
+We also currently limit ourselves to practices that exist entirely in the IDE thus can be monitored using Blaze.
 We identified potential practices of test-driven development, debugging practices, navigation practices,  eliminating static analysis bugs, and check-in frequency.
 %EMH is check-in frequency a practice?
+
 Practices of check-in frequency, static analysis bug elimination are highlighted in this paper's related work section with prior gamification studies in classroom environments.
-Johnson and Kou achieved automated monitoring of test-driven development practices with Zorro thus may provide a candidate for future gamification application.
-%reference
+Johnson and Kou achieved automated monitoring of test-driven development practices with Zorro \cite{V:Johnson2007Automated} thus may provide a candidate for future gamification application.
+%reference for debugging?
 Debugging practices have developer community opinions on optimal methods, however, detailed debug practices with modern IDEs have not received research community attention, thus are also reserved for future work.
 
-Based on studies of structured navigation by Robillard et.
-al.
-that identify structured navigation as a practice leading to effective program maintenance  \cite{wbsnipes:Robillard2004How}, we identified structured navigation as a key practice.
-Structured navigation fits the requirement of a practice that is contained in the IDE, is something that has proven benefits, and lends itself to measurement and scoring, is not an assigned task, and leaves room for developers to excel.
+Studies of developer navigation by Robillard et.
+al.\cite{wbsnipes:Robillard2004How} identify structured navigation as a practice that more effective developers use when maintaining programs.
+Structured navigation fits the requirements of a practice that is contained in the IDE, has proven benefits, lends itself to measurement and scoring, is not an assigned task, and leaves room for developers to excel.
+Thus we selected structured navigation as the practice to demonstrate gameification.
 
@@
 We categorize commands as structured navigation when they allow the developer to locate a code element following the program structure from another element.
-For example, Navigate To \textit{Ctrl+,}, Go To Definition \textit{F12}, and Find All References \textit{Ctrl+K,R} are commands that follow structure of the code, which we classify as structured navigation.
+These commands include Navigate To \textit{Ctrl+,}, Go To Definition \textit{F12}, and Find All References \textit{Ctrl+K,R} that are built-in to Visual Studio and follow the structured graph of the code.
 
@@
 To evaluate developers using recommended tools in Visual Studio, we categorized them by their tool name.
-In the pilot we were interested in developers learning about and using the Sando search tool.
+In the pilot we wanted developers to learn about and adopt the Sando search tool.
 Sando is part of the recommended developer tool suite for ABB and supports structured navigation using code search \cite{Shepherd2012Sando}.
@@
 Therefore the assigned letters could even change when an individual's ranking changed.
-Not knowing who of your colleagues was ahead of you could be a factor in motivation, but also avoids one potential source of conflict in an intact team.
+Not knowing who of your colleagues was ahead of you could be a factor that reduced motivation, but also avoids one potential source of conflict in an intact team.
 
@@
 In the third stage developers received instant feedback from Blaze on the points they accumulated during the development session.
-The up/down arrow provided a simple indicator on whether they were increasing their use of structured navigation or were using it less than their personal historical average.
-This final configuration continued until the end of the pilot.
+The tool's appearance at this stage is shown in Figure \ref{fig:blazeWindow}.
+The Info button provides a level indicator via colored contents and activates a chart display windows when clicked.
+Points shows the points accumulated for the developer's history.
+Level is the level of the game achieved based on an exponentially growing curve.
+The developer could click on the Leaderbaord tab to see how their scores compared with the others in the pilot.
+This final configuration remained as the display until the end of the pilot.
+
+\begin{figure}
+	\centering
+	\includegraphics[width=2.2in]{blazeWindow.png}
+	\caption{Blaze Tool Window}
+	\label{fig:blazeWindow}
+\end{figure}
 
@@
 Researchers were careful to create the system so that developers controlled their identification with the data and fortunately, management bought into the necessary anonymity of the participants.
-%Through this approach, we ensured that the developers are performing the task on a very natural environment with out bothering that they are being tracked.
+
 \subsection{Pilot Survey}
 
-The goal for the pilot follow-up survey was to collect feedback on the tool and the effect it had on the developers` navigation practices and knowledge.
+The goal for the pilot follow-up survey was to collect feedback on the tool and the effect it had on the developers' navigation practices and knowledge.
 The survey questions were segmented into portions that evaluate prior and post knowledge and practices for navigation, key steps in the pilot itself, the influence of game elements on their navigation practices, typical tasks and demographic information.
 
-The key questions about this pilot were did we have an effect on knowledge of the developers regarding structured navigation commands.
-We asked them to rate their knowledge of structured navigation commands prior to the pilot on a 1-10 scale to understand their baseline.
-We asked them to contrast their knowledge after the pilot to assess what they learned from the pilot.
+To find out whether Blaze had an effect on knowledge of the developers regarding structured navigation commands, we asked them to rate their knowledge of structured navigation commands prior to the pilot on a 1-10 scale to understand their baseline.
+Then we asked them to contrast their knowledge after the pilot to assess what they learned from the pilot.
 
-The other key question is did we have an effect on their navigation practices.
+The other key question is did Blaze have a perceived effect on their navigation practices.
 Here too we asked them to rate how much they used structured navigation prior to the pilot using a 4 choice rating of not at all, some, many times, and as much as possible.
 Then we asked whether the used structured navigation commands about the same amount, more,  or a lot more during the pilot.
-These two questions assessed their intent in terms of using structured navigation commands and we checked that with the data.
+These two questions assessed their perceptions of their structured navigation command use that we compared with the data collected from Blaze.
 
-A key set of questions focused on whether using Blaze increased their attention to navigation commands by asking how much specific game elements influenced their practices.
+The final assessment questions tested whether using Blaze influenced their attention to navigation commands by asking how much specific game elements influenced their practices.
 The scale started with not at all, a little, some, to a lot.
@@
 \section{Results}
+
+\subsection{Pilot Survey Results}
+
+We conducted an in-person post pilot survey in order to cross check the quantitative data from Blaze with developers' perception of their practices and their demographic information.
+
+The pilot survey queried developers' relevant demographic data to help understand some differences in results.
+We asked developers to provide their years of experience so that we could evaluate any effects experience might have.
+The developers in the study had between 7 and 14 years experience.
+One developer who dropped out of the study said that their lower level of experience compared to their peers influenced their decision to stop participating.
+
+We asked developers to rate their knowledge of the code they maintain.
+Four of the six developers reported that they maintain code that they wrote while the two most experienced developers maintain code that they recently learned.
+One developer reported during our individual analysis feedback that working in code they know well affects their navigation practices leading them to use more unstructured navigation.
+
+On the assessment of structured navigation knowledge, half of the participants replied that they had very good prior knowledge about structured navigation and they use structured navigation as much as possible.
+Other half of the participants have some understanding on structured navigation use it some of the time.
+Even developers whose navigation ratio  was lower than average, used structured navigation occasionally as much as their colleagues.
+While analyzing the points they scored with Blaze tool, we observed that the participants who rated their prior knowledge of structured navigation highly also scored higher in the pilot.
+
+\begin{figure}
+	\includegraphics[width=3.4in]{ElementInfluenceChart.pdf}
+	\caption{Influence of Feedback Elements}
+	\label{fig:elementInfluence}
+\end{figure}
+
+The chart in figure \ref{fig:elementInfluence} shows how the individual feedback features influenced developers to think about using structured navigation.
+Four out of six developers said the leaderboard influenced them "a lot" and all rated the leaderboard as having some influence.
+Three out of six  said their individual point score influenced them "a lot" and all said the score had some influence.
+The element with mixed ratings was a graphical indicator on whether they were improving over past use of structured navigation commands.
+The indicator received the influencing "a lot" rating from one participant and a "not at all" influence rating from one participant with the rest having some influence.
+The results indicate the more obvious and more clearly comparative information in the leaderboard was the strongest influence for developers.
+
+The participants requested that Blaze provide more feedback on tips to help them increase their score.
+They indicated that more obvious feedback such as hints that pop-up when they launch the tool would help them more than passively provided information.
+Their desire for personal metrics and historical views of the data encourages us that this detailed level of measurement and analysis is perceived as helpful.
+Features to meet these requests are part of future work.
 
@@
 
-%\subsubsection{Observations About Developers` Habits}
+%\subsubsection{Observations About Developers' Habits}
 %
@@
 al\cite{wbsnipes:Robillard2004How} as correlated with developers who effectively completed a maintenance task.
-Are we focusing on the right area of developers` activity in Visual Studio?
+In this analysis we ask whether we are focusing on the right area of developers' activity in Visual Studio?
 
@@
 
-This analysis shows that we cannot disprove that unstructured navigation is correlated with edit duration, thus it is an important factor in developer productivity.
-Combining unstructured navigation with reading describes a larger portion of the overall edit duration and motivates us towards making tools that improve both navigation methods and code comprehension.
+This analysis shows that we cannot disprove that unstructured navigation is correlated with edit duration, thus we conclude that it is an important factor in developer productivity.
 
@@
 What we observed visually in Figure \ref{fig:navigationaverage} is that the structured navigation practice did not increase significantly from the beginning to the end of the study.
-We evaluated the week over week differences using a Wilcoxon Rank sum test that compares week one to week two and week one to week three.
+We evaluated the week over week differences using a Wilcoxon Rank sum test \cite{RefWorks:118} that compares week one to week two and week one to week three.
 In week two when the developers were able to learn about structured navigation commands there was no significant change.
-Also in week two, two developers stopped participating, one was working in another environment and the other dropped of for the rest of the study.
-Week three there was a slight increase, however the increase is not significant.
+Also in week two, two developers stopped participating in the study.
+One was working in another environment and returned, while the other dropped for the rest of the study.
+Week three there was a slight increase in the mean, however the increase is not significant.
 Weeks four and five that follow the formal period of study show the navigation ratio continues in a similar range.
@@
 The points display and the navigation arrow were enabled in the third week to test whether the instant feedback was successful in driving increased use of the targeted features.
-Observing the result shows that during the first two weeks, we see in Figure \ref{fig:pointsbyweek} developers were fairly consistent in their use patterns.
-With the introduction of the points feedback, two developers` points accumulation spiked in week three due mostly to increased use of Sando.
+The result shows that during the first two weeks shown in Figure \ref{fig:pointsbyweek}, developers were fairly consistent in their use patterns.
+With the introduction of the points feedback, two developers' points accumulation spiked in week three due mostly to increased use of Sando.
 We informed developers at the beginning of the pilot that Sando usage would receive 10 points and using built-in structured navigation commands would receive 1 point.
+In these results we see an initial though not significant effect from the points display in Blaze that is not sustained over the remaining weeks.
+We found following discussions that developers wanted more guidance and more prompting included in the game which we discuss further in the next section.
 
@@
 
+Following our data analysis, we conducted one on one feedback sessions with the pilot participants.
+Before initiating the discussion, we assured the individual developer that the data shared with them is confidential and we will not share the same with management.
+The participants received the feedback in the form of quantitative charts and written analysis of the information.
+The key quantitative chart shown in Figure \ref{fig:developercomparison} compared their distribution of time in each category to the median for the group.
+
 \begin{figure}
@@
 
-Following our data analysis, we conducted one on one feedback sessions with the pilot participants.
-Before initiating the discussion, we assured the individual developer that the data shared with them is confidential and we will not share the same with management.
-The participants received the feedback in the form of quantitative charts and written analysis of the information.
-The key quantitative chart shown in Figure \ref{fig:developercomparison} compared their distribution of time in each category to the median for the group.
+We discovered during these feedback sessions that developers did not spend the entire time in their desktop environment because they had a virtual machine environment available for testing.
+Thus categories like testing and debugging were affected by whether or not they used the virtual environment or their desktop for these activities.
+Individual differences in those categories are considered irrelevant and not discussed here.
+The presence of the virtual environment may explain how some developers have no activity recorded on some days.
 
-We discovered during these feedback sessions that developers did not spend the entire time in their desktop environment because they had a virtual machine environment available for testing.
-Thus categories like testing, debug, and reading were affected by whether or not they used the virtual environment or their desktop for these activities.
-Individual differences in those categories are considered irrelevant and not discussed here.
-Because our game points analysis focused on navigation commands excluding reading, the presence of the virtual environment may explain how the data have missing values on some days.
-
-We also discussed with one developer who`s build effort and the underlying build frequency was much higher than median as shown in Figure \ref{fig:developercomparison}.
+%Can we get a quote here?
+One developer's build effort and the underlying build frequency was much higher than median as shown in Figure \ref{fig:developercomparison}.
 We asked the developer to explain how building more frequently helps them.
-The developer mentioned that by doing more frequent build, (s)he is trying to ensure the correctness of the work completed so far.
-The developer feels that if (s)he resolves the bugs then and there, (s)he can do it faster as the code and logic is fresh in her memory.
+The developer reported that by doing more frequent build, (s)he is trying to ensure the correctness of the work completed so far.
+The developer feels that (s)he resolves the bugs faster with frequent builds because the code and logic is fresh in her memory.
 If they build less frequently, even though it reduces the build effort, there are more bugs each time and the developer often spends more time in identifying the problem area than fixing it.
@@
 
-%Though commonly held opinion of Microsoft`s Clippy experience suggests this more invasive form of feedback is not always welcome, we shall list as future work providing more immediate suggestions on how developers can improve their navigation practices.
+%Though commonly held opinion of Microsoft's Clippy experience suggests this more invasive form of feedback is not always welcome, we shall list as future work providing more immediate suggestions on how developers can improve their navigation practices.
 Developers also reiterated that they wanted a different feedback experience from Blaze  such as immediate prompts to suggest navigation commands within the context of what they are working on.
@@
 Developers said that badges would motivate them to use some specific tools or commands more than points alone.
-This contradicts a result from the pre-study survey where respondents rated badges as less important than team or individual competition in motivating them to improve.
-The important factor in this case is badges serve as both a feedback mechanism and information showing which commands are important to use for improving structured navigation and their score.
-
-\subsection{Pilot Survey Results}
-
-We conducted an in-person post pilot survey in order to cross check the quantitative data from Blaze with developers` perception of their practices and their demographic information.
-
-The pilot survey queried developers` relevant demographic data to help understand some differences in results.
-We asked developers to provide their years of experience so that we could evaluate any effects experience might have.
-The developers in the study had between 7 and 14 years experience.
-One developer who dropped out of the study said that their lower level of experience compared to their peers influenced their decision to stop participating.
-
-We asked developers to rate their knowledge of the code they maintain.
-Four of the six developers reported that they maintain code that they wrote while the two most experienced developers maintain code that they recently learned.
-One developer reported during our individual analysis feedback that working in code they know well affects their navigation practices leading them to use more unstructured navigation.
-
-On the assessment of structured navigation knowledge, half of the participants replied that they had very good prior knowledge about structured navigation and they use structured navigation as much as possible.
-Other half of the participants have some understanding on structured navigation use it some of the time.
-Even developers whose navigation ratio  was lower than average, used structured navigation on occasion  as much as their colleagues.
-While analyzing the points they scored with Blaze tool, we observed that the participants who rated their prior knowledge of structured navigation highly also scored higher in the pilot.
-
-\begin{figure}
-	\includegraphics[width=3.4in]{ElementInfluenceChart.pdf}
-	\caption{Influence of Feedback Elements}
-	\label{fig:elementInfluence}
-\end{figure}
-
-The chart in figure \ref{fig:elementInfluence} shows how the individual feedback features influenced developers to think about using structured navigation.
-Four out of six developers said the leaderboard influenced them "a lot" and all rated the leaderboard as having some influence.
-Three out of six  said their individual point score influenced them "a lot" and all said the score had some influence.
-The element with mixed ratings was a graphical indicator on whether they were improving over past use of structured navigation commands.
-The indicator received the influencing "a lot" rating from one participant and a "not at all" influence rating from one participant with the rest having some influence.
-The results indicate the more obvious and more clearly comparative information in the leaderboard was the strongest influence for developers.
-
-The participants requested that the system to provide more feedback on tips to help them increase their score.
-They indicated that more obvious feedback such as hints that pop-up when they launch the tool would help them more than passively provided information.
-Their desire for personal metrics and historical views of the data encourages us that this detailed level of measurement and analysis is perceived as helpful.
-Features to meet these requests are part of future work.
+%  This contradicts a result from the pre-study survey where respondents rated badges as less important than team or individual competition in motivating them to improve.
+The important difference in this case is badges serve as both a feedback mechanism and information showing which commands are important to use for improving structured navigation and their score.
 
 \section{Threats to Validity}
-There are several threats to validity for this result.
-By making the purpose of our measurement and contest known although without details we may have triggered the participants to think about navigation commands they use before establishing a baseline.
-We attempt to control for this by comparing the pilot participants' data to data from Blaze users who are knowledgeable Visual Studio developers but did not receive the interventions.
-Another threat is the intact team represents people with different backgrounds and experience levels who may have established habits for navigating through source code.
-We attempt to control for this by asking post survey questions of the participants about their familiarity with the code they work in and their experience levels.
-The threat of non- generalizable of results remains particularly because the pilot was conducted within an intact team in a single industrial site in India.
-As the pre-study questionnaire shows, the developer's desirability for a gamification type of intervention vary by country.
+The threats to construct validity occur when measurements may not reflect the intended operational definition.The key measures in our study that could have results differing from their definition were navigation ratio and points scored in the game.
+
+Measuring navigation ratio involved categorizing events into structured and unstructured navigation.
+The categorization of events is specific to this study and documented herein, however, it may not be accurate to the intent of the operational definition of structure navigation depending on how the developer uses some commands like Find.
+Another factor affecting this measurement may occur when developers repeat a command such as hitting the down arrow key many times, we might consider that something other than navigation (though it is classified as such for this paper).
+
+The points measurement is more susceptible to influences of simple developer activity compared to navigation ratio.
+Thus a very busy developer could accumulate more points using a lower navigation ratio than a developer who has less activity performed more efficiently.
+
+Threats to internal validity are those conditions that create alternative reasons for the conclusions, or confound the conclusion.  
+In this study we may experience the Hawthorne effect where improvement occurs because we are seeking a change in a particular practice.
+Specifically, it could affect the study because a presentation and demonstration to the pilot organization was required in advance of deployment.
+By making the purpose of our measurement and contest known,  we may have triggered the participants to think about navigation commands they use before establishing a baseline.
+Because we focus on the same participants with the same behavior, they may have learned how to demonstrate improvement at some time other than when presented with the information in Blaze.
+This is mitigated by the data showing that some participants did not improve over the course of the study while others showed at least that they were attempting to improve.
+
+Participants in the pilot were not randomly selected, they are all volunteers from one intact team that was chosen due to their advanced practices and understanding of metrics.
+Therefore the participants may be predisposed to have an interest in the kind of tool and methodology we are piloting and have bias towards a favorable opinion.
+%We know that not all participants remained in the study or continued using Blaze, however, our conclusions are based mostly on those that continued with the study.
+
+Another threat is the intact team may be more familiar with the code they maintain thus be less likely to change their navigation practices.
+We ask post survey questions of the participants about their familiarity with the code they work in and their experience levels.
+The post-pilot survey responses confirmed some cases where developers are not likely to change their navigation practices such as when they work in code they are very familiar with.
+
+Threats to external validity include that the study was conducted in one development location of one company, other companies and locations may have different cultures incompatible with this type of monitoring and gamification activity.
+ Threats to generalizable results exist because we conducted the pilot of the tool with between 4 and 6 active developers thus the population size is too small to draw statistically significant inferences from.
+
+ The pilot was conducted in India, the country with the most positive response in our pre-study survey, perhaps where this activity was most acceptable.
+Thus other countries who rated their view of gamification lower may be less receptive.
+
+Participants in the pilot were professional software engineers with at least 7 years of experience.
+In other settings where developers have less experience or more the results may differ.
+The pre-study survey, experience was not part of the question set thus we could have results skewed to a particular age group.
 
@@
 \section{Acknowledgments}
-The authors wish to thank ABB and the pilot participants for their support of this research.
+The authors wish to thank ABB and the pilot participants in India for their support of this research.
 

### 2013-10-16 paper_id1.tex
completed good draft of introduction and abstract

@@
 \usepackage{mdframed}
+\usepackage{enumitem}
 %\usepackage{mdwmath}
@@
 \alignauthor
-Emerson Murphy-Hill
-\affaddr{North Carolina State University}\\
-\affaddr{Department of Computer Science}\\
+Emerson Murphy-Hill\\
+\affaddr{NC State University}\\
+\affaddr{Computer Science}\\
 \affaddr{Raleigh, N.C. USA}\\
@@
 We tested an idea with industrial software developers that adding game-like feedback to the development environment focused on improving practices and tool use would improve adoption of tools and practices for code navigation.
-We conducted this study in three stages, the pre-study to validate our ideas, a pilot of a tool called Blaze that implements some game features to motivate better navigation practices and use of a code search tool.
-We found that developers used the tool willingly and their feedback indicated the Blaze tool and game made them think about navigation practices more.
-We also learned that demonstrating significant improvement can be difficult and that developers want more proactive messages and recommendations on ways they can improve.
+We conducted this study in three stages: a pre-study to validate our ideas, a pilot of a tool called Blaze that implements some game features to motivate better navigation practices, and a post-pilot survey conducted via interviews.
+We found that a software engineering game with competitive elements influences developers towards changing their practices.
+%We also learned that demonstrating significant improvement can be difficult and that developers want more proactive messages and recommendations on ways they can improve.
 %
@@
 
-Software developers who maintain code face additional challenges of navigating an existing code base to incorporate changes as they build new features or fix bugs.
+Sascha is a developer with years of experience maintaining code who knows how to navigate through code fairly well.
+Sascha heard about some other navigation tools, but kept using the built-in features of the Visual Studio IDE that served him well.
+When Sascha's company setup a game around code navigation in Visual Studio, he became interested in being a part of the fun.
+The game gave bonus points for using Sando, a new code search tool, so he started using it instead of some of the built-in search commands.
+Sascha soon topped the game's leaderboard and discovered that motivated him to try the next tool to became part of the game.
+
+Like Sascha, software developers who maintain code face additional challenges of navigating an existing code base to incorporate changes as they build new features or fix bugs.
 Over the years, developers learn techniques to analyze code and locate areas that are relevant to an enhancement.
-As new techniques and tools are introduced to the development environment, developer may not learn about them unless they interact with a peer developer who uses them\cite{wbsnipes:Hill2011Peer}.
-For instance, the commands they use to navigate the code base in the development environment may make them less efficient than using newer practices or tools.
-Robillard et.
-al.
+As new practices and tools get introduced, developers may not learn about them unless they interact with a peer developer who uses them\cite{wbsnipes:Hill2011Peer}.
+%For instance, the commands they use to navigate the code base in the development environment may make them less efficient than using newer practices or tools.
+Robillard et al.
 show that developers who use commands that allow them to follow code structure are more effective and efficient than developers who browse through code \cite{wbsnipes:Robillard2004How}.
 
-Enter the toolsmith,  who responds to challenges faced by software developers by creating tools such as code search tools that make the job easier and more efficient.
+Enter the toolsmith,  who responds to developers' challenges by creating tools such as code search tools that make the job easier and more efficient.
 The toolsmith needs to spread the word about advanced tools they develop and determine their effect on developer practices over the long term.
-Is there something we can do that encourages developers to adopt new practices and tools that also provides feedback to developers and toolsmiths on improvements that result?
+Is there something we can do that encourages developers to adopt new practices and tools that also provides feedback to developers and toolsmiths on the resulting improvements?
 
+One option for motivating people to adopt new practices is to apply elements of game mechanics or ``gamify'' the practice.
+Gamification is the incorporation of game elements into an activity that people do not typically consider a game\cite{2013Oxford}.
 We envisioned a tool in our prior work that would motivate developers to adopt new practices or tools by providing constant feedback and allowing them to compare their improvements to how their peers are doing  \cite{Snipes2013Towards}.
-This tool creates a competitive game that gives continuous feedback to the developer on adopting the improved practice.
+This tool called Blaze, creates a competitive game that gives continuous feedback to the developer on adopting the improved practice.
 The more they change their practices to adopt new techniques or tools, the higher they score in the game.
 
-Gamification is the incorporation of game elements into an activity that people do not typically consider a game\cite{2013Oxford}.
-Why would software developers writing serious software be interested in this?
-Our first step was to challenge this question through a pre-study survey that queried both whether developers would be interested in gamifying their workspace and what aspects of that space would be most interesting to them.
-
-After confirming a good level of interest from the pre-study survey, we identified a set of practices and defined a game meeting expectations from the developers feedback.
-The game, implemented in a Visual Studio add-in called Blaze, integrates into the Visual Studio workspace and establishes an anonymous competition between developers to improve their navigation practices.
-
-During the pilot of Blaze we observed the variation in developer navigation practices and identified some potential reasons for this variation.
-We conducted post-pilot survey and feedback sessions that evaluated developers' acceptance of game elements in the tool and their desires for additional functionality.
-
-This work contributes an assessment of developers' response to game-like elements in the promotion of new tools and practices.
-We assess their early response through a survey of 130 software developers.
-Then conduct a pilot of a tool, Blaze, that introduces game elements into the developer's IDE focused on improving navigation practices.
+Would software developers writing serious software be motivated by this gamification concept?
+To answer this question, this work contributes an assessment of developers' response to game-like elements tied to their use of new tools and practices.
+We assess their response to the idea of gamification and relevant concerns through a survey of 130 software developers.
+Then we conduct a pilot of a tool, Blaze, that introduces game elements into the developer's IDE focused on improving navigation practices.
 With the pilot feedback we determine the acceptance, influence, and effect of deploying the Blaze tool in the pilot group.
 
-The rest of the paper is organized as follows:  Section 2 reviews the preliminary survey-based assessment, Section 3 Discusses the design considerations for developing the Blaze tool/game, Section 4 Discusses the design of the study and participant selection, Section 5 is Related work, Section 6 results from the pilot, Section 7 Threats to Validity, and Section 8 has our Conclusions.
+The rest of this paper is organized as follows:  Section 2 reviews Related work, Section 3 reviews results from the pre-study survey, Section 4 Discusses the design considerations for developing the Blaze tool/game, Section 5 Discusses the design of the Blaze pilot study and participant selection,  Section 6 results from the pilot study and post-pilot survey, Section 7 Threats to Validity, and Section 8 has our conclusions.
+%This is precisely what we set out to study for this work. 
+%
+%Our first step was to challenge this question through a pre-study survey that queried both whether developers would be interested in gamifying their workspace and what aspects of that space would be most interesting to them.
+%
+%After confirming a good level of interest from the pre-study survey, we identified a set of practices and defined a game meeting expectations from the developers feedback.
+The game, implemented in a Visual Studio add-in called Blaze, integrates into the Visual Studio workspace and establishes an anonymous competition between developers to improve their navigation practices.
+%
+%During the pilot of Blaze we observed the variation in developer navigation practices and identified some potential reasons for this variation.
+We conducted post-pilot survey and feedback sessions that evaluated developers' acceptance of game elements in the tool and their desires for additional functionality.
 
@@
 Robillard, Coelho and Murphy explore hypotheses around how developers can be more effective at performing a maintenance task \cite{wbsnipes:Robillard2004How}.
-Key conclusions are that developers are more more successful finding and fixing bugs when they create a detailed plan for implementing a change, use structured navigation with keyword search or cross-reference search, and only review methods once during their search.
+Key conclusions are that developers are more successful finding and fixing bugs when they create a detailed plan for implementing a change, use structured navigation with keyword search or cross-reference search, and only review methods once during their search.
 This builds on this work by testing methods for increasing the use of structured navigation in developers' everyday practice.
@@
 The Pro Metrics (PROM) tool provides a framework for collecting data for further analysis from tools used by developers.\cite{Coman2009Casestudy}  It provides a flexible data model and a plug-in architecture to facilitate collection from different data sources.
-Studies conducted using prom include a series of studies by it's creators on trends in time spent Pair Programming \cite{Coman2008Investigating}, benefits of refactoring on productivity \cite{Moser2008Case}, impact of refactoring on re-usability \cite{Moser2006Does}, and prediction of effort \cite{Abrahamsson2007Effort}.
+Studies conducted using PROM include a series of studies by it's creators on trends in time spent Pair Programming \cite{Coman2008Investigating}, benefits of refactoring on productivity \cite{Moser2008Case}, impact of refactoring on re-usability \cite{Moser2006Does}, and prediction of effort \cite{Abrahamsson2007Effort}.
 The PROM tool and studies that use it track the time spent editing files and methods from the IDE and code metrics obtained trough source code analysis.
@@
 
-Murphy-Hill et.
-al.
+Murphy-Hill et al.
 study a large usage history data set and apply several different algorithms to accurately suggest to users new commands in their environment to use.\cite{MurphyHill2012Improving} Algorithms based on command history performed well particularly when synchronized chronologically with the recipient's usage history.
 Novice users were better served by recommendation algorithms that did not require a long history, where more expert users benefited from more sophisticated algorithms that included more history.
-The "Most Widely Used" algorithm that recommends commands based on the collective usage profile of the team performed nearly as well as the more sophisticated algorithms.
+The ``Most Widely Used'' algorithm that recommends commands based on the collective usage profile of the team performed nearly as well as the more sophisticated algorithms.
 The paper presents the methodology for making recommendations and the evaluation of the accuracy of the recommendations.
@@
 
-\section{Pre-study Sruvery}
+\section{Pre-study Survey}
 
@@
 
-Understanding the concerns with maintaining developers'  intrinsic motivation while adding elements of achievement motivation, we developed an assessment survey questionnaire to determine whether the ABB developer community would be receptive to a gamification approach to software development.
-We designed  the survey to assess how developers are motivated to use training, practices, and tools and whether they are open to having that information shared with their peers and managers, and being recognized with badges and points.
-The developer study conducted with over 130 software developers at ABB challenged whether achievement motivation would help software engineers and whether they would accept sharing data on tools and practices they use to help others.
+Understanding the concerns with maintaining developers'  intrinsic motivation while adding elements of achievement motivation, we conducted an assessment survey questionnaire to determine whether the ABB developer community would be receptive to a gamification approach to software development.
+Another question asked about ways of motivating developers to use training, practices, and tools.
+We checked on who they would be comfortable sharing data from their development activities with.
+Finally we asked what kind of awards would be most welcome such as badges and points.
+The developer survey received over 130 responses from software developers at ABB.
 
@@
 
-Using a question from the Saatchi and Saatchi survey \cite{wbsnipes:SaatchiGameification} as directly as possible, we asked developers at ABB "How interested they would be in working for a company that incorporated some aspects of games into software engineering tools as a way to increase productivity in the workplace"?
+Using a question from the Saatchi and Saatchi survey \cite{wbsnipes:SaatchiGameification} as directly as possible, we asked developers at ABB ``How interested they would be in working for a company that incorporated some aspects of games into software engineering tools as a way to increase productivity in the workplace''?
 We segmented results by country in Figure \ref{fig:gamification} and found that gamification of software engineering is more interesting to developers in India, Poland, and Finland than in other major countries.
@@
 Another question was whether sharing detailed usage information with colleagues and the company as a whole would be a concern for developers.
-A graduated scale was used for the scope of the shared community and two questions distinguished between whether that sharing was anonymous or the information was personally identifiable.
+The question provided a graduated scale for the scope of who data were shared with and divided the questions between whether that sharing was anonymous or not.
 Figure \ref{fig:comfortwithsharing}  shows that over 90\% of respondents are comfortable with sharing either anonymous or non-anonymous information with their team members.
 Their comfort level decreases as the scope of who the information is shared with increases.
-People were less comfortable sharing with anyone (the category for people outside ABB) particularly if the data were not anonymous.
-Two conclusions from this feedback are that people are very willing to share information that could help their team.
+People were less comfortable sharing with anyone (the category for people outside ABB) particularly if the data were not kept anonymous.
+The first conclusion from these questions on sharing data is that people are very willing to share information that could help their team.
 Second, sharing within the company is acceptable if we take care to make the data anonymous.
@@
 
-Developers were asked what would motivate them to use tools and practices and given 4 choices to rank.
+We asked developers what would motivate them to use tools and practices and gave them 4 choices to rank.
 The top motivator for using tools and practices is coworker recommendation with 75\% of respondents in Figure \ref{fig:toolandpracticemotivators}   ranking in 1st or 2nd place.
-The lowest was having badges posted on their social profile where 33\% ranked this in 1st or 2nd place.
+The lowest was having badges posted on their social network profile where 33\% ranked this in 1st or 2nd place.
 Badges also ranked low in motivations for taking more software engineering training.
@@
 
-Developers were asked to rank awards for different actions by which they would find the most desirable.
+Developers also ranked the desirability of awards for different actions.
 Overall 68\% of developers ranked receiving an award for consistently performing quality practices as either 1st or 2nd place.
-This supports a conclusion that focusing on awarding quality practices would motivate developers to do them consistently.
+This supports a conclusion that focusing on awarding good practices would motivate developers to do them consistently.
 
  To summarize, the high-level conclusions from each key question are:
-\begin{itemize}
+\begin{itemize}[itemsep=0mm]
 \item 95\% of developers responding would try tools and practices suggested by an automated recommendation system 
-\item Developers favor a team-oriented competition competition over individual awards
+\item Developers favor a team-oriented competition over individual awards
 \item Awards for performing good quality development practices ranked highest
@@
 \end{itemize}
-Thus, this pre-study survey provided a strong positive indicator for going forward with a software engineering tool that focused on sharing and rewarding developers for using good practices and tools.
+Thus, this pre-study survey provided a strong positive indicator for going forward with a software engineering tool that focused on sharing developer data and rewarding developers for using good practices and tools.
+
 \section{Game Design}
 
-%\begin{enumerate}
+%\begin{enumerate}[itemsep=0mm]
 %\item Define the practice we wish to measure compliance with
@@
 
-Following the pre-study results, we began implement a game-like tool for developers to use in Visual Studio.
-We defined the goals for the tool and designed the game elements that it would include.
-Here we discuss  relevant aspects of selecting a practice suitable for gamification, designing the game, and competitive element considerations.
+Following the pre-study findings, we implemented a game-like tool for developers to use in Visual Studio.
+The tool focused on the objectives of motivating the adoption of good practices by including elements of games in the development environment.
+The following sections describe how we selected a practice suitable for gamification, designed the game, and included competitive elements in the game.
 
 \subsection{Select a Practice}
-The first step is to select a development practice to incorporate into the game.  
-Selecting a practice to study involves identifying a practice, defining specific characteristics of the practice to measure, and identifying the actions in Visual Studio that relate to the practice.
-In their paper on game design patterns, Hamari et.
-al.
+
+We defined the first criteria for selecting a practice based on prior work on applying gamification to other domains outside software development. 
+In their paper on game design patterns, Hamari et al.
 issue a warning that assigning achievements to required tasks reduces intrinsic motivation because player autonomy is reduced by the achievements \cite{wbsnipes:Hamari2011Framework} .
-Thus we avoided giving points for actions like fixing bugs or completing tasks.
-We also currently limit ourselves to practices that exist entirely in the IDE thus can be monitored using Blaze.
-We identified potential practices of test-driven development, debugging practices, navigation practices,  eliminating static analysis bugs, and check-in frequency.
-%EMH is check-in frequency a practice?
+Thus we avoided selecting practices like bug fixing or task completion because they are requirements that must be done.
 
-Practices of check-in frequency, static analysis bug elimination are highlighted in this paper's related work section with prior gamification studies in classroom environments.
+Another criteria is imposed by the instrument scope where we must limit ourselves to practices that exist entirely in the IDE that can be monitored using Blaze.
+We identified potential practices of test-driven development, debugging practices, navigation practices,  eliminating static analysis bugs, and frequent CM submissions.
+
+Frequent CM submission and static analysis bug elimination are highlighted in this paper's related work section with prior gamification studies in classroom environments.
 Johnson and Kou achieved automated monitoring of test-driven development practices with Zorro \cite{V:Johnson2007Automated} thus may provide a candidate for future gamification application.
@@
 
-Studies of developer navigation by Robillard et.
-al.\cite{wbsnipes:Robillard2004How} identify structured navigation as a practice that more effective developers use when maintaining programs.
+Studies of developer effectiveness by Robillard et al.\cite{wbsnipes:Robillard2004How} identify structured navigation as a practice that more effective developers use when maintaining programs.
 Structured navigation fits the requirements of a practice that is contained in the IDE, has proven benefits, lends itself to measurement and scoring, is not an assigned task, and leaves room for developers to excel.
-Thus we selected structured navigation as the practice to demonstrate gameification.
+Thus we selected structured navigation as the practice to demonstrate gamification.
 
@@
 
-To setup the game for a practice, the Blaze tool provides an XML configuration file where the researcher can configure Blaze to categorize and assign points to the commands that are part of a software engineering practice.
+To setup the game, the Blaze tool provides an XML configuration file where the researcher can configure Blaze to categorize and assign points to the commands that are part of a software engineering practice.
 Blaze allows the researcher to define multiple command category levels.
@@
 
-We categorize commands as structured navigation when they allow the developer to locate a code element following the program structure from another element.
+We categorize commands as structured navigation when they allow the developer to locate a code element by following the program structure from another element.
 These commands include Navigate To \textit{Ctrl+,}, Go To Definition \textit{F12}, and Find All References \textit{Ctrl+K,R} that are built-in to Visual Studio and follow the structured graph of the code.
 
-Commands that we considered as unstructured navigation include scrolling through a file using keys or the scroll-bar, opening files from the Solution Explorer, Find commands (e.g.
-Find in Files), Go-to line number, and children of these commands (e.g.
-Find Next).
+Commands that we considered as unstructured navigation include scrolling through a file using cursor keys or the scroll-bar, opening files from the Solution Explorer, Find commands (e.g.
+Find in Files), children of these commands (e.g.
+Find Next), and Go-to line number.
 
@@
 We gave additional emphasis to using the Sando search tool by assigning it 10 points.
-In creating levels, following the guidance from Hamari \cite{wbsnipes:Hamari2011Framework}, we built levels following an exponential curve.
+To create levels, following the guidance from Hamari \cite{wbsnipes:Hamari2011Framework}, we applied an exponential curve based on points scored.
 Users initially level up after a working day or two extending to achieving the next level after a week or even a month following an exponential curve.
-The levels make the game motivating for beginners and challenging for more experienced developers.
 We designed the points and levels such that above average developers can complete all levels during the study period.
@@
 \subsection{Competitive Elements}
+
 By selecting an intact team, we hope to leverage aspects of motivation such as the Leaderboard to spur feelings of competition among teammates.
-The leaderboard was presented so that developers could see their own score in relation to the top 5 people on the leaderboard.
+The leaderboard allows developers to see their own score in relation to the top 5 people on the team.
 However, the identifier of the other participants was generated as 3 capital letters assigned based on their position on the leaderboard.
-Therefore the assigned letters could even change when an individual's ranking changed.
-Not knowing who of your colleagues was ahead of you could be a factor that reduced motivation, but also avoids one potential source of conflict in an intact team.
-
-Competitive dynamics may be more effective when teams from different areas can compete as a team against other teams..
-This tried and true method both builds esprit de corps and enhances achievement towards the intended outcome.
-This study design is potential future work.
+Not knowing who of your colleagues was ahead of you could be a factor that reduced motivation but avoids one potential source of conflict in an intact team.
 
@@
 
-Utilizing the unobtrusive continuous monitoring capabilities of Blaze, we designed the pilot to reveal feedback to the developer in 3 stages of 1 week duration each.
-% Longtiudinal studies allow us to design an experiment and determine the effects over two or more sample periods using standard MANOVA (Multiple ANalysis of VAriance) models.
-\cite{longitudinalbook}
-We employed the three stage model defining the first stage as a baseline capturing the current practices of the pilot group.
-The second stage employed the intervention that included an informative web site on structured navigation commands in Visual Studio, information about the Sando search tool, and information about how points would be awarded during the pilot.
-The third stage provided the participants with continuous feedback in the from of points accumulated and an indicator arrow that points up when their navigation ratio is higher than their historical average.
+Utilizing the continuous monitoring capabilities of Blaze, we designed the pilot to conduct a longitudinal study of the effect of Blaze on developers' navigation practices.
+The Blaze tool revealed information about navigation in 3 stages of 1 week duration each.
+A post-pilot survey gathered developers' opinions of Blaze and some confounding factors that could influence results.
 
@@
 The About Blaze page contained details of the point scoring system and background on Blaze's purpose.
-The usefulness of the communication site is rated as part of our study completion survey.
+The usefulness of the communication site is rated as part of our post-pilot  survey.
 
@@
 Level is the level of the game achieved based on an exponentially growing curve.
-The developer could click on the Leaderbaord tab to see how their scores compared with the others in the pilot.
+The developer could click on the Leaderboard tab to see how their scores compared with the others in the pilot.
 This final configuration remained as the display until the end of the pilot.
@@
 Hence, we conducted this study with an intact team of 6 developers working in an R\&D facility in India on a large industrial software system.
-We approached a team which develops and maintains an industrial software product.
+We approached a team who develops and maintains an industrial software product.
 Initially, we met the whole team along with the management staff to explain the research project and the Blaze tool we developed.
@@
 The other questions included whether they reviewed the training materials on structured navigation commands, installed the Sando code search tool, and what types of development tasks they would consider using structured navigation for in general.
-Another demographic question is how well they know the code they work in rated from "I work in code that I wrote", "I maintain code that I am familiar with", to "I recently learned the code"
+Another demographic question is how well they know the code they work in rated from ``I work in code that I wrote'', ``I maintain code that I am familiar with'', to ``I recently learned the code''
 
@@
 The chart in figure \ref{fig:elementInfluence} shows how the individual feedback features influenced developers to think about using structured navigation.
-Four out of six developers said the leaderboard influenced them "a lot" and all rated the leaderboard as having some influence.
-Three out of six  said their individual point score influenced them "a lot" and all said the score had some influence.
+Four out of six developers said the leaderboard influenced them ``a lot'' and all rated the leaderboard as having some influence.
+Three out of six  said their individual point score influenced them ``a lot'' and all said the score had some influence.
 The element with mixed ratings was a graphical indicator on whether they were improving over past use of structured navigation commands.
-The indicator received the influencing "a lot" rating from one participant and a "not at all" influence rating from one participant with the rest having some influence.
+The indicator received the influencing ``a lot'' rating from one participant and a ``not at all'' influence rating from one participant with the rest having some influence.
 The results indicate the more obvious and more clearly comparative information in the leaderboard was the strongest influence for developers.
@@
 Analysis of the log extracts several key information bits from this raw source.
-First we classify the event names at a higher level into categories related activities such as Navigating, Reading, Editing, Building, Debugging,  Testing, and using a  known Tool.
+First we classify the events at a higher level into categories related activities such as Navigating, Editing, Building, Debugging,  Testing, and using a  known Tool.
+Each event name is mapped to a category and we focus on a sub-division of the Navigation category that separates Structured Navigation from Unstructured Navigation.
 
@@
 When the developer edits a new file, all the events after the last edit to the previous file are part of the new sequence for the new file.
-Although not at the method level, this is similar to the say Robillard et.
-al.
-generated sequence based on located methods for their study.\cite{wbsnipes:Robillard2004How}
+For example, the developer edits file A, then navigates through files B and C finally landing at line 100 of file D where they make another edit.
+All the events between the edit of file A and the edit of file D are considered part of the edit session ending with several edits of file D.
+Multiple edits to file D that occur before editing a different file are included in the same sequence that ends with the edits to file D.
+Although not at the method level, this method of generating sequences is similar to how Robillard et al.
+generated sequences based on located methods for their study \cite{wbsnipes:Robillard2004How}.
 
@@
 
-We selected navigation as the subject of the pilot game because it was identified by Robillard et.
-al\cite{wbsnipes:Robillard2004How} as correlated with developers who effectively completed a maintenance task.
+We selected navigation as the subject of the pilot game because it was identified by Robillard et al.\cite{wbsnipes:Robillard2004How} as correlated with developers who effectively completed a maintenance task.
 In this analysis we ask whether we are focusing on the right area of developers' activity in Visual Studio?
@@
 The process identified the top 5 out of 15 possible categories as follows:
-\begin{enumerate}
+\begin{enumerate}[itemsep=0mm]
 \item Unstructured Navigation
@@
 Structured navigation events included: 
-\begin{itemize}
+\begin{itemize}[itemsep=0mm]
 \item Navigate To (Ctrl+,) is a fuzzy search interface that locates symbols matching your search terms.
@@
 
+We discovered during these feedback sessions that developers did not spend the entire time in their desktop environment because they had a virtual machine environment available for testing.
+Thus categories like testing and debugging were affected by whether or not they used the virtual environment or their desktop for these activities.
+Individual differences in those categories are considered irrelevant and not discussed here.
+The presence of the virtual environment may explain how some developers have no activity recorded on some days.
+
 \begin{figure}
@@
 \end{figure}
-
-We discovered during these feedback sessions that developers did not spend the entire time in their desktop environment because they had a virtual machine environment available for testing.
-Thus categories like testing and debugging were affected by whether or not they used the virtual environment or their desktop for these activities.
-Individual differences in those categories are considered irrelevant and not discussed here.
-The presence of the virtual environment may explain how some developers have no activity recorded on some days.
 

### 2013-10-20 paper_id1.tex
paper submitted to everyone for review on Friday AM

@@
 As software development practices evolve, toolsmiths face the continuous challenge of communicating new practices and tools to the community.
-We tested an idea with industrial software developers that adding game-like feedback to the development environment focused on improving practices and tool use would improve adoption of tools and practices for code navigation.
-We conducted this study in three stages: a pre-study to validate our ideas, a pilot of a tool called Blaze that implements some game features to motivate better navigation practices, and a post-pilot survey conducted via interviews.
-We found that a software engineering game with competitive elements influences developers towards changing their practices.
+We tested an idea with industrial software developers that adding game-like feedback to the development environment would improve adoption of tools and practices for code navigation.
+We conducted this study in three stages: a pre-study to validate our ideas, pilot of a tool with some game features to motivate better navigation practices, and feedback interviews for a post-pilot survey.
+We found that a software engineering game with competitive elements influences some developers towards changing their practices.
 %We also learned that demonstrating significant improvement can be difficult and that developers want more proactive messages and recommendations on ways they can improve.
@@
 Sascha is a developer with years of experience maintaining code who knows how to navigate through code fairly well.
-Sascha heard about some other navigation tools, but kept using the built-in features of the Visual Studio IDE that served him well.
-When Sascha's company setup a game around code navigation in Visual Studio, he became interested in being a part of the fun.
+Sascha heard about some other navigation tools, but kept using the built-in features of the Visual Studio IDE.
+When Sascha's company started a game around code navigation in Visual Studio, he became interested in being a part of the fun.
 The game gave bonus points for using Sando, a new code search tool, so he started using it instead of some of the built-in search commands.
-Sascha soon topped the game's leaderboard and discovered that motivated him to try the next tool to became part of the game.
+Sascha soon topped the game's leaderboard and discovered he became motivated to try the next practices included in the game.
 
 Like Sascha, software developers who maintain code face additional challenges of navigating an existing code base to incorporate changes as they build new features or fix bugs.
-Over the years, developers learn techniques to analyze code and locate areas that are relevant to an enhancement.
+Over the years, developers learn techniques to analyze code and locate relevant areas for an enhancement.
 As new practices and tools get introduced, developers may not learn about them unless they interact with a peer developer who uses them\cite{wbsnipes:Hill2011Peer}.
@@
 The toolsmith needs to spread the word about advanced tools they develop and determine their effect on developer practices over the long term.
-Is there something we can do that encourages developers to adopt new practices and tools that also provides feedback to developers and toolsmiths on the resulting improvements?
+Is there something we can do to encourage developers to adopt new practices and tools that also provides feedback to developers and toolsmiths on the resulting improvements?
 
@@
 Gamification is the incorporation of game elements into an activity that people do not typically consider a game\cite{2013Oxford}.
-We envisioned a tool in our prior work that would motivate developers to adopt new practices or tools by providing constant feedback and allowing them to compare their improvements to how their peers are doing  \cite{Snipes2013Towards}.
-This tool called Blaze, creates a competitive game that gives continuous feedback to the developer on adopting the improved practice.
+We envisioned a tool in our prior work to motivate developers by providing constant feedback and allowing them to compare their improvements to how their peers are doing  \cite{Snipes2013Towards}.
+This tool called Blaze, creates a competitive game with continuous feedback on how well the developer adopts the improved practice.
 The more they change their practices to adopt new techniques or tools, the higher they score in the game.
@@
 To answer this question, this work contributes an assessment of developers' response to game-like elements tied to their use of new tools and practices.
-We assess their response to the idea of gamification and relevant concerns through a survey of 130 software developers.
-Then we conduct a pilot of a tool, Blaze, that introduces game elements into the developer's IDE focused on improving navigation practices.
+We assess 130 developers' response and relevant concerns about the idea of gamification through a web-based survey.
+We piloted Blaze, our tool that introduces game elements focused on improving navigation practices into the developer's IDE.
 With the pilot feedback we determine the acceptance, influence, and effect of deploying the Blaze tool in the pilot group.
 
-The rest of this paper is organized as follows:  Section 2 reviews Related work, Section 3 reviews results from the pre-study survey, Section 4 Discusses the design considerations for developing the Blaze tool/game, Section 5 Discusses the design of the Blaze pilot study and participant selection,  Section 6 results from the pilot study and post-pilot survey, Section 7 Threats to Validity, and Section 8 has our conclusions.
+The rest of this paper is organized as follows:  Section 2 reviews related work, Section 3 reviews results from the pre-study survey, Section 4 discusses the design considerations for developing the Blaze tool/game, Section 5 discusses the design of the Blaze pilot study,  Section 6 results from the pilot study and post-pilot survey, Section 7 covers threats to talidity, and Section 8 has our conclusions.
 %This is precisely what we set out to study for this work. 
@@
 \section{Related Work}
-%update and expand
+
+The two areas of related work we discuss are research studies using other monitoring methods and studies applying motivation and gamification to other areas.
+
 \subsection{Monitoring Practice Studies }
+
+Robillard, Coelho and Murphy explore hypotheses around how developers can be more effective at performing a maintenance task \cite{wbsnipes:Robillard2004How}.
+Robillard et al.
+say developers are more successful finding and fixing bugs when they create a detailed plan for implementing a change, use structured navigation with keyword search or cross-reference search, and only review methods once during their search.
+We build on their work by testing methods for increasing the use of structured navigation in developers' everyday practice.
+
 Johnson and Kou defined Zorro\cite{V:Johnson2007Automated}, a system for detecting whether developers use Test Driven Development techniques based on data from Hackystat.
 Zorro divides development activities into episodes delimited by events such as configuration management code check-in, start of a unit test run, or start of a build.
-Using the distinct events that developers follow within these episodes, Zorro determines whether the episode followed Test-Driven Development practices per their classic definition of test first - code - test pass, or a different scenario.
-In two student-based studies comparing Zorro classifications with a simultaneous observational screen video, Zorro achieved between 70\% \cite{Kou2010Operational} and 89\% \cite{V:Johnson2007Automated} accuracy in classifying episodes into their proper TDD scenarios.
+Using the distinct events developers follow within these episodes, Zorro determines whether the episode followed Test-Driven Development practices per their classic definition of test first - code - test pass, or a different scenario.
+In two student-based studies comparing Zorro classifications with a simultaneous observational screen video, Zorro achieved between 70\% \cite{Kou2010Operational} and 89\% \cite{V:Johnson2007Automated} accuracy when classifying episodes into their proper TDD scenarios.
 The studies did not, however, attempt to use or evaluate the influence of instant feedback aimed at encouraging participants to follow the classic definition of Test Driven Development.
-In our study, we attempt this leap to present data collected directly to developers attempting to influence them to change their practices.
+In our study, we attempt to present data collected directly to developers in order to influence them to change their practices.
 
@@
 They look for specific refactoring commands in Eclipse and determine the amount of time developers use tools versus hand refactoring the code.
-For this study we build a tool that captures the use of all commands within the Visual Studio environment including refactoring, navigation, and edit commands.
-
-Robillard, Coelho and Murphy explore hypotheses around how developers can be more effective at performing a maintenance task \cite{wbsnipes:Robillard2004How}.
-Key conclusions are that developers are more successful finding and fixing bugs when they create a detailed plan for implementing a change, use structured navigation with keyword search or cross-reference search, and only review methods once during their search.
-This builds on this work by testing methods for increasing the use of structured navigation in developers' everyday practice.
+For this study we focus on code navigation as a high-value area for improving developer efficiency.
 
 The Pro Metrics (PROM) tool provides a framework for collecting data for further analysis from tools used by developers.\cite{Coman2009Casestudy}  It provides a flexible data model and a plug-in architecture to facilitate collection from different data sources.
-Studies conducted using PROM include a series of studies by it's creators on trends in time spent Pair Programming \cite{Coman2008Investigating}, benefits of refactoring on productivity \cite{Moser2008Case}, impact of refactoring on re-usability \cite{Moser2006Does}, and prediction of effort \cite{Abrahamsson2007Effort}.
-The PROM tool and studies that use it track the time spent editing files and methods from the IDE and code metrics obtained trough source code analysis.
+Studies conducted using PROM include a series of studies on trends in time spent Pair Programming \cite{Coman2008Investigating}, benefits of refactoring on productivity \cite{Moser2008Case}, impact of refactoring on re-usability \cite{Moser2006Does}, and prediction of effort \cite{Abrahamsson2007Effort}.
+Studies using the PROM tool correlate the time spent editing with code metrics obtained trough source code analysis.
 Applying these data to specific research questions such as whether refactoring improves productivity \cite{Moser2008Case} shows the utility of combining automated effort data with code metric data.
@@
 Murphy-Hill et al.
-study a large usage history data set and apply several different algorithms to accurately suggest to users new commands in their environment to use.\cite{MurphyHill2012Improving} Algorithms based on command history performed well particularly when synchronized chronologically with the recipient's usage history.
+study a large usage history data set and apply several different algorithms to accurately suggest IDE commands to novice users.\cite{MurphyHill2012Improving} Algorithms based on command history performed well particularly when synchronized chronologically with the recipient's usage history.
 Novice users were better served by recommendation algorithms that did not require a long history, where more expert users benefited from more sophisticated algorithms that included more history.
-The ``Most Widely Used'' algorithm that recommends commands based on the collective usage profile of the team performed nearly as well as the more sophisticated algorithms.
-The paper presents the methodology for making recommendations and the evaluation of the accuracy of the recommendations.
-Our work goes the next step of evaluating whether a specific recommended practice gets adopted by developers when influenced through gamification methods.
+The ``Most Widely Used'' algorithm recommends commands based on the collective usage profile of the team and performed nearly as well as the more sophisticated algorithms.
+The paper presents the methodology for making recommendations and evaluate their accuracy.
+Our work goes takes a further step evaluating how well tools can influence developers to use a specific recommended practice.
 
@@
 
-When discussing gamification, folks raise a concern about replacing an activity that is intrinsically motivated with extrinsic motivation provided by points and achievements.
+When discussing gamification, folks raise a concern about replacing an activity like software development, which is intrinsically motivated, with extrinsic motivation provided by points and achievements.
 To discover what is behind this concern, several studies are useful to understand what should be done.
 
-Early studies attempt to evaluate achievement motivation in students and relate that to their performance in school.
-The study by Hermans describes results from applying questionnaire to evaluate achievement motivation in students \cite{wbsnipes:Hermans1970Questionnaire}.
-Each question is ranked for its correlation to achievement motivation.
-Some key questions about the diligence with which students approach their work are correlated with achievement motivation.
+Beecham and colleagues provide a thorough analysis of existing literature on studies of motivation in Software Engineering \cite{Beecham2008Motivation}.
+They claim most common motivator found in cited references is ''the work'' motivates us.
+The list of de-motivators is also useful with common job satisfaction items like stress and inequity in recognition, plus poor quality software (low accomplishment) and lack of influence in decision making.
+The paper also lists characteristics of people in the professions including need for independence (autonomy), desire to learn new skills and tackle new challenges.
 
-Maehr proposes an affirmative theory whereby  individuals achieve as a member of a social group choosing the behavior that will meet the expectations and values of the group that is significant to them \cite{wbsnipes:MaehrCulture}.
-The presence of a social group where the person is motivated to belong and excel within is important for creating achievement motivation.
-Maehr states that since achievement behaviors can be triggered by circumstance, that they are present in most people regardless of background.
+Maehr proposes an affirmative theory whereby  individuals achieve as a member of a social group choosing the behavior that meets the expectations and values of the group that is significant to them\cite{wbsnipes:MaehrCulture}.
+A key factor in raising achievement motivation is establishing a social group where the person is motivated to belong and excel.
+Maehr states that since achievement behaviors can be triggered by circumstance, they are present in most people regardless of background.
 
-Beecham and colleagues provide a thorough analysis of existing literature on studies of motivation in Software Engineering \cite{wbsnipes:Beecham2008Motivation}.
-They provide a useful inventory of aspects of the job motivate software engineers with the most common being ''the work'' motivates us.
-The list of de-motivators is also useful with common job satisfaction items like stress and inequity in recognition, plus poor quality software (low accomplishment) and lack of influence in decision making.
-The paper also lists characteristics of people in the professions including need for independence (autonomy), desire to learn new skills/tackle challenges.
+Hamari et al.
+analyze achievements patterns in game systems and provide guidance on defining successful achievement systems.
+They recommend  achievement awards be setup as a secondary form of scoring so players have multiple objectives available for a particular game.
+Players should feel like they earned achievements in the game, they should not be given for trivial tasks.
+
 Singer and Schneider demonstrate the use of a message board and points for encouraging students to increase their frequency of commits to the source code repository \cite{Singer2012It}.
@@
 Their approach is to consider how gamification could be applied throughout the software development life-cycle to motivate many practices in different phases.
-They provide guidance on assessing gamification in the software development domain that includes analyzing the actors and game mechanics, defining integration with existing tools and processes, and designing the evaluation of results.
+They provide guidance on assessing gamification in the software development domain including analyzing the actors and game mechanics, defining integration with existing tools and processes, and designing the evaluation of results.
 Dbois and Tamburrelli demonstrate this guidance using a study of students working on a class project.
@@
 The subject group can see each others' rankings while the control group only sees their personal rankings.
-Results though inconclusive indicate that competition increases the desired results.
-In our study we apply both individual feedback and competitive based feedback (anonymized) to participants and seek to observe a change in their behavior after an intervention.
+Results though inconclusive, show competition increases the desired results.
+In our study we provide participants with both individual feedback and anonymized competitive feedback and seek to observe a change in their behavior.
 
@@
 Understanding the concerns with maintaining developers'  intrinsic motivation while adding elements of achievement motivation, we conducted an assessment survey questionnaire to determine whether the ABB developer community would be receptive to a gamification approach to software development.
-Another question asked about ways of motivating developers to use training, practices, and tools.
-We checked on who they would be comfortable sharing data from their development activities with.
-Finally we asked what kind of awards would be most welcome such as badges and points.
 The developer survey received over 130 responses from software developers at ABB.
@@
 Using a question from the Saatchi and Saatchi survey \cite{wbsnipes:SaatchiGameification} as directly as possible, we asked developers at ABB ``How interested they would be in working for a company that incorporated some aspects of games into software engineering tools as a way to increase productivity in the workplace''?
-We segmented results by country in Figure \ref{fig:gamification} and found that gamification of software engineering is more interesting to developers in India, Poland, and Finland than in other major countries.
-The responses indicating that developers were at least somewhat interested is 75-88\% In India, Poland, and Finland.
+We segmented results by country in Figure \ref{fig:gamification} and found gamification of software engineering is more interesting to developers in India, Poland, and Finland than in other major countries.
+The responses indicate developers were at least somewhat interested is 75-88\% In India, Poland, and Finland.
 For the US, Sweden, and Switzerland the responses indicate 60-65\% are at least somewhat interested.
 Because we nearly identically reused their original question, we compare the overall response from our survey to the Saatchi and Saatchi survey\cite{wbsnipes:SaatchiGameification} upon which this question was based.
-Our results land on both sides of the 75\% of 18-45 year olds that Saatchi and Saatchi found were at least somewhat interested.
+Our results land on both sides of the 75\% mark resulting from the Saatchi and Saatchi study.
 
@@
 We asked developers how likely they would be to try tools and practices recommended to them through an automated usage tracking system.
-Answers showed that 95\% of the developers surveyed are likely to try the recommended tools and practices.
-This indicates that a recommendation system would positively impact the deployment of software engineering tools and practices at ABB.
+Answers showed 95\% of the developers surveyed are likely to try the recommended tools and practices.
+Thus, a recommendation system would positively impact the deployment of software engineering tools and practices at ABB.
 
 Another question was whether sharing detailed usage information with colleagues and the company as a whole would be a concern for developers.
-The question provided a graduated scale for the scope of who data were shared with and divided the questions between whether that sharing was anonymous or not.
-Figure \ref{fig:comfortwithsharing}  shows that over 90\% of respondents are comfortable with sharing either anonymous or non-anonymous information with their team members.
+The question provided a graduated scale for the scope of who data were shared with and divided the questions between whether sharing was anonymous or not.
+Figure \ref{fig:comfortwithsharing}  shows over 90\% of respondents are comfortable with sharing either anonymous or non-anonymous information with their team members.
 Their comfort level decreases as the scope of who the information is shared with increases.
 People were less comfortable sharing with anyone (the category for people outside ABB) particularly if the data were not kept anonymous.
-The first conclusion from these questions on sharing data is that people are very willing to share information that could help their team.
+The first conclusion from these questions on sharing data is people are very willing to share information that could help their team.
 Second, sharing within the company is acceptable if we take care to make the data anonymous.
@@
 We conclude that a facility to share coworker recommendations on tools and practices has the best opportunity to increase their adoption.
-Since a competition among teams ranks slightly higher than management mandate, we have the choice to apply one or the other or both as motivation for using key practices.
 
@@
 Overall 68\% of developers ranked receiving an award for consistently performing quality practices as either 1st or 2nd place.
-This supports a conclusion that focusing on awarding good practices would motivate developers to do them consistently.
+Thus, focusing awards on good practices would motivate developers to do them consistently.
 
@@
 \end{itemize}
-Thus, this pre-study survey provided a strong positive indicator for going forward with a software engineering tool that focused on sharing developer data and rewarding developers for using good practices and tools.
+Thus, this pre-study survey provided a strong positive indicator for going forward with a software engineering tool focused on sharing developer data and rewarding developers for using good practices and tools.
 
@@
 Following the pre-study findings, we implemented a game-like tool for developers to use in Visual Studio.
-The tool focused on the objectives of motivating the adoption of good practices by including elements of games in the development environment.
+The tool focused on motivating the adoption of good practices by including elements of games in the development environment.
 The following sections describe how we selected a practice suitable for gamification, designed the game, and included competitive elements in the game.
@@
 
-We defined the first criteria for selecting a practice based on prior work on applying gamification to other domains outside software development. 
+We defined the first criteria for selecting a practice based on prior work  applying gamification to other domains outside software development. 
 In their paper on game design patterns, Hamari et al.
-issue a warning that assigning achievements to required tasks reduces intrinsic motivation because player autonomy is reduced by the achievements \cite{wbsnipes:Hamari2011Framework} .
-Thus we avoided selecting practices like bug fixing or task completion because they are requirements that must be done.
+issue a warning that assigning achievements to required tasks reduces intrinsic motivation because player autonomy is reduced by the achievements \cite{wbsnipes:Hamari2011Framework}.
+Thus we avoided selecting practices like bug fixing or task completion because they are simply requirements of the job.
 
-Another criteria is imposed by the instrument scope where we must limit ourselves to practices that exist entirely in the IDE that can be monitored using Blaze.
+Another criteria is imposed by the instrument scope where we must limit ourselves to practices performed entirely in the IDE so they can be monitored using Blaze.
 We identified potential practices of test-driven development, debugging practices, navigation practices,  eliminating static analysis bugs, and frequent CM submissions.
@@
 
-Studies of developer effectiveness by Robillard et al.\cite{wbsnipes:Robillard2004How} identify structured navigation as a practice that more effective developers use when maintaining programs.
-Structured navigation fits the requirements of a practice that is contained in the IDE, has proven benefits, lends itself to measurement and scoring, is not an assigned task, and leaves room for developers to excel.
+Studies of developer effectiveness by Robillard et al.\cite{wbsnipes:Robillard2004How} identify structured navigation as a practice more effective developers use when maintaining programs.
+Structured navigation fits the requirements of a practice contained in the IDE, has proven benefits, lends itself to measurement and scoring, is not an assigned task, and leaves room for developers to excel.
 Thus we selected structured navigation as the practice to demonstrate gamification.
@@
 Blaze allows the researcher to define multiple command category levels.
-Thus we can categorize commands that are navigation then further classify them into structured and unstructured navigation.
+Thus we can categorize commands as navigation then further classify them into structured and unstructured navigation.
 
 We categorize commands as structured navigation when they allow the developer to locate a code element by following the program structure from another element.
-These commands include Navigate To \textit{Ctrl+,}, Go To Definition \textit{F12}, and Find All References \textit{Ctrl+K,R} that are built-in to Visual Studio and follow the structured graph of the code.
+These include  built-in Visual Studio commands like Navigate To \textit{Ctrl+,}, Go To Definition \textit{F12}, and Find All References \textit{Ctrl+K,R}  that follow the structured graph of the code.
 
-Commands that we considered as unstructured navigation include scrolling through a file using cursor keys or the scroll-bar, opening files from the Solution Explorer, Find commands (e.g.
+Commands we considered as unstructured navigation include scrolling through a file using cursor keys or the scroll-bar, opening files from the Solution Explorer, Find commands (e.g.
 Find in Files), children of these commands (e.g.
@@
 The design of achievement awards should gradually encourage the participant to reach higher and higher levels of success in the activity.
-Achievements must feel earned to the participant so that internally they recognize the effort to get the award was significant and feel satisfied  \cite{wbsnipes:Hamari2011Framework}.
+Achievements must feel earned to the participant so they recognize the effort to get the award was significant and feel satisfied  \cite{wbsnipes:Hamari2011Framework}.
 Hamari and Eranti describe a general relationship between achievements in games and regular game play      In games achievements are typically a parallel scoring system to the main game play.
@@
 To create levels, following the guidance from Hamari \cite{wbsnipes:Hamari2011Framework}, we applied an exponential curve based on points scored.
-Users initially level up after a working day or two extending to achieving the next level after a week or even a month following an exponential curve.
-We designed the points and levels such that above average developers can complete all levels during the study period.
+Users initially ``level up'' after  working a day or two; the next level may be achieved after a week or even a month following an exponential curve.
+We designed the points and levels so above average developers can complete all levels during the study period.
 
@@
 However, the identifier of the other participants was generated as 3 capital letters assigned based on their position on the leaderboard.
-Not knowing who of your colleagues was ahead of you could be a factor that reduced motivation but avoids one potential source of conflict in an intact team.
+Not knowing who of your colleagues was ahead of you could be  reduce motivation, but this avoids one potential source of conflict in an intact team.
 
@@
 
-Part of the research objective was to determine whether we can achieve adoption of a practice with alternative interventions that are more deliberate than word of mouth and perhaps more scalable than webinars or presentations.
-As previously discussed we sought to provide Blaze as a means to influence developers to use structured navigation practices.
-We constructed the interventions during the pilot to roll out in stages each week that they use the tool.
+Part of the research objective was to determine whether we can achieve adoption of a practice with alternative interventions.
+As previously discussed, we sought to provide Blaze to influence developers towards useing more structured navigation practices.
+We constructed the interventions during the pilot to roll out in stages each week they use the tool.
 This allowed us to establish a baseline and determine the changes as the pilot continued.
@@
 After a week's data collection marking the end of the first stage, Blaze started to automatically pop-up a window containing a button for a web page with information.
-The page included a link to download Sando  and information about using that tool in a video.
+The page included a link to download Sando  and information about using Sando in a video demo.
 The sub-page on navigation contained a page of tutorial on the built-in structured navigation commands in Visual Studio.
@@
 It was important to gain the confidence of the participants that their data were kept confidential so their activities would be close as possible to real-world.
-if the participants felt like they are being tracked and the data is going to use for some study that would identify them to their peers or managers, then they may deviate from the natural development style.
+if the participants felt like they are being tracked and the data would identify them to their peers or managers, then they may refuse to participate or deviate from their normal development style.
 
-We  assured the team about the confidentiality of the data and the management also informed the team that the data collected by the research group will not be used for any other purpose other than the study.
+We  assured the team about the confidentiality of the data and the management reassured the data collected by the research group will not be used for any other purpose other than the study.
 Developers were asked to volunteer for the pilot by their management, however, managers did not know who participated in the study.
-Their name and data were kept confidential to ensure that no one apart from the research team knows that he/she is participating in the pilot.
+Their name and data were kept confidential to ensure no one apart from the research team knows he/she is participating in the pilot.
 Mechanisms in our data collection to identify individuals were under the control of each developer, they could chose to share or keep confidential the association of a generated unique id with researchers or each other.
@@
 Management wished to have no knowledge of who was participating and took great care to remain detached from the pilot proceedings.
-Data at this level of detail has not been collected in ABB development organizations except for a few minutes of video study that might be conducted.
-Researchers were careful to create the system so that developers controlled their identification with the data and fortunately, management bought into the necessary anonymity of the participants.
+Data at this level of detail has not been collected in ABB development organizations.
+Researchers were careful to create the system where developers controlled their identification with the data and fortunately, management bought into the necessary anonymity of the participants.
 
@@
 Then we asked whether the used structured navigation commands about the same amount, more,  or a lot more during the pilot.
-These two questions assessed their perceptions of their structured navigation command use that we compared with the data collected from Blaze.
+These two questions assessed their perceptions of structured navigation command use.
+We compared the answers with the usage data collected from Blaze.
 
@@
 The pilot survey queried developers' relevant demographic data to help understand some differences in results.
-We asked developers to provide their years of experience so that we could evaluate any effects experience might have.
+We asked developers to provide their years of experience so we could evaluate any effects experience might have.
 The developers in the study had between 7 and 14 years experience.
-One developer who dropped out of the study said that their lower level of experience compared to their peers influenced their decision to stop participating.
+One developer who dropped out of the study said their lower level of experience compared to their peers influenced their decision to stop participating.
 
 We asked developers to rate their knowledge of the code they maintain.
-Four of the six developers reported that they maintain code that they wrote while the two most experienced developers maintain code that they recently learned.
+Four of the six developers reported they maintain code that they wrote while the two most experienced developers maintain code that they recently learned.
 One developer reported during our individual analysis feedback that working in code they know well affects their navigation practices leading them to use more unstructured navigation.
 
-On the assessment of structured navigation knowledge, half of the participants replied that they had very good prior knowledge about structured navigation and they use structured navigation as much as possible.
+On the assessment of structured navigation knowledge, half of the participants replied they had very good prior knowledge about structured navigation and they use structured navigation as much as possible.
 Other half of the participants have some understanding on structured navigation use it some of the time.
 Even developers whose navigation ratio  was lower than average, used structured navigation occasionally as much as their colleagues.
-While analyzing the points they scored with Blaze tool, we observed that the participants who rated their prior knowledge of structured navigation highly also scored higher in the pilot.
+While analyzing the points they scored with Blaze tool, we observed the participants who rated their prior knowledge of structured navigation highly also scored higher in the pilot.
 
@@
 
-The participants requested that Blaze provide more feedback on tips to help them increase their score.
-They indicated that more obvious feedback such as hints that pop-up when they launch the tool would help them more than passively provided information.
+The participants requested Blaze provide more feedback on tips to help them increase their score.
+They asked for more obvious feedback such as hints that pop-up when they launch the tool would help them more than passively provided information.
 Their desire for personal metrics and historical views of the data encourages us that this detailed level of measurement and analysis is perceived as helpful.
@@
 Blaze logs events from actions the developer takes or actions from Visual Studio itself.
- The event is basically a GUI event that the program manages with event handlers that are registered with the application to listen for the event.
+ The event is basically a GUI event managed with Visual Studio by event handlers registered with the application to listen for the event.
 Blaze becomes a global event handler listening for all events by registering for them in Visual Studio.
@@
 Analysis of the log extracts several key information bits from this raw source.
-First we classify the events at a higher level into categories related activities such as Navigating, Editing, Building, Debugging,  Testing, and using a  known Tool.
-Each event name is mapped to a category and we focus on a sub-division of the Navigation category that separates Structured Navigation from Unstructured Navigation.
+First Blaze classifies the events at a higher level into categories related activities such as Navigating, Editing, Building, Debugging,  Testing, and using a  known Tool.
+Each event name maps to a category from the list and sub-division of the Navigation category separates Structured Navigation from Unstructured Navigation.
 
-The second information we extract is aggregating the events along the time-line into sessions.
-A session is an abstract concept that groups multiple events into a sequence with a beginning and an end.
+In another processing step, Blaze aggregates events along the time-line into sessions.
+A session is an abstract concept grouping multiple events into a sequence with a beginning and an end.
 Of the many possible ways to define sessions, we chose to define sequences as all the events that transpire before and during a set of edits to a specific file.
@@
 All the events between the edit of file A and the edit of file D are considered part of the edit session ending with several edits of file D.
-Multiple edits to file D that occur before editing a different file are included in the same sequence that ends with the edits to file D.
-Although not at the method level, this method of generating sequences is similar to how Robillard et al.
+Multiple edits to file D that occur before editing a different file are included in the same sequence ending with the edits to file D.
+Although not at the method level, this way of generating sequences is similar to how Robillard et al.
 generated sequences based on located methods for their study \cite{wbsnipes:Robillard2004How}.
@@
 
-In order to test whether navigation is an important area to address, we construct a hypothesis that use of unstructured navigation is an important factor in developer productivity.
+In order to test whether navigation is an important area to address, we construct a hypothesis that using unstructured navigation is an important factor in developer productivity.
 Edit session duration, we assert by opinion, is a factor related to developer productivity though there are many other factors we could consider.
-With the data available from Blaze, we can explore the categories of events that correlate with the duration of edit sessions over the period of the pilot.
+With the data available from Blaze, we can explore the categories of events correlated with the duration of edit sessions over the period of the pilot.
 
 Using Weka\cite{Hall2009WEKA}, we performed an attribute selection process to identify the most significant attributes related to edit session duration.
-The attribute search method we chose was Greedy Step-wise search that selects attributes based on their correlation with the target and stops adding attributes when the next attribute decreases the correlation.
+The attribute search method we chose was Greedy Step-wise search selecting attributes based on their correlation with the target and stops adding attributes when the next attribute decreases the correlation.
 The process identified the top 5 out of 15 possible categories as follows:
@@
 The linear model found a positive correlation with unstructured navigation with a p-value <0.001 on 1694 degrees of freedom.
-The model had an Adjusted R-squared of 0.64 showing a good portion of the variation in edit duration is explained by that category.
+The model had an Adjusted R-squared of 0.64 showing a good portion of the variation in edit duration is explained by the Unstructured Navigation category.
 
-This analysis shows that we cannot disprove that unstructured navigation is correlated with edit duration, thus we conclude that it is an important factor in developer productivity.
+This analysis shows we cannot disprove that unstructured navigation is correlated with edit duration, thus we conclude it is an important factor in developer productivity.
 
@@
 During the pilot we staged the deployment of information feedback to developers so they would first have access to learn about structured navigation and then get instant feedback in the form of points for using structured navigation commands.
-Our hypothesis was that this instant feedback and information would result in an increase in the number of structured navigation commands used in Visual Studio.
+We hypothesized instant feedback and information would result in an increase in the number of structured navigation commands used in Visual Studio.
 
 To evaluate whether developers use more structured navigation when they receive instant feedback, we monitor the change in navigation ratio over the pilot study period.
-Expecting that as developers learn about navigation commands and tools that count towards the score in the second week they will begin using the practices more.
+We expected developers learning about navigation commands and tools that count towards the score would begin using the practices more during the second week.
 In the third week we turn on points feedback and the leaderboard expecting this will encourage developers to use more structured navigation.
@@
 
-What we observed visually in Figure \ref{fig:navigationaverage} is that the structured navigation practice did not increase significantly from the beginning to the end of the study.
-We evaluated the week over week differences using a Wilcoxon Rank sum test \cite{RefWorks:118} that compares week one to week two and week one to week three.
+We observe visually in Figure \ref{fig:navigationaverage} the structured navigation practice did not increase significantly from the beginning to the end of the study.
+We evaluated the week over week differences using a Wilcoxon Rank sum test \cite{RefWorks:118} comparing week one to week two and week one to week three.
 In week two when the developers were able to learn about structured navigation commands there was no significant change.
@@
 Week three there was a slight increase in the mean, however the increase is not significant.
-Weeks four and five that follow the formal period of study show the navigation ratio continues in a similar range.
+Weeks four and five follow the formal period of study show the navigation ratio continues in a similar range.
 Overall, we did not see a significant effect towards increasing navigation ratio for the interventions Blaze provided to the environment.
@@
 The developers using Blaze received feedback in the form of points given for using structured navigation commands and using the Sando search tool\cite{Shepherd2012Sando}.
-Per the pilot design, the points display was disabled until the third week following installation so that a baseline without instant feedback could be established.
+Per the pilot design, the points display was disabled until the third week following installation to establish the efffect of instant feedback apart from other information.
 The points display and the navigation arrow were enabled in the third week to test whether the instant feedback was successful in driving increased use of the targeted features.
-The result shows that during the first two weeks shown in Figure \ref{fig:pointsbyweek}, developers were fairly consistent in their use patterns.
+During the first two weeks shown in Figure \ref{fig:pointsbyweek}, developers were fairly consistent in their use patterns.
 With the introduction of the points feedback, two developers' points accumulation spiked in week three due mostly to increased use of Sando.
 We informed developers at the beginning of the pilot that Sando usage would receive 10 points and using built-in structured navigation commands would receive 1 point.
-In these results we see an initial though not significant effect from the points display in Blaze that is not sustained over the remaining weeks.
-We found following discussions that developers wanted more guidance and more prompting included in the game which we discuss further in the next section.
+
+In these results we see an initial though not statistically significant effect from the points display in Blaze in Week three.
+In week three, the developers could see their points feedback for the first time.
+Two developers took more notice of this than others and used Sando more in the third week.
+Their use of Sando continued in weeks 4 and 5 though with lower frequency.
+The effect of points looks greater than navigation ratio during Week 3 because of the additional emphasis of the 10 point bonus given for Sando usage.
+Developers in thir post-pilot feedback felt more guidance and more prompting included in the game would help guide them towards higher scores.
 
@@
 Following our data analysis, we conducted one on one feedback sessions with the pilot participants.
-Before initiating the discussion, we assured the individual developer that the data shared with them is confidential and we will not share the same with management.
+Before initiating the discussion, we assured the individual developer the data shared with them is confidential and we will not share the same with management.
 The participants received the feedback in the form of quantitative charts and written analysis of the information.
@@
 Thus categories like testing and debugging were affected by whether or not they used the virtual environment or their desktop for these activities.
-Individual differences in those categories are considered irrelevant and not discussed here.
+Because those categories are not part of navigation, they are not discussed here.
 The presence of the virtual environment may explain how some developers have no activity recorded on some days.
@@
 We asked the developer to explain how building more frequently helps them.
-The developer reported that by doing more frequent build, (s)he is trying to ensure the correctness of the work completed so far.
-The developer feels that (s)he resolves the bugs faster with frequent builds because the code and logic is fresh in her memory.
+By doing more frequent build, (s)he is trying to ensure the correctness of the work completed so far.
+The developer feels like (s)he resolves the bugs faster with frequent builds because the code and logic is fresh in her memory.
 If they build less frequently, even though it reduces the build effort, there are more bugs each time and the developer often spends more time in identifying the problem area than fixing it.
-Hence the developer feels that more frequent builds improves their productivity.
+Hence the developer feels more frequent builds improves  productivity.
 
-In another case, we observed a higher level of unstructured navigation than median and asked the developer about that.
-They explained that even though they agree that structured navigation is better than unstructured navigation, (s)he prefers unstructured navigation in some specific instances.
+In another case, we observed a higher level of unstructured navigation than median and asked the developer to explain.
+Even though they agree  structured navigation is better than unstructured navigation, (s)he prefers unstructured navigation in some specific instances.
 The developer is the author of the code and (s)he knows the code completely.
@@
 %Though commonly held opinion of Microsoft's Clippy experience suggests this more invasive form of feedback is not always welcome, we shall list as future work providing more immediate suggestions on how developers can improve their navigation practices.
-Developers also reiterated that they wanted a different feedback experience from Blaze  such as immediate prompts to suggest navigation commands within the context of what they are working on.
+Developers also reiterated how they wanted a different feedback experience from Blaze  such as immediate prompts to suggest navigation commands within the context of what they are working on.
 Moving towards more immediate feedback, we shared a set of achievement badges tied to specific navigation commands  as we gave the more quantitative feedback to the developers.
-Developers said that badges would motivate them to use some specific tools or commands more than points alone.
+Developers said badges would both inform and motivate them to use some specific tools or commands more than points alone.
 %  This contradicts a result from the pre-study survey where respondents rated badges as less important than team or individual competition in motivating them to improve.
@@
 \section{Threats to Validity}
-The threats to construct validity occur when measurements may not reflect the intended operational definition.The key measures in our study that could have results differing from their definition were navigation ratio and points scored in the game.
+The threats to construct validity occur when measurements may not reflect the intended operational definition.The key measures of navigation ratio and points in our study could have results differing from their definition.
 
@@
 The categorization of events is specific to this study and documented herein, however, it may not be accurate to the intent of the operational definition of structure navigation depending on how the developer uses some commands like Find.
-Another factor affecting this measurement may occur when developers repeat a command such as hitting the down arrow key many times, we might consider that something other than navigation (though it is classified as such for this paper).
+Another factor affecting this measurement may occur when developers repeat a command such as hitting the down arrow key many times, we might consider something other than navigation (though it is classified as such for this paper).
 
@@
 Because we focus on the same participants with the same behavior, they may have learned how to demonstrate improvement at some time other than when presented with the information in Blaze.
-This is mitigated by the data showing that some participants did not improve over the course of the study while others showed at least that they were attempting to improve.
+This is mitigated by the data showing some participants did not improve over the course of the study while others showed they were attempting to improve.
 
-Participants in the pilot were not randomly selected, they are all volunteers from one intact team that was chosen due to their advanced practices and understanding of metrics.
+Participants in the pilot were not randomly selected, they are all volunteers from one intact team chosen due to their advanced practices and understanding of metrics.
 Therefore the participants may be predisposed to have an interest in the kind of tool and methodology we are piloting and have bias towards a favorable opinion.
@@
 
-Threats to external validity include that the study was conducted in one development location of one company, other companies and locations may have different cultures incompatible with this type of monitoring and gamification activity.
+Threats to external validity include the study was conducted in one development location of one company, other companies and locations may have different cultures incompatible with this type of monitoring and gamification activity.
  Threats to generalizable results exist because we conducted the pilot of the tool with between 4 and 6 active developers thus the population size is too small to draw statistically significant inferences from.
@@
 \section{Conclusions}
-Developers accept gamification ideas into their environment even when coupled with detailed development activity monitoring.
-We identified specific concerns particularly around maintaining anonymity of the developer as their data are shared with other developers.
-The participants' feedback indicates a desire for more instructional hints on how to succeed in the game similar to how real games provide hints when the user might be stuck.
-Implementing a more comprehensive set of practices in the game would expand the usefulness of Blaze as a training tool.
-Other ideas might include making the game more subtle so that it is not as obvious a competition, but builds the elements of self-improvement that move the experience from an extrinsically motivated one to an intrinsically motivated one.
+We set out to demonstrate whether gamification of software development practices could succeed in the industrial environment at ABB.
+Through a series of three steps we conducted a pre-study survey, pilot of a gamified navigation practices with the Blaze tool, and post-pilot survey.
+A common thread through all three is developers are interested in this idea to help them learn and improve their practices.
+Between 60\% and 80\% of developers indicated elements like points and leaderboards influenced them in both pre-study and post-pilot surveys.
+The recorded usage data shows some developers responded positively when they started receiving points feedback.
+We learned developers need more detailed and immediate feedback from Blaze on ways to improve their practices.
+Some developers also appreciated charts providing a historical view of how they spend their time by category compared with the median for the team.
+This compliments their feedback on the game showing developers want more information on how to become more effective at their job.
+Futrue work includes providing a mechanism alerting developers when they are editing the same code, shares recomendations from mentor devopers based on observed data, and allows developers to share their data with a mentor for feedback.
+
+%  Other ideas might include making the game more subtle so that it is not as obvious a competition, but builds the elements of self-improvement that move the experience from an extrinsically motivated one to an intrinsically motivated one.
 %\end{document}  % This is where a 'short' article might terminate

### 2013-10-23 paper_id1.tex
first submitted version

@@
 
-\title{Experiences Gamifying Developer Practices}
-%\subtitle{}
+\title{Experiences Gamifying Developer Adoption\\ of Practices and Tools}
 % 
@@
 We tested an idea with industrial software developers that adding game-like feedback to the development environment would improve adoption of tools and practices for code navigation.
-We conducted this study in three stages: a pre-study to validate our ideas, pilot of a tool with some game features to motivate better navigation practices, and feedback interviews for a post-pilot survey.
-We found that a software engineering game with competitive elements influences some developers towards changing their practices.
+We present results from a pre-study survey of 130 developers' opinions on gamification and motivation, data from a pilot by an intact team of six developers of a usage monitoring tool with game features tied to code navigation practices, and feedback collected in post-pilot interviews.
+Most developers expressed interest in gamification in our survey, though some have strong negative opinions.
+Pilot results show that two of the six pilot developers adjusted their practices when presented with competitive game elements.
+%We found that a software engineering game with competitive elements influences some developers towards changing their practices.
 %We also learned that demonstrating significant improvement can be difficult and that developers want more proactive messages and recommendations on ways they can improve.
@@
 
-Sascha is a developer with years of experience maintaining code who knows how to navigate through code fairly well.
-Sascha heard about some other navigation tools, but kept using the built-in features of the Visual Studio IDE.
+Sascha is a developer with years of experience maintaining code, who knows how to navigate through code fairly well.
+Sascha heard about some other navigation tools, but kept using the built-in features of the Visual Studio IDE mainly out of habit.
 When Sascha's company started a game around code navigation in Visual Studio, he became interested in being a part of the fun.
-The game gave bonus points for using Sando, a new code search tool, so he started using it instead of some of the built-in search commands.
-Sascha soon topped the game's leaderboard and discovered he became motivated to try the next practices included in the game.
+Sascha tried the code navigation tools and practices that were part of the game and soon found himself at the top of the game's leaderboard.
+After learning the new practices Sascha spends less time navigating code to accomplish his maintenance tasks.
 
-Like Sascha, software developers who maintain code face additional challenges of navigating an existing code base to incorporate changes as they build new features or fix bugs.
-Over the years, developers learn techniques to analyze code and locate relevant areas for an enhancement.
-As new practices and tools get introduced, developers may not learn about them unless they interact with a peer developer who uses them\cite{wbsnipes:Hill2011Peer}.
-%For instance, the commands they use to navigate the code base in the development environment may make them less efficient than using newer practices or tools.
-Robillard et al.
-show that developers who use commands that allow them to follow code structure are more effective and efficient than developers who browse through code \cite{wbsnipes:Robillard2004How}.
-
-Enter the toolsmith,  who responds to developers' challenges by creating tools such as code search tools that make the job easier and more efficient.
-The toolsmith needs to spread the word about advanced tools they develop and determine their effect on developer practices over the long term.
-Is there something we can do to encourage developers to adopt new practices and tools that also provides feedback to developers and toolsmiths on the resulting improvements?
+Like Sascha, software developers who maintain code face challenges of navigating an existing code base to incorporate changes as they build new features or fix bugs.
+Toolsmiths respond to these challenges by creating new tools for assisting developers through difficult activities like code navigation.
+However, as new practices and tools become available, developers may be motivated to start using them unless they interact with a peer developer who promotes them\cite{wbsnipes:Hill2011Peer}.
+Of course they are intrinsically motivated to improve, but urgent demands interfere with experimenting using new tools and they become less efficient than otherwise possible.
+Thus we need to both communicate with and motivate developers like Sascha to start using new tools by showing them how they benefit from using the new tools.
 
@@
 Gamification is the incorporation of game elements into an activity that people do not typically consider a game\cite{2013Oxford}.
-We envisioned a tool in our prior work to motivate developers by providing constant feedback and allowing them to compare their improvements to how their peers are doing  \cite{Snipes2013Towards}.
-This tool called Blaze, creates a competitive game with continuous feedback on how well the developer adopts the improved practice.
+We envisioned a tool in our prior work that would motivate developers by providing constant feedback and allowing them to compare their improvements to how their peers are doing  \cite{Snipes2013Towards}.
+This tool, called Blaze, creates a competitive game with continuous feedback on how well the developer adopts the improved practice.
 The more they change their practices to adopt new techniques or tools, the higher they score in the game.
@@
 We piloted Blaze, our tool that introduces game elements focused on improving navigation practices into the developer's IDE.
-With the pilot feedback we determine the acceptance, influence, and effect of deploying the Blaze tool in the pilot group.
+Developers' reactions to gamification were mostly positive with a few detractors, and results of piloting show that game elements influence some developers but have no effect on others' use of tools and practices.
 
-The rest of this paper is organized as follows:  Section 2 reviews related work, Section 3 reviews results from the pre-study survey, Section 4 discusses the design considerations for developing the Blaze tool/game, Section 5 discusses the design of the Blaze pilot study,  Section 6 results from the pilot study and post-pilot survey, Section 7 covers threats to talidity, and Section 8 has our conclusions.
-%This is precisely what we set out to study for this work. 
-%
-%Our first step was to challenge this question through a pre-study survey that queried both whether developers would be interested in gamifying their workspace and what aspects of that space would be most interesting to them.
-%
-%After confirming a good level of interest from the pre-study survey, we identified a set of practices and defined a game meeting expectations from the developers feedback.
-The game, implemented in a Visual Studio add-in called Blaze, integrates into the Visual Studio workspace and establishes an anonymous competition between developers to improve their navigation practices.
-%
-%During the pilot of Blaze we observed the variation in developer navigation practices and identified some potential reasons for this variation.
-We conducted post-pilot survey and feedback sessions that evaluated developers' acceptance of game elements in the tool and their desires for additional functionality.
+The rest of this paper is organized as follows:  Section 2 reviews related work, Section 3 reviews results from the pre-study survey, Section 4 discusses the design considerations for developing the Blaze tool/game, Section 5 discusses the design of the Blaze pilot study,  Section 6 results from the pilot study and post-pilot survey, Section 7 covers threats to validity, and Section 8 has our conclusions.
 
@@
 They provide guidance on assessing gamification in the software development domain including analyzing the actors and game mechanics, defining integration with existing tools and processes, and designing the evaluation of results.
-Dbois and Tamburrelli demonstrate this guidance using a study of students working on a class project.
+Dubois and Tamburrelli demonstrate this guidance using a study of students working on a class project.
 Defining a control group and subject group of students, both groups receive the same feedback about their code comments, test coverage, code size, and static analysis rule violations.
@@
 For the US, Sweden, and Switzerland the responses indicate 60-65\% are at least somewhat interested.
+
+Comments from participants reflected a more interesting controversial reaction than these measurements might indicate.
+A developer from India said ``Gamification is a juvenile way of getting programmers to do something that you want.''  A different respondent expressed a concern that ``Gamification of best practices can have nasty side effects.''  Another participant said ``there is already high a level of competition between software development teams..'' indicating they are under considerable pressured to get the work done.
+Another developer reacted with a statement that identifies the kind of intrinsic motivation we hope for ``I am using practices and tools because I want to do my job well',' indicating that gamification was unnecessary to motivate people to improve.
+
+On the positive side, a developer from Poland said ``In a previous work place we had a 'game' between developers - scoring broken/unbroken builds and new/passed/failed tests.
+It was helping to improve quality and It worked well.
+Recommended especially for junior developers.
+''  Other comments indicated competition or points would be acceptable ``as long as it doesn't interrupt or slow down my work.
+''
+
 Because we nearly identically reused their original question, we compare the overall response from our survey to the Saatchi and Saatchi survey\cite{wbsnipes:SaatchiGameification} upon which this question was based.
@@
 	\includegraphics[width=3.4in]{ComfortWithSharing.pdf}
-	\caption{Comfortable with sharing usage information}
+	\caption{Comfort with sharing usage information}
 	\label{fig:comfortwithsharing}
@@
 
-We asked developers what would motivate them to use tools and practices and gave them 4 choices to rank.
+We asked developers what would motivate them to use tools and practices and gave them four choices to rank.
 The top motivator for using tools and practices is coworker recommendation with 75\% of respondents in Figure \ref{fig:toolandpracticemotivators}   ranking in 1st or 2nd place.
-The lowest was having badges posted on their social network profile where 33\% ranked this in 1st or 2nd place.
-Badges also ranked low in motivations for taking more software engineering training.
+Of several comments, a developer in Sweden captured the following scene, ``[The] most motivating [thing] is when a new practice / tool is discussed during a continuous improvement meeting or even coffee break, where one co-developer tells about the benefits and there is a joint discussion on where it (tool or practice) could be best applied.''
+
+The team goal ranked second behind coworker recommendation.
+This response indicated that developers would be motivated to use new tools and practices if that helped their team in a competion between development teams at ABB.
+The result shows that team competition would be first or second on the list for  47\% of developers to motivate their use of practices or tools.
+
+Just behind team competition is management mandate where developers would be asked to follow specific practices by the organization.
+The lowest was having badges posted on their social network profile where 32\% ranked this in 1st or 2nd place.
+Badges and management mandate received the most negative written comments in the survey.
+A developer from the United States captured the lower opinion of badges that several participants expressed ``I am motivated because ABB allows me to choose the tools that best work for me.
+A badge will not really do anything for me.''  Badges also ranked low in motivations for taking more software engineering training.
+
 We conclude that a facility to share coworker recommendations on tools and practices has the best opportunity to increase their adoption.
@@
 \item 95\% of developers responding would try tools and practices suggested by an automated recommendation system 
-\item Developers favor a team-oriented competition over individual awards
+\item Developers are motivated by collaboration and team goals more than mandates and individual awards
 \item Awards for performing good quality development practices ranked highest
@@
 \section{Game Design}
-
-%\begin{enumerate}[itemsep=0mm]
-%\item Define the practice we wish to measure compliance with
-%\item Develop a mapping of the practice to actions that are captured by the blaze tool from the IDE
-%\item Weight the actions according to their importance to the practice and assign points to the actions in the configuration file
-%\item Develop the intervention method
-%\item Determine the phases and length of the phase in the study according to baseline, intervention, and continuous feedback
-%\item Select participants to deploy
-%\item Define a pre-study questionnaire if one is necessary to collect associated information
-%\item Define a post-study questionnaire if necessary to collect post study information
-%\item Communicate with the participants about the study
-%\item Execute the study and report results
-%\end{enumerate}
-
-%distinguish pilot from pre-study.  How does gamification apply?
 
@@
 
-Another criteria is imposed by the instrument scope where we must limit ourselves to practices performed entirely in the IDE so they can be monitored using Blaze.
+Another criteria imposed by the instrument limits us to practices performed entirely in the IDE so they can be monitored using Blaze.
 We identified potential practices of test-driven development, debugging practices, navigation practices,  eliminating static analysis bugs, and frequent CM submissions.
@@
 
-Using these guidelines we assigned points to structured navigation commands which received 1 point each while unstructured navigation commands received 0 points.
-We gave additional emphasis to using the Sando search tool by assigning it 10 points.
+Using these guidelines we assigned points to structured navigation commands which received one point each while unstructured navigation commands received zero points.
+We gave additional emphasis to using the Sando search tool by assigning it ten points.
 To create levels, following the guidance from Hamari \cite{wbsnipes:Hamari2011Framework}, we applied an exponential curve based on points scored.
@@
 
-\subsection{Competitive Elements}
-
-By selecting an intact team, we hope to leverage aspects of motivation such as the Leaderboard to spur feelings of competition among teammates.
-The leaderboard allows developers to see their own score in relation to the top 5 people on the team.
-However, the identifier of the other participants was generated as 3 capital letters assigned based on their position on the leaderboard.
+By selecting an intact team, we hope to leverage aspects of motivation such as the Leaderboard to spur feelings of competition between teammates.
+The leaderboard allows developers to see their own score in relation to the top five people on the team.
+However, the identifier of the other participants was generated as three capital letters assigned based on their position on the leaderboard.
 Not knowing who of your colleagues was ahead of you could be  reduce motivation, but this avoids one potential source of conflict in an intact team.
@@
 Utilizing the continuous monitoring capabilities of Blaze, we designed the pilot to conduct a longitudinal study of the effect of Blaze on developers' navigation practices.
-The Blaze tool revealed information about navigation in 3 stages of 1 week duration each.
+The Blaze tool revealed information about navigation in three stages of one week duration each.
 A post-pilot survey gathered developers' opinions of Blaze and some confounding factors that could influence results.
@@
 Part of the research objective was to determine whether we can achieve adoption of a practice with alternative interventions.
-As previously discussed, we sought to provide Blaze to influence developers towards useing more structured navigation practices.
+As previously discussed, we sought to provide Blaze to influence developers towards using more structured navigation practices.
 We constructed the interventions during the pilot to roll out in stages each week they use the tool.
@@
 	\centering
-	\includegraphics[width=2.2in]{blazeWindow.png}
+	\includegraphics[width=1.6in]{blazeWindow.png}
+\includegraphics[width=1.6in]{blazeLeaderboard.png}
 	\caption{Blaze Tool Window}
@@
 
-%\subsubsection{Observations About Developers' Habits}
-%
-%We are interested in understanding whether developers are consistently using practices that make them inefficient.
-One way to check that is to determine whether our measurements are associated with specific developers in the study, in other words, can we independently classify the edits performed by each developer.
-If this is possible, then we can discuss whether developer have consistent ways of navigating and understanding for each edit.
-Thus the hypothesis is developers have consistent ways of navigating and understanding code while they are making changes.
-%
-%To test this hypothesis we construct a classification tree using Weka \cite{Hall2009WEKA} with the J48 classifier.
-Slicing the data and training the model using 10-fold cross-validation, the model identifies 53\% of the 647 edit sessions with the correct developer among 6 participants.
-The tree uses nearly all available fields including time for all categories and time of day for edits in its attempt to associate sessions with their developer generating a Kappa statistic of 0.403  and square root of the relative squared error at 101\%.
-%
-%While there is some consistency to how developers work, we cannot say that individual developers are always consistent in their work patterns.
-Thus developers are somewhat flexible in the commands they use when navigating and making changes to code and may be more open to adopting new tools and techniques.\
-
 \subsubsection{Observations About Drivers of Edit Session Duration}
@@
 The attribute search method we chose was Greedy Step-wise search selecting attributes based on their correlation with the target and stops adding attributes when the next attribute decreases the correlation.
-The process identified the top 5 out of 15 possible categories as follows:
+The top 5 out of the 15 possible categories ranked by this process are as follows:
 \begin{enumerate}[itemsep=0mm]
@@
 Using R\cite{Rcitation}, we constructed a linear regression model for Unstructured Navigation to determine the correlation with edit duration.
-The linear model found a positive correlation with unstructured navigation with a p-value <0.001 on 1694 degrees of freedom.
+The linear model found a positive correlation with unstructured navigation with a p-value <.001 on 1694 degrees of freedom.
 The model had an Adjusted R-squared of 0.64 showing a good portion of the variation in edit duration is explained by the Unstructured Navigation category.
@@
 \begin{itemize}[itemsep=0mm]
-\item Navigate To (Ctrl+,) is a fuzzy search interface that locates symbols matching your search terms.
-\item Go To Definition (F12) brings up the code where the selected symbol is defined. 
-\item View Call Hierarchy (Ctrl+K Ctrl+T) provides a two way analysis of a symbol's dependencies and uses. 
+\item Navigate To (Ctrl+,) is a fuzzy search interface that lists identifiers matching the selected string 
+\item Go To Definition (F12) brings up the code that defines the selected identifier 
+\item View Call Hierarchy (Ctrl+K Ctrl+T) provides a two way analysis of an identifier's dependencies and uses
 \item Class View (Ctrl+W, C) provides a browser and search function for classes and class hierarchy
-\item Find All References provides a list of references for an object
-\item Navigate to Event Handler brings up the event handler from the XAML editor
-\item View Class Diagram generates a class diagram for the project
-\item View Object Browser is a search tool for objects in the solution
+\item Find All References provides a list of lines that reference an identifier
+\item Navigate to Event Handler in the XAML editor shows the event handler for an object
+\item View Class Diagram generates a class diagram 
+\item View Object Browser is a search tool and browser
 \end{itemize}
@@
 The developers using Blaze received feedback in the form of points given for using structured navigation commands and using the Sando search tool\cite{Shepherd2012Sando}.
-Per the pilot design, the points display was disabled until the third week following installation to establish the efffect of instant feedback apart from other information.
+Per the pilot design, the points display was disabled until the third week following installation to establish the effect of instant feedback apart from other information.
 The points display and the navigation arrow were enabled in the third week to test whether the instant feedback was successful in driving increased use of the targeted features.
@@
 With the introduction of the points feedback, two developers' points accumulation spiked in week three due mostly to increased use of Sando.
-We informed developers at the beginning of the pilot that Sando usage would receive 10 points and using built-in structured navigation commands would receive 1 point.
+We informed developers at the beginning of the pilot that Sando usage would receive ten points and using built-in structured navigation commands would receive one point.
 
@@
 In week three, the developers could see their points feedback for the first time.
-Two developers took more notice of this than others and used Sando more in the third week.
-Their use of Sando continued in weeks 4 and 5 though with lower frequency.
-The effect of points looks greater than navigation ratio during Week 3 because of the additional emphasis of the 10 point bonus given for Sando usage.
-Developers in thir post-pilot feedback felt more guidance and more prompting included in the game would help guide them towards higher scores.
+Two developers took notice of the points and used Sando more in the third week.
+Their use of Sando continued at a lower level in weeks four and five.
+The effect of points feedback is greater than navigation ratio during week three because of the additional emphasis of the ten point bonus given for Sando usage.
+Results from our post-pilot survey where at least half the developers said the points and leaderboard influenced them "a lot" also support our conclusion that some though not all developers were influenced by the game elements in Blaze.
 
@@
 
-\subsubsection{Direct Feedback Discussions}
+\subsubsection{Post-Pilot Feedback Discussions}
 
@@
 We discovered during these feedback sessions that developers did not spend the entire time in their desktop environment because they had a virtual machine environment available for testing.
-Thus categories like testing and debugging were affected by whether or not they used the virtual environment or their desktop for these activities.
+Thus categories like testing and debugging were affected by whether they used the virtual environment or their desktop for these activities.
 Because those categories are not part of navigation, they are not discussed here.
@@
 
-%Can we get a quote here?
 One developer's build effort and the underlying build frequency was much higher than median as shown in Figure \ref{fig:developercomparison}.
@@
 If they build less frequently, even though it reduces the build effort, there are more bugs each time and the developer often spends more time in identifying the problem area than fixing it.
-Hence the developer feels more frequent builds improves  productivity.
+Hence the developer feels more frequent builds improves her productivity.
 
@@
 The developer is the author of the code and (s)he knows the code completely.
-``I know that the line I need to fix is in XYZ method and it is 4 blocks down from where I am currently.
+``I know that the line I need to fix is in XYZ method and it is four blocks down from where I am currently.
 It takes much less time for me to reach the fix location using unstructured navigation.
@@
 
-%Though commonly held opinion of Microsoft's Clippy experience suggests this more invasive form of feedback is not always welcome, we shall list as future work providing more immediate suggestions on how developers can improve their navigation practices.
-Developers also reiterated how they wanted a different feedback experience from Blaze  such as immediate prompts to suggest navigation commands within the context of what they are working on.
-Moving towards more immediate feedback, we shared a set of achievement badges tied to specific navigation commands  as we gave the more quantitative feedback to the developers.
-Developers said badges would both inform and motivate them to use some specific tools or commands more than points alone.
-%  This contradicts a result from the pre-study survey where respondents rated badges as less important than team or individual competition in motivating them to improve.
-The important difference in this case is badges serve as both a feedback mechanism and information showing which commands are important to use for improving structured navigation and their score.
+Developers also reiterated how they expected a different experience from Blaze  such as immediate prompts to suggest navigation commands within the context of what they are working on.
+This contradicts some developer comments from the pre-study survey that indicated interruptions from a game tool would be distracting and perhaps unwelcome.
+
+In the feedback, we shared a set of achievement badges tied to specific navigation commands with developers.
+Developers said badges would both inform and motivate them to use some specific tools or commands.
+This contradicts a result from the pre-study survey where respondents rated badges last in their influence motivating them to improve.
+The difference in this opinion may result from the context where developers, when presented with badges, see how badges serve as both a feedback mechanism and information on the commands for improving structured navigation and their score.
 
@@
 Measuring navigation ratio involved categorizing events into structured and unstructured navigation.
-The categorization of events is specific to this study and documented herein, however, it may not be accurate to the intent of the operational definition of structure navigation depending on how the developer uses some commands like Find.
-Another factor affecting this measurement may occur when developers repeat a command such as hitting the down arrow key many times, we might consider something other than navigation (though it is classified as such for this paper).
+The categorization of events is specific to this study and documented herein, however, it may not be accurate to the intent of the operational definition of structured navigation depending on how the developer uses some commands like Find.
+Another factor affecting this measurement may occur when developers repeat a command such as hitting the down arrow key many times, that may be something other than navigation (though it is classified as such for this paper).
 
@@
 By making the purpose of our measurement and contest known,  we may have triggered the participants to think about navigation commands they use before establishing a baseline.
-Because we focus on the same participants with the same behavior, they may have learned how to demonstrate improvement at some time other than when presented with the information in Blaze.
-This is mitigated by the data showing some participants did not improve over the course of the study while others showed they were attempting to improve.
+This threat is mitigated by the data showing four participants did not improve over the course of the study while two demonstrated some improvement.
 
@@
 Therefore the participants may be predisposed to have an interest in the kind of tool and methodology we are piloting and have bias towards a favorable opinion.
-%We know that not all participants remained in the study or continued using Blaze, however, our conclusions are based mostly on those that continued with the study.
 
-Another threat is the intact team may be more familiar with the code they maintain thus be less likely to change their navigation practices.
-We ask post survey questions of the participants about their familiarity with the code they work in and their experience levels.
-The post-pilot survey responses confirmed some cases where developers are not likely to change their navigation practices such as when they work in code they are very familiar with.
+The intact team may be more familiar with the code they maintain thus be less likely to change their navigation practices.
+The post-pilot survey responses confirmed developers are less inclined to use structured navigation when they work in code they are very familiar with.
 
 Threats to external validity include the study was conducted in one development location of one company, other companies and locations may have different cultures incompatible with this type of monitoring and gamification activity.
- Threats to generalizable results exist because we conducted the pilot of the tool with between 4 and 6 active developers thus the population size is too small to draw statistically significant inferences from.
+
+ Threats to generalizable results exist because we conducted the pilot of the tool with six developers thus the population size is too small to draw statistically significant inferences from.
 
@@
 
-Participants in the pilot were professional software engineers with at least 7 years of experience.
+Participants in the pilot were professional software engineers with at least seven years of experience.
 In other settings where developers have less experience or more the results may differ.
@@
 We set out to demonstrate whether gamification of software development practices could succeed in the industrial environment at ABB.
-Through a series of three steps we conducted a pre-study survey, pilot of a gamified navigation practices with the Blaze tool, and post-pilot survey.
-A common thread through all three is developers are interested in this idea to help them learn and improve their practices.
-Between 60\% and 80\% of developers indicated elements like points and leaderboards influenced them in both pre-study and post-pilot surveys.
-The recorded usage data shows some developers responded positively when they started receiving points feedback.
+Through a series of three steps we conducted a pre-study survey with 130 developers, a pilot of a gamified navigation practices with six developers in an intact team using the Blaze tool, and post-pilot survey with the same six developers.
+A common thread of findings through all three steps is developers are interested in the idea of using games to help them learn and improve their practices.
+On average, 74\% of developers were interested in gamified workspace in our pre-study survey, and five out of six developers indicated elements like points and leaderboards influenced them in our post-pilot survey.
+The recorded usage data shows two out of six developers in our pilot responded positively when they started receiving points feedback.
 We learned developers need more detailed and immediate feedback from Blaze on ways to improve their practices.
@@
 This compliments their feedback on the game showing developers want more information on how to become more effective at their job.
-Futrue work includes providing a mechanism alerting developers when they are editing the same code, shares recomendations from mentor devopers based on observed data, and allows developers to share their data with a mentor for feedback.
-
-%  Other ideas might include making the game more subtle so that it is not as obvious a competition, but builds the elements of self-improvement that move the experience from an extrinsically motivated one to an intrinsically motivated one.
-%\end{document}  % This is where a 'short' article might terminate
+Future work includes providing a mechanism that shares tool recommendations between developers, and allowing developers to share their data with a mentor for feedback.
 

### 2013-10-23 paper_id51.tex
Submitted version at 4:30 Tuesday

--- 2013-10-23 paper_id1.tex
@@
 
-As software development practices evolve, toolsmiths face the continuous challenge of communicating new practices and tools to the community.
+As software development practices evolve, toolsmiths face the continuous challenge of getting developers to adopt new practices and tools.
 We tested an idea with industrial software developers that adding game-like feedback to the development environment would improve adoption of tools and practices for code navigation.
-We present results from a pre-study survey of 130 developers' opinions on gamification and motivation, data from a pilot by an intact team of six developers of a usage monitoring tool with game features tied to code navigation practices, and feedback collected in post-pilot interviews.
-Most developers expressed interest in gamification in our survey, though some have strong negative opinions.
+We present results from a pre-study survey of 130 developers' opinions on gamification and motivation, usage data from a pilot with an intact team of six developers of a game on code navigation practices, and feedback collected in post-pilot interviews.
+Our pre-study survey showed that most developers were interested in gamification, though some have strong negative opinions.
 Pilot results show that two of the six pilot developers adjusted their practices when presented with competitive game elements.
@@
 
-% A category with the (minimum) three required fields
-\category{H.4}{Information Systems Applications}{Miscellaneous}
-%A category including the fourth, optional field follows...
-\category{D.2.8}{Software Engineering}{Metrics}[complexity measures, performance measures]
+\category{D.2.9}{Software}{Software Engineering} {Management} [Productivity]
 
-\terms{Human Factors, Management}
+\terms{Management, Economics, Human computer interaction}
 
-\keywords{empirical, training, learning, usability}
+\keywords{Development, Productivity, Code Navigation}
 
@@
 Of course they are intrinsically motivated to improve, but urgent demands interfere with experimenting using new tools and they become less efficient than otherwise possible.
-Thus we need to both communicate with and motivate developers like Sascha to start using new tools by showing them how they benefit from using the new tools.
+Thus, we need to both communicate with and motivate developers like Sascha to start using new tools.
 
@@
 We piloted Blaze, our tool that introduces game elements focused on improving navigation practices into the developer's IDE.
-Developers' reactions to gamification were mostly positive with a few detractors, and results of piloting show that game elements influence some developers but have no effect on others' use of tools and practices.
+Developers' reactions to gamification were mostly positive with a few detractors, and results of piloting show that game elements influence some developers, but have no effect on others' use of tools and practices.
 
@@
 
-The two areas of related work we discuss are research studies using other monitoring methods and studies applying motivation and gamification to other areas.
+The two areas of related work we discuss are research studies using other monitoring methods and studies applying gamification to other areas.
 
@@
 They look for specific refactoring commands in Eclipse and determine the amount of time developers use tools versus hand refactoring the code.
-For this study we focus on code navigation as a high-value area for improving developer efficiency.
+For this study, we focus on code navigation as a high-value area for improving developer efficiency.
 
@@
 Studies conducted using PROM include a series of studies on trends in time spent Pair Programming \cite{Coman2008Investigating}, benefits of refactoring on productivity \cite{Moser2008Case}, impact of refactoring on re-usability \cite{Moser2006Does}, and prediction of effort \cite{Abrahamsson2007Effort}.
-Studies using the PROM tool correlate the time spent editing with code metrics obtained trough source code analysis.
-Applying these data to specific research questions such as whether refactoring improves productivity \cite{Moser2008Case} shows the utility of combining automated effort data with code metric data.
-Taking a different focus on techniques, our study used more detailed events captured from the IDE  to detect how the user was navigating through the code in addition to knowing how long the editor is open for a particular file.
+Studies using the PROM tool correlate the time spent editing with code metrics obtained through source code analysis.
+Applying these data to specific research questions ,such as whether refactoring improves productivity \cite{Moser2008Case}, shows the utility of combining automated effort data with code metric data.
+Taking a different focus on techniques, our study used more detailed events captured from the IDE  to detect commands the uses to navigating through the code in addition to knowing duration of edit sessions for code modules.
 
 Murphy-Hill et al.
-study a large usage history data set and apply several different algorithms to accurately suggest IDE commands to novice users.\cite{MurphyHill2012Improving} Algorithms based on command history performed well particularly when synchronized chronologically with the recipient's usage history.
-Novice users were better served by recommendation algorithms that did not require a long history, where more expert users benefited from more sophisticated algorithms that included more history.
+study a large usage history data set and apply several different algorithms to accurately suggest IDE commands to novice users\cite{MurphyHill2012Improving}.
+Algorithms based on command history performed well particularly when synchronized chronologically with the recipient's usage history.
+Novice users were better served by recommendation algorithms that did not require a long history, while more expert users benefited from more sophisticated algorithms that included more history.
 The ``Most Widely Used'' algorithm recommends commands based on the collective usage profile of the team and performed nearly as well as the more sophisticated algorithms.
-The paper presents the methodology for making recommendations and evaluate their accuracy.
-Our work goes takes a further step evaluating how well tools can influence developers to use a specific recommended practice.
+The paper presents the methodology for making recommendations and evaluating their accuracy.
+Our work takes a further step evaluating how well tools can influence developers to use a specific recommended commands.
 
@@
 When discussing gamification, folks raise a concern about replacing an activity like software development, which is intrinsically motivated, with extrinsic motivation provided by points and achievements.
-To discover what is behind this concern, several studies are useful to understand what should be done.
+To discover what is behind this concern, we reviewed several studies on motivation and gamification.
 
 Beecham and colleagues provide a thorough analysis of existing literature on studies of motivation in Software Engineering \cite{Beecham2008Motivation}.
-They claim most common motivator found in cited references is ''the work'' motivates us.
+They claim that the most common motivator found in cited references is ''the work''.
 The list of de-motivators is also useful with common job satisfaction items like stress and inequity in recognition, plus poor quality software (low accomplishment) and lack of influence in decision making.
@@
 
-Maehr proposes an affirmative theory whereby  individuals achieve as a member of a social group choosing the behavior that meets the expectations and values of the group that is significant to them\cite{wbsnipes:MaehrCulture}.
+Maehr proposes an affirmative theory whereby  individuals achieve as a member of a social group, choosing the behavior that meets the expectations and values of the group that is significant to them\cite{wbsnipes:MaehrCulture}.
 A key factor in raising achievement motivation is establishing a social group where the person is motivated to belong and excel.
@@
 They recommend  achievement awards be setup as a secondary form of scoring so players have multiple objectives available for a particular game.
-Players should feel like they earned achievements in the game, they should not be given for trivial tasks.
+Players should feel like they earned achievements in the game; they should not be given for trivial tasks.
 
@@
 The subject group can see each others' rankings while the control group only sees their personal rankings.
-Results though inconclusive, show competition increases the desired results.
-In our study we provide participants with both individual feedback and anonymized competitive feedback and seek to observe a change in their behavior.
+Results, though inconclusive, show competition increases the desired results.
+In our study, we provide participants with both individual feedback and anonymized competitive feedback and seek to observe a change in their behavior.
 
@@
 We segmented results by country in Figure \ref{fig:gamification} and found gamification of software engineering is more interesting to developers in India, Poland, and Finland than in other major countries.
-The responses indicate developers were at least somewhat interested is 75-88\% In India, Poland, and Finland.
+The responses indicate 75-88\% of developers in India, Poland, and Finland were at least somewhat interested.
 For the US, Sweden, and Switzerland the responses indicate 60-65\% are at least somewhat interested.
 
-Comments from participants reflected a more interesting controversial reaction than these measurements might indicate.
+Comments from participants reflected a more interesting controversial reaction than these measurements might otherwise indicate.
 A developer from India said ``Gamification is a juvenile way of getting programmers to do something that you want.''  A different respondent expressed a concern that ``Gamification of best practices can have nasty side effects.''  Another participant said ``there is already high a level of competition between software development teams..'' indicating they are under considerable pressured to get the work done.
@@
 
-On the positive side, a developer from Poland said ``In a previous work place we had a 'game' between developers - scoring broken/unbroken builds and new/passed/failed tests.
-It was helping to improve quality and It worked well.
-Recommended especially for junior developers.
-''  Other comments indicated competition or points would be acceptable ``as long as it doesn't interrupt or slow down my work.
+On the positive side, a developer from Poland said, ``In a previous work place we had a 'game' between developers - scoring broken/unbroken builds and new/passed/failed tests.
+It was helping to improve quality and it worked well.
+[It is] recommended especially for junior developers.
+''  Other comments indicated competition or points are acceptable ``as long as it doesn't interrupt or slow down my work.
 ''
@@
 Because we nearly identically reused their original question, we compare the overall response from our survey to the Saatchi and Saatchi survey\cite{wbsnipes:SaatchiGameification} upon which this question was based.
-Our results land on both sides of the 75\% mark resulting from the Saatchi and Saatchi study.
+Our results land on both sides of the 75\% mark reported by the Saatchi and Saatchi study.
 
@@
 Another question was whether sharing detailed usage information with colleagues and the company as a whole would be a concern for developers.
-The question provided a graduated scale for the scope of who data were shared with and divided the questions between whether sharing was anonymous or not.
+The question provided a graduated scale for the scope of who data was shared with and divided the questions between whether sharing was anonymous or not.
 Figure \ref{fig:comfortwithsharing}  shows over 90\% of respondents are comfortable with sharing either anonymous or non-anonymous information with their team members.
 Their comfort level decreases as the scope of who the information is shared with increases.
-People were less comfortable sharing with anyone (the category for people outside ABB) particularly if the data were not kept anonymous.
-The first conclusion from these questions on sharing data is people are very willing to share information that could help their team.
+People were less comfortable sharing with anyone (the category for people outside ABB) particularly if the data was not kept anonymous.
+The first conclusion from these questions on sharing data is that people are very willing to share information that could help their team.
 Second, sharing within the company is acceptable if we take care to make the data anonymous.
@@
 We asked developers what would motivate them to use tools and practices and gave them four choices to rank.
-The top motivator for using tools and practices is coworker recommendation with 75\% of respondents in Figure \ref{fig:toolandpracticemotivators}   ranking in 1st or 2nd place.
+In Figure \ref{fig:toolandpracticemotivators}, we see that 75\% of respondents ranked coworker recommendation in 1st or 2nd place.
 Of several comments, a developer in Sweden captured the following scene, ``[The] most motivating [thing] is when a new practice / tool is discussed during a continuous improvement meeting or even coffee break, where one co-developer tells about the benefits and there is a joint discussion on where it (tool or practice) could be best applied.''
@@
 A developer from the United States captured the lower opinion of badges that several participants expressed ``I am motivated because ABB allows me to choose the tools that best work for me.
-A badge will not really do anything for me.''  Badges also ranked low in motivations for taking more software engineering training.
-
-We conclude that a facility to share coworker recommendations on tools and practices has the best opportunity to increase their adoption.
+A badge will not really do anything for me.''  We conclude that a facility to share coworker recommendations on tools and practices has the best opportunity to increase their adoption.
 
@@
 
-Developers also ranked the desirability of awards for different actions.
+In another question, developers ranked several actions that they would like to receive awards for.
 Overall 68\% of developers ranked receiving an award for consistently performing quality practices as either 1st or 2nd place.
-Thus, focusing awards on good practices would motivate developers to do them consistently.
+Thus, providing awards to developers for using good practices would motivate developers to do them consistently.
 
@@
 \item Developers are motivated by collaboration and team goals more than mandates and individual awards
-\item Awards for performing good quality development practices ranked highest
+\item Awards for good quality development practices ranked highest
 \item Over 90\% of respondents approved of sharing data with their team members
 \end{itemize}
-Thus, this pre-study survey provided a strong positive indicator for going forward with a software engineering tool focused on sharing developer data and rewarding developers for using good practices and tools.
+Thus, this pre-study survey provided a positive indicator for going forward with a software engineering tool focused on sharing developer data and rewarding developers for using good practices and tools.
 
@@
 In their paper on game design patterns, Hamari et al.
-issue a warning that assigning achievements to required tasks reduces intrinsic motivation because player autonomy is reduced by the achievements \cite{wbsnipes:Hamari2011Framework}.
+issued a warning that assigning achievements to required tasks reduces intrinsic motivation because player autonomy is reduced by the achievements \cite{wbsnipes:Hamari2011Framework}.
 Thus we avoided selecting practices like bug fixing or task completion because they are simply requirements of the job.
@@
 
-To evaluate developers using recommended tools in Visual Studio, we categorized them by their tool name.
-In the pilot we wanted developers to learn about and adopt the Sando search tool.
+To evaluate developers use of recommended tools in Visual Studio, we categorized them by their tool name.
+As part of the pilot, we wanted developers to learn about and adopt the Sando search tool.
 Sando is part of the recommended developer tool suite for ABB and supports structured navigation using code search \cite{Shepherd2012Sando}.
@@
 Achievements must feel earned to the participant so they recognize the effort to get the award was significant and feel satisfied  \cite{wbsnipes:Hamari2011Framework}.
-Hamari and Eranti describe a general relationship between achievements in games and regular game play      In games achievements are typically a parallel scoring system to the main game play.
-They make the game more engaging providing multiple ways to increase your score and multiple challenges in one interface.
+Hamari and Eranti describe a general relationship between achievements in games and regular game play.
+Achievements in games typically create a parallel scoring system to the main game play.
+They make the game more engaging by providing multiple ways to increase your score and multiple challenges in one interface.
 
-Using these guidelines we assigned points to structured navigation commands which received one point each while unstructured navigation commands received zero points.
-We gave additional emphasis to using the Sando search tool by assigning it ten points.
-To create levels, following the guidance from Hamari \cite{wbsnipes:Hamari2011Framework}, we applied an exponential curve based on points scored.
-Users initially ``level up'' after  working a day or two; the next level may be achieved after a week or even a month following an exponential curve.
-We designed the points and levels so above average developers can complete all levels during the study period.
+Considering these guidelines, we assigned one point for each use of structured navigation commands while unstructured navigation commands received zero points.
+We gave additional emphasis to using the Sando search tool by assigning ten points for each use.
+To create levels following the guidance from Hamari \cite{wbsnipes:Hamari2011Framework}, we applied an exponential curve based on points scored.
+Users initially ``level up'' after  working a day or two; the next level may require a week's work to achieve as the difficulty increases.
+We designed the points and levels so above average developers can pass all levels during the study period.
 
-By selecting an intact team, we hope to leverage aspects of motivation such as the Leaderboard to spur feelings of competition between teammates.
+By selecting an intact team for the pilot, we hope to leverage aspects of motivation such as the Leaderboard to spur feelings of competition between teammates.
 The leaderboard allows developers to see their own score in relation to the top five people on the team.
-However, the identifier of the other participants was generated as three capital letters assigned based on their position on the leaderboard.
-Not knowing who of your colleagues was ahead of you could be  reduce motivation, but this avoids one potential source of conflict in an intact team.
+However, the other participants' identifiers are auto-generated as three capital letters assigned based on their position on the leaderboard.
+Not knowing who of your colleagues was ahead of you could reduce motivation, but this avoids one potential source of conflict in an intact team.
 
@@
 
-Part of the research objective was to determine whether we can achieve adoption of a practice with alternative interventions.
-As previously discussed, we sought to provide Blaze to influence developers towards using more structured navigation practices.
+Part of the research objective was to determine whether we can drive adoption of a practice through gamification.
+As previously discussed, designed the game in Blaze to influence developers towards using more structured navigation practices.
 We constructed the interventions during the pilot to roll out in stages each week they use the tool.
@@
 
-In the initial stage, participants were informed of the purpose of the study what to expect from the Blaze tool.
+In the initial stage, participants were informed of the purpose of the study, and what to expect from the Blaze tool.
 The participants installed the tool in their Visual Studio environment.
-The tool did not present any window for them to see and simply collected data in the background.
+During this stage, Blaze did not present a window for them to see and simply collected data in the background.
 
@@
 The sub-page on navigation contained a page of tutorial on the built-in structured navigation commands in Visual Studio.
-The About Blaze page contained details of the point scoring system and background on Blaze's purpose.
+The ``About Blaze'' page contained details of the point scoring system and background on Blaze's purpose.
 The usefulness of the communication site is rated as part of our post-pilot  survey.
 
-In the third stage developers received instant feedback from Blaze on the points they accumulated during the development session.
+In the third stage, developers received instant feedback from Blaze on the points they accumulated during the development session.
 The tool's appearance at this stage is shown in Figure \ref{fig:blazeWindow}.
-The Info button provides a level indicator via colored contents and activates a chart display windows when clicked.
+The ``Info'' button provides a level indicator via colored contents and activates a chart display windows when clicked.
 Points shows the points accumulated for the developer's history.
@@
 
-As mentioned in Figure \ref{fig:gamification}, India is one country where as per our survey, the acceptance to gamification is the highest.
+As mentioned in Figure \ref{fig:gamification} from our survey, the acceptance of gamification is highest in India.
 Hence, we conducted this study with an intact team of 6 developers working in an R\&D facility in India on a large industrial software system.
-We approached a team who develops and maintains an industrial software product.
 Initially, we met the whole team along with the management staff to explain the research project and the Blaze tool we developed.
-We  explained the aim of the study, what data is collected, and how the data will be processed and managed.
+We  explained the aim of the study, what data is collected, and how the data would be processed and managed.
 
-It was important to gain the confidence of the participants that their data were kept confidential so their activities would be close as possible to real-world.
-if the participants felt like they are being tracked and the data would identify them to their peers or managers, then they may refuse to participate or deviate from their normal development style.
+It was important to gain the confidence of the participants that their data were kept confidential so their activities would be as close as possible to real-world.
+If the participants felt like they are being tracked and the data would identify them to their peers or managers, then they may refuse to participate or deviate from their normal development style.
 
-We  assured the team about the confidentiality of the data and the management reassured the data collected by the research group will not be used for any other purpose other than the study.
+We  assured the team about the confidentiality of the data, and the management reassured them that the data collected by the research group will not be used for any other purpose other than the study.
 Developers were asked to volunteer for the pilot by their management, however, managers did not know who participated in the study.
-Their name and data were kept confidential to ensure no one apart from the research team knows he/she is participating in the pilot.
-Mechanisms in our data collection to identify individuals were under the control of each developer, they could chose to share or keep confidential the association of a generated unique id with researchers or each other.
+Their names and data were kept confidential to ensure that no one apart from the research team knows he/she is participating in the pilot.
+To distinguish individuals in the data each user is assigned a generated unique ID.
+Developers could chose to share the unique ID with us or keep it confidential.
+All were willing to share their unique ID with us during the follow-up feedback sessions.
 
@@
 Data at this level of detail has not been collected in ABB development organizations.
-Researchers were careful to create the system where developers controlled their identification with the data and fortunately, management bought into the necessary anonymity of the participants.
+Researchers were careful to create the system where developers controlled whether we could associate their identity with their data, and fortunately management bought into the necessity for participant anonymity.
 
@@
 The goal for the pilot follow-up survey was to collect feedback on the tool and the effect it had on the developers' navigation practices and knowledge.
-The survey questions were segmented into portions that evaluate prior and post knowledge and practices for navigation, key steps in the pilot itself, the influence of game elements on their navigation practices, typical tasks and demographic information.
+The survey questions were segmented into portions that evaluate prior and post knowledge of navigation practices, key steps in the pilot itself, the influence of game elements on their navigation practices, typical tasks and demographic information.
 
-To find out whether Blaze had an effect on knowledge of the developers regarding structured navigation commands, we asked them to rate their knowledge of structured navigation commands prior to the pilot on a 1-10 scale to understand their baseline.
+To find out whether Blaze increased developers'  knowledge of structured navigation commands, we asked them to rate their knowledge of structured navigation commands prior to the pilot on a 1-10 scale.
 Then we asked them to contrast their knowledge after the pilot to assess what they learned from the pilot.
@@
 The other key question is did Blaze have a perceived effect on their navigation practices.
-Here too we asked them to rate how much they used structured navigation prior to the pilot using a 4 choice rating of not at all, some, many times, and as much as possible.
-Then we asked whether the used structured navigation commands about the same amount, more,  or a lot more during the pilot.
+Here too we asked them to rate how much they used structured navigation prior to the pilot using a 4 choice rating of ``not at all'', ``some'', ``many times'', and ``as much as possible''.
+Then we asked whether they used structured navigation commands ``about the same amount'', ``more'',  or ``a lot more'' during the pilot.
 These two questions assessed their perceptions of structured navigation command use.
@@
 The final assessment questions tested whether using Blaze influenced their attention to navigation commands by asking how much specific game elements influenced their practices.
-The scale started with not at all, a little, some, to a lot.
+The responses available were ``not at all'', ``a little'', ``some'', and ``a lot''.
 
 The other questions included whether they reviewed the training materials on structured navigation commands, installed the Sando code search tool, and what types of development tasks they would consider using structured navigation for in general.
-Another demographic question is how well they know the code they work in rated from ``I work in code that I wrote'', ``I maintain code that I am familiar with'', to ``I recently learned the code''
+Another demographic question asked how well they know the code they work in rated from ``I work in code that I wrote'', ``I maintain code that I am familiar with'', to ``I recently learned the code''.
 
@@
 We asked developers to rate their knowledge of the code they maintain.
-Four of the six developers reported they maintain code that they wrote while the two most experienced developers maintain code that they recently learned.
-One developer reported during our individual analysis feedback that working in code they know well affects their navigation practices leading them to use more unstructured navigation.
+Four of the six developers reported they maintain code that they wrote, while the two most experienced developers maintain code that they recently learned.
+One developer reported during our individual feedback sessions that working in code they know well affects their navigation practices leading them to use more unstructured navigation.
 
-On the assessment of structured navigation knowledge, half of the participants replied they had very good prior knowledge about structured navigation and they use structured navigation as much as possible.
-Other half of the participants have some understanding on structured navigation use it some of the time.
-Even developers whose navigation ratio  was lower than average, used structured navigation occasionally as much as their colleagues.
-While analyzing the points they scored with Blaze tool, we observed the participants who rated their prior knowledge of structured navigation highly also scored higher in the pilot.
+On the assessment of prior structured navigation knowledge, half of the participants replied they had very good prior knowledge about structured navigation and they use structured navigation ``as much as possible''.
+The other half of the participants rated their prior knowledge of structured navigation as good and reported that they use it ``some'' of the time.
+Of the three developers who rated their prior knowledge and use of structured navigation higher, two of them had the highest point scores of the group.
 
@@
 The chart in figure \ref{fig:elementInfluence} shows how the individual feedback features influenced developers to think about using structured navigation.
-Four out of six developers said the leaderboard influenced them ``a lot'' and all rated the leaderboard as having some influence.
-Three out of six  said their individual point score influenced them ``a lot'' and all said the score had some influence.
+Four out of six developers said the leaderboard influenced them ``a lot'', and all rated the leaderboard as having at least ``a little'' influence.
+Three out of six  said their individual point score influenced them ``a lot'', and all said the score had at least ``a little'' influence.
 The element with mixed ratings was a graphical indicator on whether they were improving over past use of structured navigation commands.
-The indicator received the influencing ``a lot'' rating from one participant and a ``not at all'' influence rating from one participant with the rest having some influence.
+The indicator received the influencing ``a lot'' rating from two participants and a ``not at all'' influence rating from one participant with the rest having at least ``a little'' influence.
 The results indicate the more obvious and more clearly comparative information in the leaderboard was the strongest influence for developers.
 
-The participants requested Blaze provide more feedback on tips to help them increase their score.
-They asked for more obvious feedback such as hints that pop-up when they launch the tool would help them more than passively provided information.
+The participants requested that Blaze provide more feedback on tips to help them increase their score.
+They indicated that more obvious feedback, such as hints that pop-up when they launch the tool, would help them more than passively provided information.
 Their desire for personal metrics and historical views of the data encourages us that this detailed level of measurement and analysis is perceived as helpful.
@@
 An optional field for Artifact Reference captures data such as the file-name being edited and the currently selected line number.
-Finally an anonymized unique identifier for the user of blaze allows us to investigate differences between developers.
+An anonymized unique identifier for each Blaze user allows us to investigate differences between developers.
 
-Analysis of the log extracts several key information bits from this raw source.
-First Blaze classifies the events at a higher level into categories related activities such as Navigating, Editing, Building, Debugging,  Testing, and using a  known Tool.
-Each event name maps to a category from the list and sub-division of the Navigation category separates Structured Navigation from Unstructured Navigation.
+To assist in log analysis, Blaze classifies the events into categories for related activities such as navigating, editing, building, debugging,  testing, and using a  known tool.
+Each event name maps to a category from the list and sub-division of the navigation category separates structured navigation from unstructured navigation.
 
@@
 Of the many possible ways to define sessions, we chose to define sequences as all the events that transpire before and during a set of edits to a specific file.
-When the developer edits a new file, all the events after the last edit to the previous file are part of the new sequence for the new file.
+When the developer edits a new file, all the events after the last edit to the previous file are included in the new sequence for the new file.
 For example, the developer edits file A, then navigates through files B and C finally landing at line 100 of file D where they make another edit.
-All the events between the edit of file A and the edit of file D are considered part of the edit session ending with several edits of file D.
+All the events between the edit of file A and the edit of file D are considered part of the edit session ending with the edit of file D.
 Multiple edits to file D that occur before editing a different file are included in the same sequence ending with the edits to file D.
@@
 We selected navigation as the subject of the pilot game because it was identified by Robillard et al.\cite{wbsnipes:Robillard2004How} as correlated with developers who effectively completed a maintenance task.
-In this analysis we ask whether we are focusing on the right area of developers' activity in Visual Studio?
+In this analysis, we ask whether we are focusing on the right area of developers' activity in Visual Studio?
 
@@
 Using Weka\cite{Hall2009WEKA}, we performed an attribute selection process to identify the most significant attributes related to edit session duration.
-The attribute search method we chose was Greedy Step-wise search selecting attributes based on their correlation with the target and stops adding attributes when the next attribute decreases the correlation.
+The attribute search method we chose was Greedy Step-wise search, which selects attributes based on their correlation with the target, and stops adding attributes when the next attribute decreases the correlation.
 The top 5 out of the 15 possible categories ranked by this process are as follows:
 \begin{enumerate}[itemsep=0mm]
-\item Unstructured Navigation
+\item Unstructured navigation
 \item Debug - events from running the debugger
@@
 
-Using R\cite{Rcitation}, we constructed a linear regression model for Unstructured Navigation to determine the correlation with edit duration.
+To determine the correlation with edit duration, we used R\cite{Rcitation} to construct a linear regression model for unstructured navigation .
 The linear model found a positive correlation with unstructured navigation with a p-value <.001 on 1694 degrees of freedom.
-The model had an Adjusted R-squared of 0.64 showing a good portion of the variation in edit duration is explained by the Unstructured Navigation category.
+The model had an Adjusted R-squared of 0.64 showing a good portion of the variation in edit duration is explained by the unstructured navigation category.
 
@@
 
-To evaluate whether developers improved their practices for structured navigation we established a metric Navigation ratio as the number as the number of structured navigation events in a session or period of time over the number of unstructured navigation events.
+To evaluate whether developers improved their practices for structured navigation we established a metric, Navigation ratio, as the number as the number of structured navigation events in a session or period of time over the number of unstructured navigation events.
+Our hypothesis is that instant feedback from points combined with game information would result in an increase in the navigation ratio for developers in the pilot.
 
@@
 
-Unstructured navigation events included selecting a file in an explorer window or selecting the tab for a file, using arrow and page up/down keys to go up/down through a file, scrolling, clicking on a file element, and using any of the built-in find commands such as Find In Files or Quick Find.
+Unstructured navigation events included selecting a file in an explorer window or selecting the tab for a file, using arrow and page up/down keys to go up/down through a file, scrolling, clicking on a file element, and using any of the built-in find commands such as ``Find in Files'' or ``Quick Find''.
 
-During the pilot we staged the deployment of information feedback to developers so they would first have access to learn about structured navigation and then get instant feedback in the form of points for using structured navigation commands.
-We hypothesized instant feedback and information would result in an increase in the number of structured navigation commands used in Visual Studio.
+During the pilot, we staged the deployment of information feedback to developers so they would first have access to learn about structured navigation and then get instant feedback in the form of points for using structured navigation commands.
 
@@
 We expected developers learning about navigation commands and tools that count towards the score would begin using the practices more during the second week.
-In the third week we turn on points feedback and the leaderboard expecting this will encourage developers to use more structured navigation.
+In the third week, we turn on points feedback and the leaderboard expecting this will encourage developers to use more structured navigation.
 
@@
 
-We observe visually in Figure \ref{fig:navigationaverage} the structured navigation practice did not increase significantly from the beginning to the end of the study.
+We observe visually in Figure \ref{fig:navigationaverage}, that the navigation ratio did not increase significantly from the beginning to the end of the study.
 We evaluated the week over week differences using a Wilcoxon Rank sum test \cite{RefWorks:118} comparing week one to week two and week one to week three.
-In week two when the developers were able to learn about structured navigation commands there was no significant change.
-Also in week two, two developers stopped participating in the study.
-One was working in another environment and returned, while the other dropped for the rest of the study.
-Week three there was a slight increase in the mean, however the increase is not significant.
+In week two, when the developers were able to learn about structured navigation commands, there was no significant change.
+Also in week two, two developers were not providing data.
+One was working in another environment and returned to the study, while the other dropped for the rest of the study.
+In week three there was a slight increase in the mean, however the increase is not significant.
 Weeks four and five follow the formal period of study show the navigation ratio continues in a similar range.
@@
 
-In these results we see an initial though not statistically significant effect from the points display in Blaze in Week three.
-In week three, the developers could see their points feedback for the first time.
-Two developers took notice of the points and used Sando more in the third week.
+In these results, we see an effect from the points display in Blaze in week three when the developers could see their points feedback for the first time.
+We conducted a Wilcoxon Rank sum test on points similar to the way we did for navigation ratio to determine significance.
+Across the pilot developers, the difference is not significant at the .05 level.
+The p-value for week two against week three is .19, and for week one against week three it is .31.
+We do know from feedback discussions and usage data that two developers took notice of the points and used Sando more in the third week.
 Their use of Sando continued at a lower level in weeks four and five.
 The effect of points feedback is greater than navigation ratio during week three because of the additional emphasis of the ten point bonus given for Sando usage.
-Results from our post-pilot survey where at least half the developers said the points and leaderboard influenced them "a lot" also support our conclusion that some though not all developers were influenced by the game elements in Blaze.
+Results from our post-pilot survey where at least half the developers said the points and leaderboard influenced them "a lot" also supports our conclusion that two developers were influenced by the game elements in Blaze.
 
@@
 Following our data analysis, we conducted one on one feedback sessions with the pilot participants.
-Before initiating the discussion, we assured the individual developer the data shared with them is confidential and we will not share the same with management.
-The participants received the feedback in the form of quantitative charts and written analysis of the information.
+Before initiating the discussion, we assured the individual developer that the data shared with them is confidential and we will not share the same with management.
+The participants received the feedback in the form of one-on one presentations of quantitative charts and written analysis of the information.
 The key quantitative chart shown in Figure \ref{fig:developercomparison} compared their distribution of time in each category to the median for the group.
@@
 Because those categories are not part of navigation, they are not discussed here.
-The presence of the virtual environment may explain how some developers have no activity recorded on some days.
+The developers use the virtual environment  explains some instances where developers have no activity recorded on some days.
 
@@
 
-One developer's build effort and the underlying build frequency was much higher than median as shown in Figure \ref{fig:developercomparison}.
+One developer's build effort and the underlying build frequency was much higher than the group median, see Figure \ref{fig:developercomparison}.
 We asked the developer to explain how building more frequently helps them.
@@
 Developers also reiterated how they expected a different experience from Blaze  such as immediate prompts to suggest navigation commands within the context of what they are working on.
-This contradicts some developer comments from the pre-study survey that indicated interruptions from a game tool would be distracting and perhaps unwelcome.
+This contradicts some developers' comments from the pre-study survey that indicated interruptions from a game would be distracting and unwelcome.
 
@@
 This contradicts a result from the pre-study survey where respondents rated badges last in their influence motivating them to improve.
-The difference in this opinion may result from the context where developers, when presented with badges, see how badges serve as both a feedback mechanism and information on the commands for improving structured navigation and their score.
+The difference in this opinion may result from the context where developers, when presented with badges, see how badges serve as both a feedback mechanism and guidance on the commands for improving their score and practices.
 
@@
 The categorization of events is specific to this study and documented herein, however, it may not be accurate to the intent of the operational definition of structured navigation depending on how the developer uses some commands like Find.
-Another factor affecting this measurement may occur when developers repeat a command such as hitting the down arrow key many times, that may be something other than navigation (though it is classified as such for this paper).
+Another factor affecting this measurement may occur when developers repeat a command such as hitting the down arrow key many times, that may be something other than navigation (though it is classified as such for this study).
+In feedback, we learned developers were reviewing and reading code which may account for these repetitive actions.
 
 The points measurement is more susceptible to influences of simple developer activity compared to navigation ratio.
-Thus a very busy developer could accumulate more points using a lower navigation ratio than a developer who has less activity performed more efficiently.
+Thus, a very busy developer could accumulate more points using a lower navigation ratio than a developer who has less activity performed more efficiently.
 
 Threats to internal validity are those conditions that create alternative reasons for the conclusions, or confound the conclusion.  
-In this study we may experience the Hawthorne effect where improvement occurs because we are seeking a change in a particular practice.
-Specifically, it could affect the study because a presentation and demonstration to the pilot organization was required in advance of deployment.
+The Hawthorne effect, where improvement occurs because we are seeking a change in a particular practice, could impact our results because a presentation  to the pilot organization was required in advance of deployment.
 By making the purpose of our measurement and contest known,  we may have triggered the participants to think about navigation commands they use before establishing a baseline.
@@
 
-Threats to external validity include the study was conducted in one development location of one company, other companies and locations may have different cultures incompatible with this type of monitoring and gamification activity.
+Threats to external validity include the fact that the study was conducted in one development location of one company.
+Other companies and locations may have different cultures incompatible with this type of monitoring activity.
 
@@
  The pilot was conducted in India, the country with the most positive response in our pre-study survey, perhaps where this activity was most acceptable.
-Thus other countries who rated their view of gamification lower may be less receptive.
+Thus, other countries who rated their view of gamification lower may be less receptive.
 
 Participants in the pilot were professional software engineers with at least seven years of experience.
-In other settings where developers have less experience or more the results may differ.
-The pre-study survey, experience was not part of the question set thus we could have results skewed to a particular age group.
+In other settings where developers have different levels of experience, the results may differ.
+In  the pre-study survey, experience was not part of the question set, thus it could have results skewed to a particular age group.
 

### 2014-03-14 paper_id51.tex
Changed bib file location
Submitted on Thursday

@@
 \documentclass{sig-alternate}
+\pdfpagewidth=8.5in
+          \pdfpageheight=11in
 % *** CITATION PACKAGES ***
@@
 \usepackage{enumitem}
+\usepackage{url}
 %\usepackage{mdwmath}
@@
 \begin{document}
-%
+%ICSE '14, May 31 - June 07 2014, Hyderabad, India
+
 % --- Author Metadata here ---
-\conferenceinfo{International Conference on Software Engineering}{2014 Hyderbad, India}
-%\CopyrightYear{2007} % Allows default copyright year (20XX) to be over-ridden - IF NEED BE.
-%\crdata{0-12345-67-8/90/01}  % Allows default copyright data (0-89791-88-6/97/05) to be over-ridden - IF NEED BE.
+\conferenceinfo{ICSE}{'14, May 31 -- June 7, 2014, Hyderabad, India}
+\CopyrightYear{is held by the owner/author(s). Publication rights licensed to ACM.}
+\crdata{978-1-4503-2768-8/14/06}
+
 % --- End of Author Metadata ---
@@
 We tested an idea with industrial software developers that adding game-like feedback to the development environment would improve adoption of tools and practices for code navigation.
-We present results from a pre-study survey of 130 developers' opinions on gamification and motivation, usage data from a pilot with an intact team of six developers of a game on code navigation practices, and feedback collected in post-pilot interviews.
+We present results from a pre-study survey of 130 developers' opinions on gamification and motivation, usage data from a study with an intact team of six developers of a game on code navigation practices, and feedback collected in post-study interviews.
 Our pre-study survey showed that most developers were interested in gamification, though some have strong negative opinions.
-Pilot results show that two of the six pilot developers adjusted their practices when presented with competitive game elements.
+Study results show that two of the six study developers adjusted their practices when presented with competitive game elements.
 %We found that a software engineering game with competitive elements influences some developers towards changing their practices.
@@
 Sascha is a developer with years of experience maintaining code, who knows how to navigate through code fairly well.
-Sascha heard about some other navigation tools, but kept using the built-in features of the Visual Studio IDE mainly out of habit.
+Sascha heard about some other navigation tools, but kept using the built-in features of the Visual Studio Integrated Development Environment (IDE) mainly out of habit.
 When Sascha's company started a game around code navigation in Visual Studio, he became interested in being a part of the fun.
 Sascha tried the code navigation tools and practices that were part of the game and soon found himself at the top of the game's leaderboard.
+After participating in the game, Sascha used the new practices more often and felt recognized for his achievement.
+%Deleted the following due to Review #1 comments who concluded the paper was about improving productivity using structured navigation.
 After learning the new practices Sascha spends less time navigating code to accomplish his maintenance tasks.
 
-Like Sascha, software developers who maintain code face challenges of navigating an existing code base to incorporate changes as they build new features or fix bugs.
+Toolsmiths and researchers are continuously finding ways to improve the work of software developers.
+However, as new practices and tools become available, developers like Sascha may not be motivated to start using them unless they interact with a peer developer who promotes them \cite{wbsnipes:Hill2011Peer}.
+Of course they are intrinsically motivated to improve, but urgent demands interfere with experimenting using new tools and they become less efficient than otherwise possible.
+Thus, we need to both communicate with and motivate developers like Sascha to start using new tools.
+
+%Like Sascha, software developers who maintain code face challenges of navigating an existing code base to incorporate changes as they build new features or fix bugs.
 Toolsmiths respond to these challenges by creating new tools for assisting developers through difficult activities like code navigation.
-However, as new practices and tools become available, developers may be motivated to start using them unless they interact with a peer developer who promotes them\cite{wbsnipes:Hill2011Peer}.
+However, as new practices and tools become available, developers may not be motivated to start using them unless they interact with a peer developer who promotes them \cite{wbsnipes:Hill2011Peer}.
 Of course they are intrinsically motivated to improve, but urgent demands interfere with experimenting using new tools and they become less efficient than otherwise possible.
@@
 We assess 130 developers' response and relevant concerns about the idea of gamification through a web-based survey.
-We piloted Blaze, our tool that introduces game elements focused on improving navigation practices into the developer's IDE.
-Developers' reactions to gamification were mostly positive with a few detractors, and results of piloting show that game elements influence some developers, but have no effect on others' use of tools and practices.
+As a research study, we piloted the Blaze tool that introduces game elements focused on improving navigation practices to developers using Visual Studio.
+Developers' reactions to gamification were mostly positive with a few detractors, and results of the study show that game elements influence some developers, but have no effect on others' use of tools and practices.
 
-The rest of this paper is organized as follows:  Section 2 reviews related work, Section 3 reviews results from the pre-study survey, Section 4 discusses the design considerations for developing the Blaze tool/game, Section 5 discusses the design of the Blaze pilot study,  Section 6 results from the pilot study and post-pilot survey, Section 7 covers threats to validity, and Section 8 has our conclusions.
-
-\section{Related Work}
-
-The two areas of related work we discuss are research studies using other monitoring methods and studies applying gamification to other areas.
-
-\subsection{Monitoring Practice Studies }
-
-Robillard, Coelho and Murphy explore hypotheses around how developers can be more effective at performing a maintenance task \cite{wbsnipes:Robillard2004How}.
-Robillard et al.
-say developers are more successful finding and fixing bugs when they create a detailed plan for implementing a change, use structured navigation with keyword search or cross-reference search, and only review methods once during their search.
-We build on their work by testing methods for increasing the use of structured navigation in developers' everyday practice.
-
-Johnson and Kou defined Zorro\cite{V:Johnson2007Automated}, a system for detecting whether developers use Test Driven Development techniques based on data from Hackystat.
-Zorro divides development activities into episodes delimited by events such as configuration management code check-in, start of a unit test run, or start of a build.
-Using the distinct events developers follow within these episodes, Zorro determines whether the episode followed Test-Driven Development practices per their classic definition of test first - code - test pass, or a different scenario.
-In two student-based studies comparing Zorro classifications with a simultaneous observational screen video, Zorro achieved between 70\% \cite{Kou2010Operational} and 89\% \cite{V:Johnson2007Automated} accuracy when classifying episodes into their proper TDD scenarios.
-The studies did not, however, attempt to use or evaluate the influence of instant feedback aimed at encouraging participants to follow the classic definition of Test Driven Development.
-In our study, we attempt to present data collected directly to developers in order to influence them to change their practices.
-
-Murphy-Hill, Parnin, and Black \cite{V:MurphyHill2012How} use the Mylyn Monitor to explore whether or not developers use the automated refactoring tools present in Eclipse.
-They look for specific refactoring commands in Eclipse and determine the amount of time developers use tools versus hand refactoring the code.
-For this study, we focus on code navigation as a high-value area for improving developer efficiency.
-
-The Pro Metrics (PROM) tool provides a framework for collecting data for further analysis from tools used by developers.\cite{Coman2009Casestudy}  It provides a flexible data model and a plug-in architecture to facilitate collection from different data sources.
-Studies conducted using PROM include a series of studies on trends in time spent Pair Programming \cite{Coman2008Investigating}, benefits of refactoring on productivity \cite{Moser2008Case}, impact of refactoring on re-usability \cite{Moser2006Does}, and prediction of effort \cite{Abrahamsson2007Effort}.
-Studies using the PROM tool correlate the time spent editing with code metrics obtained through source code analysis.
-Applying these data to specific research questions ,such as whether refactoring improves productivity \cite{Moser2008Case}, shows the utility of combining automated effort data with code metric data.
-Taking a different focus on techniques, our study used more detailed events captured from the IDE  to detect commands the uses to navigating through the code in addition to knowing duration of edit sessions for code modules.
-
-Murphy-Hill et al.
-study a large usage history data set and apply several different algorithms to accurately suggest IDE commands to novice users\cite{MurphyHill2012Improving}.
-Algorithms based on command history performed well particularly when synchronized chronologically with the recipient's usage history.
-Novice users were better served by recommendation algorithms that did not require a long history, while more expert users benefited from more sophisticated algorithms that included more history.
-The ``Most Widely Used'' algorithm recommends commands based on the collective usage profile of the team and performed nearly as well as the more sophisticated algorithms.
-The paper presents the methodology for making recommendations and evaluating their accuracy.
-Our work takes a further step evaluating how well tools can influence developers to use a specific recommended commands.
-
-\subsection{Gamification}
-
-When discussing gamification, folks raise a concern about replacing an activity like software development, which is intrinsically motivated, with extrinsic motivation provided by points and achievements.
-To discover what is behind this concern, we reviewed several studies on motivation and gamification.
-
-Beecham and colleagues provide a thorough analysis of existing literature on studies of motivation in Software Engineering \cite{Beecham2008Motivation}.
-They claim that the most common motivator found in cited references is ''the work''.
-The list of de-motivators is also useful with common job satisfaction items like stress and inequity in recognition, plus poor quality software (low accomplishment) and lack of influence in decision making.
-The paper also lists characteristics of people in the professions including need for independence (autonomy), desire to learn new skills and tackle new challenges.
-
-Maehr proposes an affirmative theory whereby  individuals achieve as a member of a social group, choosing the behavior that meets the expectations and values of the group that is significant to them\cite{wbsnipes:MaehrCulture}.
-A key factor in raising achievement motivation is establishing a social group where the person is motivated to belong and excel.
-Maehr states that since achievement behaviors can be triggered by circumstance, they are present in most people regardless of background.
-
-Hamari et al.
-analyze achievements patterns in game systems and provide guidance on defining successful achievement systems.
-They recommend  achievement awards be setup as a secondary form of scoring so players have multiple objectives available for a particular game.
-Players should feel like they earned achievements in the game; they should not be given for trivial tasks.
-
-Singer and Schneider demonstrate the use of a message board and points for encouraging students to increase their frequency of commits to the source code repository \cite{Singer2012It}.
-The communication mechanism enabled students to see each others' progress and resulted in more frequent commits than baseline.
-Participants valued the communication and collaboration aspect and some valued the competition enough to change their opinion on optimum commit frequency.
-The subsequent thesis by Singer \cite{Singer2013a} describes results of an experiment conducted across two iterations of a class where active feedback on commits was deployed to one course and another course served as the control group where commit frequency was simply monitored.
-Results show an increase in the frequency of commits at a statistically significant level.
-This study uses fine grained instrumentation to identify practices used between commits such as structured navigation.
-
-Dubois and Tamburrelli discuss theoretical concepts behind applying gamification to software development \cite{Dubois2013Understanding}.
-Their approach is to consider how gamification could be applied throughout the software development life-cycle to motivate many practices in different phases.
-They provide guidance on assessing gamification in the software development domain including analyzing the actors and game mechanics, defining integration with existing tools and processes, and designing the evaluation of results.
-Dubois and Tamburrelli demonstrate this guidance using a study of students working on a class project.
-Defining a control group and subject group of students, both groups receive the same feedback about their code comments, test coverage, code size, and static analysis rule violations.
-The subject group can see each others' rankings while the control group only sees their personal rankings.
-Results, though inconclusive, show competition increases the desired results.
-In our study, we provide participants with both individual feedback and anonymized competitive feedback and seek to observe a change in their behavior.
+The rest of this paper is organized as follows:  Section 2 reviews results from the pre-study survey validating the acceptance of gamification with developers, Section 3 contains the design considerations for developing the Blaze tool/game, Section 4 gives the study design for attempting to influence developers` work patterns with Blaze, Section 5 results from the study and post-study survey, Section 6 covers related work, Section 7 discusses threats to validity, and Section 8 has our conclusions.
 
@@
 
-Understanding the concerns with maintaining developers'  intrinsic motivation while adding elements of achievement motivation, we conducted an assessment survey questionnaire to determine whether the ABB developer community would be receptive to a gamification approach to software development.
+We conducted an assessment survey questionnaire to determine whether the developer community in ABB would be receptive to a gamification approach to software development.
+ABB is a multi-national corporation and established leader in power and automation technologies that enable utility and industry customers to improve performance while lowering environmental impact.
 The developer survey received over 130 responses from software developers at ABB.
@@
 
-Using a question from the Saatchi and Saatchi survey \cite{wbsnipes:SaatchiGameification} as directly as possible, we asked developers at ABB ``How interested they would be in working for a company that incorporated some aspects of games into software engineering tools as a way to increase productivity in the workplace''?
-We segmented results by country in Figure \ref{fig:gamification} and found gamification of software engineering is more interesting to developers in India, Poland, and Finland than in other major countries.
-The responses indicate 75-88\% of developers in India, Poland, and Finland were at least somewhat interested.
-For the US, Sweden, and Switzerland the responses indicate 60-65\% are at least somewhat interested.
+A consulting firm, Saatchi and Saatchi, use a question in a survey that is often cited in gamification initiatives.
+This study reuses this question \cite{wbsnipes:SaatchiGameification} tailoring it to software development by asking ``How interested they would be in working for a company that incorporated some aspects of games into software engineering tools as a way to increase productivity in the workplace''?
+Results are segmented by country in Figure \ref{fig:gamification} and found gamification of software engineering is more interesting to developers in India, Poland, and Finland than in other major countries.
+The responses indicate 74\% of developers in India, Poland, and Finland were at least somewhat interested, while the US, Sweden, and Switzerland had slightly lower interest in applying gamification to their work.
 
 Comments from participants reflected a more interesting controversial reaction than these measurements might otherwise indicate.
-A developer from India said ``Gamification is a juvenile way of getting programmers to do something that you want.''  A different respondent expressed a concern that ``Gamification of best practices can have nasty side effects.''  Another participant said ``there is already high a level of competition between software development teams..'' indicating they are under considerable pressured to get the work done.
-Another developer reacted with a statement that identifies the kind of intrinsic motivation we hope for ``I am using practices and tools because I want to do my job well',' indicating that gamification was unnecessary to motivate people to improve.
+A developer from India said ``Gamification is a juvenile way of getting programmers to do something that you want.''  A different respondent expressed a general concern that ``Gamification of best practices can have nasty side effects.''  Another participant said ``there is already high a level of competition between software development teams.'' indicating they are under considerable pressured to get the work done.
+Another developer reacted with a statement that expresses a positive intrinsic motivation ``I am using practices and tools because I want to do my job well',' indicating that gamification was unnecessary to motivate people to improve.
 
@@
 
-Because we nearly identically reused their original question, we compare the overall response from our survey to the Saatchi and Saatchi survey\cite{wbsnipes:SaatchiGameification} upon which this question was based.
+The Figure \ref{fig:gamification} compares the overall response from our survey to the Saatchi and Saatchi survey\cite{wbsnipes:SaatchiGameification} upon which this question was based.
 Our results land on both sides of the 75\% mark reported by the Saatchi and Saatchi study.
@@
 The question provided a graduated scale for the scope of who data was shared with and divided the questions between whether sharing was anonymous or not.
-Figure \ref{fig:comfortwithsharing}  shows over 90\% of respondents are comfortable with sharing either anonymous or non-anonymous information with their team members.
-Their comfort level decreases as the scope of who the information is shared with increases.
-People were less comfortable sharing with anyone (the category for people outside ABB) particularly if the data was not kept anonymous.
+Figure \ref{fig:comfortwithsharing}  shows, as the community with which the data is shared gets smaller, more developer are comfortable with sharing the data.
+If we take scope as the main concept, and limit the sharing to ``My Team'', we can say that 90\% of respondents are comfortable with sharing either anonymous or non-anonymous data if the scope of data were limited to ``My Team''.
+Developers were less comfortable sharing with anyone (the category for people outside ABB) particularly if the data was not kept anonymous.
 The first conclusion from these questions on sharing data is that people are very willing to share information that could help their team.
-Second, sharing within the company is acceptable if we take care to make the data anonymous.
+Second, sharing within the company is acceptable with anonymized data.
 
@@
 
-The team goal ranked second behind coworker recommendation.
-This response indicated that developers would be motivated to use new tools and practices if that helped their team in a competion between development teams at ABB.
-The result shows that team competition would be first or second on the list for  47\% of developers to motivate their use of practices or tools.
+The second ranked motivator for using tools and practices, ``Team Goal'', shows the survey response of ``It elevated my team`s achievement level in a game among software engineering teams at ABB.''   This response indicated that 47\% of developers would be motivated to use new tools and practices if that helped their team in a competition between development teams.
 
@@
 A developer from the United States captured the lower opinion of badges that several participants expressed ``I am motivated because ABB allows me to choose the tools that best work for me.
-A badge will not really do anything for me.''  We conclude that a facility to share coworker recommendations on tools and practices has the best opportunity to increase their adoption.
+A badge will not really do anything for me.''
+
+ The respones on this question, supported by findings in \cite{MurphyHill2012Improving,wbsnipes:Hill2011Peer,Dubois2013Understanding}, influenced the development of the studied game to include a competitive leaderboard and a common recommendation contest to influence ABB developers to change their practices.
 
@@
 
-Following the pre-study findings, we implemented a game-like tool for developers to use in Visual Studio.
+Learning from the pre-study findings, we implemented a game-like tool for developers to use in Visual Studio.
 The tool focused on motivating the adoption of good practices by including elements of games in the development environment.
-The following sections describe how we selected a practice suitable for gamification, designed the game, and included competitive elements in the game.
+This section describes the selection of a practice suitable for gamification, considerations for designing the game, and competitive elements in the game.
 
@@
 In their paper on game design patterns, Hamari et al.
-issued a warning that assigning achievements to required tasks reduces intrinsic motivation because player autonomy is reduced by the achievements \cite{wbsnipes:Hamari2011Framework}.
+stated that assigning achievements to required tasks reduces intrinsic motivation because player autonomy is reduced by the achievements \cite{wbsnipes:Hamari2011Framework}.
 Thus we avoided selecting practices like bug fixing or task completion because they are simply requirements of the job.
@@
 Another criteria imposed by the instrument limits us to practices performed entirely in the IDE so they can be monitored using Blaze.
-We identified potential practices of test-driven development, debugging practices, navigation practices,  eliminating static analysis bugs, and frequent CM submissions.
+Potential practices for gamification include  test-driven development, refactoring tool use, debugging practices, navigation practices,  eliminating static analysis bugs, and frequent configuration management submissions.
 
-Frequent CM submission and static analysis bug elimination are highlighted in this paper's related work section with prior gamification studies in classroom environments.
+Frequent configuration management submission and static analysis bug elimination are highlighted in this paper's related work section with prior gamification studies in classroom environments.
 Johnson and Kou achieved automated monitoring of test-driven development practices with Zorro \cite{V:Johnson2007Automated} thus may provide a candidate for future gamification application.
-%reference for debugging?
-Debugging practices have developer community opinions on optimal methods, however, detailed debug practices with modern IDEs have not received research community attention, thus are also reserved for future work.
+Refactoring tool use also has studies that make it a good candidate for gamification \cite{V:MurphyHill2012How,Moser2008Case}.
 
-Studies of developer effectiveness by Robillard et al.\cite{wbsnipes:Robillard2004How} identify structured navigation as a practice more effective developers use when maintaining programs.
-Structured navigation fits the requirements of a practice contained in the IDE, has proven benefits, lends itself to measurement and scoring, is not an assigned task, and leaves room for developers to excel.
+Studies of developer effectiveness by Robillard et al.\cite{wbsnipes:Robillard2004How} (discussed in Related Work \ref{sec:RelatedWork}) identify structured navigation as a practice more effective developers use when maintaining programs.
+Structured navigation is the use of navigation commands and tools that follow or represent the structure of the code.
+Using structured navigation provides a shortcut that confirms the existence of and affirms the mental model the developer has of the code.
+Robillard et al.`s  experiment~\cite{wbsnipes:Robillard2004How} showed developers who maintain mental models and navigate structurally are more effective than other developers when given the same maintenance task and code artifacts.
+Structured navigation fits the requirements of a practice contained in the IDE, shows benefits, and is not specifically assigned task for a developer.
 Thus we selected structured navigation as the practice to demonstrate gamification.
@@
 
-We categorize commands as structured navigation when they allow the developer to locate a code element by following the program structure from another element.
-These include  built-in Visual Studio commands like Navigate To \textit{Ctrl+,}, Go To Definition \textit{F12}, and Find All References \textit{Ctrl+K,R}  that follow the structured graph of the code.
+To evaluate whether developers improved their practices for structured navigation we established a metric, Navigation ratio, as the number of structured navigation events in a session or period of time over the number of unstructured navigation events.
+Our hypothesis is that instant feedback from points combined with game information would result in an increase in the navigation ratio for developers in the study.
 
-Commands we considered as unstructured navigation include scrolling through a file using cursor keys or the scroll-bar, opening files from the Solution Explorer, Find commands (e.g.
-Find in Files), children of these commands (e.g.
-Find Next), and Go-to line number.
+Structured navigation events included: 
+\begin{itemize}[itemsep=0mm]
+\item Navigate To (Ctrl+,) is a fuzzy search interface that lists identifiers matching the selected string 
+\item Go To Definition (F12) brings up the code that defines the selected identifier 
+\item View Call Hierarchy (Ctrl+K Ctrl+T) provides a two way analysis of an identifier's dependencies and uses
+\item Class View (Ctrl+W, C) provides a browser and search function for classes and class hierarchy
+\item Find All References provides a list of lines that reference an identifier
+\item Navigate to Event Handler in the XAML editor shows the event handler for an object
+\item View Class Diagram generates a class diagram 
+\item View Object Browser is a search tool and browser
+\end{itemize}
 
-To evaluate developers use of recommended tools in Visual Studio, we categorized them by their tool name.
-As part of the pilot, we wanted developers to learn about and adopt the Sando search tool.
+Unstructured navigation events included selecting a file in an explorer window or selecting the tab for a file, using arrow and page up/down keys to go up/down through a file, scrolling, clicking on a file element, and using any of the built-in find commands such as ``Find in Files'' or ``Quick Find''.
+
+To evaluate developers` use of recommended tools in Visual Studio, we categorized them by their tool name.
+As part of the study, we wanted developers to learn about and adopt the Sando search tool.
 Sando is part of the recommended developer tool suite for ABB and supports structured navigation using code search \cite{Shepherd2012Sando}.
@@
 
-By selecting an intact team for the pilot, we hope to leverage aspects of motivation such as the Leaderboard to spur feelings of competition between teammates.
+By selecting an intact team for the study, we hope to leverage aspects of motivation such as the Leaderboard to spur feelings of competition between teammates.
 The leaderboard allows developers to see their own score in relation to the top five people on the team.
@@
 
-Utilizing the continuous monitoring capabilities of Blaze, we designed the pilot to conduct a longitudinal study of the effect of Blaze on developers' navigation practices.
+Utilizing the continuous monitoring capabilities of Blaze, we designed the study to conduct a longitudinal study of the effect of Blaze on developers' navigation practices.
 The Blaze tool revealed information about navigation in three stages of one week duration each.
-A post-pilot survey gathered developers' opinions of Blaze and some confounding factors that could influence results.
+A post-study survey gathered developers' opinions of Blaze and some confounding factors that could influence results.
 
@@
 
-Part of the research objective was to determine whether we can drive adoption of a practice through gamification.
-As previously discussed, designed the game in Blaze to influence developers towards using more structured navigation practices.
-We constructed the interventions during the pilot to roll out in stages each week they use the tool.
-This allowed us to establish a baseline and determine the changes as the pilot continued.
+The key research objective was to determine whether we can drive adoption of a practice through gamification.
+As previously discussed, we designed the game in Blaze to influence developers towards using more structured navigation practices.
+We constructed the interventions during the study to roll out in stages each week they use the tool.
+This allowed us to establish a baseline and determine the changes as the study continued.
 
@@
 The ``About Blaze'' page contained details of the point scoring system and background on Blaze's purpose.
-The usefulness of the communication site is rated as part of our post-pilot  survey.
+The usefulness of the communication site is rated as part of our post-study  survey.
 
@@
 The tool's appearance at this stage is shown in Figure \ref{fig:blazeWindow}.
-The ``Info'' button provides a level indicator via colored contents and activates a chart display windows when clicked.
+The ``Info'' button provides a level indicator via colored contents and activates a chart display window when clicked.
 Points shows the points accumulated for the developer's history.
 Level is the level of the game achieved based on an exponentially growing curve.
-The developer could click on the Leaderboard tab to see how their scores compared with the others in the pilot.
-This final configuration remained as the display until the end of the pilot.
+The developer could click on the Leaderboard tab to see how their scores compared with the others in the study.
+This final configuration remained as the display until the end of the study.
 
@@
 
-\subsection{Pilot Participants}
+\subsection{Study Participants}
 
@@
 We  assured the team about the confidentiality of the data, and the management reassured them that the data collected by the research group will not be used for any other purpose other than the study.
-Developers were asked to volunteer for the pilot by their management, however, managers did not know who participated in the study.
-Their names and data were kept confidential to ensure that no one apart from the research team knows he/she is participating in the pilot.
+Developers were asked to volunteer for the study by their management, however, managers did not know who participated in the study.
+Their names and data were kept confidential to ensure that no one apart from the research team knows he/she is participating in the study.
 To distinguish individuals in the data each user is assigned a generated unique ID.
@@
 
-Management wished to have no knowledge of who was participating and took great care to remain detached from the pilot proceedings.
+Management wished to have no knowledge of who was participating and took great care to remain detached from the study proceedings.
 Data at this level of detail has not been collected in ABB development organizations.
@@
 
-\subsection{Pilot Survey}
+\subsection{Study Survey}
 
-The goal for the pilot follow-up survey was to collect feedback on the tool and the effect it had on the developers' navigation practices and knowledge.
-The survey questions were segmented into portions that evaluate prior and post knowledge of navigation practices, key steps in the pilot itself, the influence of game elements on their navigation practices, typical tasks and demographic information.
+The goal for the study follow-up survey was to collect feedback on the tool and the effect it had on the developers' navigation practices and knowledge.
+The survey questions were segmented into portions that evaluate prior and post knowledge of navigation practices, key steps in the study itself, the influence of game elements on their navigation practices, typical tasks and demographic information.
 
-To find out whether Blaze increased developers'  knowledge of structured navigation commands, we asked them to rate their knowledge of structured navigation commands prior to the pilot on a 1-10 scale.
-Then we asked them to contrast their knowledge after the pilot to assess what they learned from the pilot.
+To find out whether Blaze increased developers'  knowledge of structured navigation commands, we asked them to rate their knowledge of structured navigation commands prior to the study on a 1-10 scale.
+Then we asked them to contrast their knowledge after the study to assess what they learned.
 
 The other key question is did Blaze have a perceived effect on their navigation practices.
-Here too we asked them to rate how much they used structured navigation prior to the pilot using a 4 choice rating of ``not at all'', ``some'', ``many times'', and ``as much as possible''.
-Then we asked whether they used structured navigation commands ``about the same amount'', ``more'',  or ``a lot more'' during the pilot.
+Here too we asked them to rate how much they used structured navigation prior to the study using a 4 choice rating of ``not at all'', ``some'', ``many times'', and ``as much as possible''.
+Then we asked whether they used structured navigation commands ``about the same amount'', ``more'',  or ``a lot more'' during the study.
 These two questions assessed their perceptions of structured navigation command use.
@@
 
-\subsection{Pilot Survey Results}
+\subsection{Study Survey Results}
 
-We conducted an in-person post pilot survey in order to cross check the quantitative data from Blaze with developers' perception of their practices and their demographic information.
+We conducted an in-person post study survey in order to cross check the quantitative data from Blaze with developers' perception of their practices and their demographic information.
 
-The pilot survey queried developers' relevant demographic data to help understand some differences in results.
+The study survey queried developers' relevant demographic data to help understand some differences in results.
 We asked developers to provide their years of experience so we could evaluate any effects experience might have.
@@
 
-\subsection{Pilot Data Observations}
+\subsection{Study Data Observations}
 
@@
 In another processing step, Blaze aggregates events along the time-line into sessions.
-A session is an abstract concept grouping multiple events into a sequence with a beginning and an end.
-Of the many possible ways to define sessions, we chose to define sequences as all the events that transpire before and during a set of edits to a specific file.
-When the developer edits a new file, all the events after the last edit to the previous file are included in the new sequence for the new file.
+A session, similar to episodes defined by ~\cite{Kou2010Operational}, is an abstract concept grouping multiple events into a sequence with a beginning and an end.
+Of the many possible ways to define sessions, we chose to define edit sessions to contain all the events that transpire between edits of different files.
+When the developer edits a file different from the file last edited, all the events after the last edit to the previous file are included in the new session for the new file.
 For example, the developer edits file A, then navigates through files B and C finally landing at line 100 of file D where they make another edit.
 All the events between the edit of file A and the edit of file D are considered part of the edit session ending with the edit of file D.
-Multiple edits to file D that occur before editing a different file are included in the same sequence ending with the edits to file D.
-Although not at the method level, this way of generating sequences is similar to how Robillard et al.
-generated sequences based on located methods for their study \cite{wbsnipes:Robillard2004How}.
+When this edit is followed by multiple edits to file D, the additional edits are included in the same session ending with the edits to file D.
+Events transpiring between edits may involve activities not considered as edit related such as building and testing.
+This session definition allows us to quantify how much time in each category of events elapses before the developer edits another area of code.
+Although not at the method level, this way of generating sessions is similar to how Robillard et al.
+generated sessions based on located methods for their study \cite{wbsnipes:Robillard2004How}.
 
 \subsubsection{Observations About Drivers of Edit Session Duration}
-
-We selected navigation as the subject of the pilot game because it was identified by Robillard et al.\cite{wbsnipes:Robillard2004How} as correlated with developers who effectively completed a maintenance task.
-In this analysis, we ask whether we are focusing on the right area of developers' activity in Visual Studio?
+\label{sec:driversOfEditSession}
+We selected navigation as the subject of the Blaze study because it was identified by Robillard et al.\cite{wbsnipes:Robillard2004How} as correlated with developers who effectively completed a maintenance task.
+In this analysis, we ask whether we are focusing on the right area of developers' activity in Visual Studio.
 
@@
 Edit session duration, we assert by opinion, is a factor related to developer productivity though there are many other factors we could consider.
-With the data available from Blaze, we can explore the categories of events correlated with the duration of edit sessions over the period of the pilot.
+With the data available from Blaze, we can explore the categories of events correlated with the duration of edit sessions over the study period.
 
-Using Weka\cite{Hall2009WEKA}, we performed an attribute selection process to identify the most significant attributes related to edit session duration.
+Using Weka~\cite{Hall2009WEKA}, we performed an attribute selection process to identify the most significant attributes related to edit session duration.
 The attribute search method we chose was Greedy Step-wise search, which selects attributes based on their correlation with the target, and stops adding attributes when the next attribute decreases the correlation.
-The top 5 out of the 15 possible categories ranked by this process are as follows:
+Defining the attributes as the sum of developer time spent by category in each edit session, the top 5 out of the 15 possible categories ranked by this process are as follows:
 \begin{enumerate}[itemsep=0mm]
@@
 
-To determine the correlation with edit duration, we used R\cite{Rcitation} to construct a linear regression model for unstructured navigation .
+To determine the correlation with edit duration, we used R~\cite{Rcitation} to construct a linear regression model for unstructured navigation.
 The linear model found a positive correlation with unstructured navigation with a p-value <.001 on 1694 degrees of freedom.
@@
 
-To evaluate whether developers improved their practices for structured navigation we established a metric, Navigation ratio, as the number as the number of structured navigation events in a session or period of time over the number of unstructured navigation events.
-Our hypothesis is that instant feedback from points combined with game information would result in an increase in the navigation ratio for developers in the pilot.
+During the study, we staged the deployment of information feedback to developers so they would first have access to learn about structured navigation and then get instant feedback in the form of points for using structured navigation commands.
 
-Structured navigation events included: 
-\begin{itemize}[itemsep=0mm]
-\item Navigate To (Ctrl+,) is a fuzzy search interface that lists identifiers matching the selected string 
-\item Go To Definition (F12) brings up the code that defines the selected identifier 
-\item View Call Hierarchy (Ctrl+K Ctrl+T) provides a two way analysis of an identifier's dependencies and uses
-\item Class View (Ctrl+W, C) provides a browser and search function for classes and class hierarchy
-\item Find All References provides a list of lines that reference an identifier
-\item Navigate to Event Handler in the XAML editor shows the event handler for an object
-\item View Class Diagram generates a class diagram 
-\item View Object Browser is a search tool and browser
-\end{itemize}
-
-Unstructured navigation events included selecting a file in an explorer window or selecting the tab for a file, using arrow and page up/down keys to go up/down through a file, scrolling, clicking on a file element, and using any of the built-in find commands such as ``Find in Files'' or ``Quick Find''.
-
-During the pilot, we staged the deployment of information feedback to developers so they would first have access to learn about structured navigation and then get instant feedback in the form of points for using structured navigation commands.
-
-To evaluate whether developers use more structured navigation when they receive instant feedback, we monitor the change in navigation ratio over the pilot study period.
+To evaluate whether developers use more structured navigation when they receive instant feedback, we monitor the change in navigation ratio over the study period.
 We expected developers learning about navigation commands and tools that count towards the score would begin using the practices more during the second week.
@@
 The developers using Blaze received feedback in the form of points given for using structured navigation commands and using the Sando search tool\cite{Shepherd2012Sando}.
-Per the pilot design, the points display was disabled until the third week following installation to establish the effect of instant feedback apart from other information.
+Per the study design, the points display was disabled until the third week following installation to establish the effect of instant feedback apart from other information.
 The points display and the navigation arrow were enabled in the third week to test whether the instant feedback was successful in driving increased use of the targeted features.
@@
 With the introduction of the points feedback, two developers' points accumulation spiked in week three due mostly to increased use of Sando.
-We informed developers at the beginning of the pilot that Sando usage would receive ten points and using built-in structured navigation commands would receive one point.
+We informed developers at the beginning of the study that Sando usage would receive ten points and using built-in structured navigation commands would receive one point.
 
@@
 We conducted a Wilcoxon Rank sum test on points similar to the way we did for navigation ratio to determine significance.
-Across the pilot developers, the difference is not significant at the .05 level.
+Across developers, the difference is not significant at the .05 level.
 The p-value for week two against week three is .19, and for week one against week three it is .31.
@@
 The effect of points feedback is greater than navigation ratio during week three because of the additional emphasis of the ten point bonus given for Sando usage.
-Results from our post-pilot survey where at least half the developers said the points and leaderboard influenced them "a lot" also supports our conclusion that two developers were influenced by the game elements in Blaze.
+Results from our post-study survey where at least half the developers said the points and leaderboard influenced them ``a lot'' also supports our conclusion that two developers were influenced by the game elements in Blaze.
 
@@
 
-\subsubsection{Post-Pilot Feedback Discussions}
+\subsubsection{Post-Study Feedback Discussions}
 
-Following our data analysis, we conducted one on one feedback sessions with the pilot participants.
+Following our data analysis, we conducted one-on-one feedback sessions with the participants.
 Before initiating the discussion, we assured the individual developer that the data shared with them is confidential and we will not share the same with management.
-The participants received the feedback in the form of one-on one presentations of quantitative charts and written analysis of the information.
+The participants received the feedback in the form of one-on-one presentations of quantitative charts and written analysis of the information.
 The key quantitative chart shown in Figure \ref{fig:developercomparison} compared their distribution of time in each category to the median for the group.
+The categories were identified based on what would be most relevant to the developer where some categories match findings in the Weka analysis in section \ref{sec:driversOfEditSession}.
+The ``Actions'' category identified by Weka is omitted in the chart because it is a collection of events not categorized at a level useful for developers.
+The category for structured navigation shows the practice we are trying to encourage as the opposite of the more common unstructured navigation practice.
 
@@
 Because those categories are not part of navigation, they are not discussed here.
-The developers use the virtual environment  explains some instances where developers have no activity recorded on some days.
+Due to this, there are some instances where developers have no activity recorded on some days.
 
@@
 
+\section{Related Work}
+\label{sec:RelatedWork}
+The two areas of related work we discuss are research studies using other monitoring methods and studies applying gamification to other areas.
+
+\subsection{Monitoring Practice Studies}
+
+Robillard, Coelho and Murphy explore hypotheses around how developers can be more effective at performing a maintenance task \cite{wbsnipes:Robillard2004How}.
+Robillard et al.
+say developers are more successful finding and fixing bugs when they create a detailed plan for implementing a change, use structured navigation with keyword search or cross-reference search, and only review methods once during their search.
+We build on their work by testing methods for increasing the use of structured navigation in developers' everyday practice.
+
+Johnson and Kou defined Zorro \cite{V:Johnson2007Automated}, a system for detecting whether developers use Test Driven Development techniques based on data from Hackystat, a monitoring framework developed by Johnson.
+Zorro divides development activities into episodes delimited by events such as configuration management code check-in, start of a unit test run, or start of a build.
+Using the distinct events developers follow within these episodes, Zorro determines whether the episode followed Test-Driven Development practices in their prescribed order or a different order.
+In two student-based studies comparing Zorro classifications with a simultaneous observational screen video, Zorro achieved between 70\% \cite{Kou2010Operational} and 89\% \cite{V:Johnson2007Automated} accuracy when classifying episodes into their proper TDD scenarios.
+Data from our study also classifies a series of events into sessions (a similar concept to episodes) enabling us to determine the most significant factor for edit session duration.
+
+Murphy-Hill, Parnin, and Black \cite{V:MurphyHill2012How} use the Mylyn Monitor to explore whether or not developers use the automated refactoring tools present in Eclipse.
+They look for specific refactoring commands in Eclipse and determine the amount of time developers use tools versus hand refactoring the code.
+For this study, we focus on code navigation as a high-value area for improving developer efficiency.
+
+The Pro Metrics (PROM) tool provides a framework for collecting data for further analysis from tools used by developers.\cite{Coman2009Casestudy}  It provides a flexible data model and a plug-in architecture to facilitate collection from different data sources.
+Studies conducted using PROM include studying benefits of refactoring on productivity \cite{Moser2008Case} and impact of refactoring on re-usability \cite{Moser2006Does}.
+Studies using the PROM tool correlate the time spent editing with code metrics obtained through source code analysis.
+Taking a different focus on techniques, this study uses more detailed events captured from the IDE to detect commands used to navigating through the code in addition to knowing duration of edit sessions for code modules.
+
+Murphy-Hill et al.
+study a large usage history data set and apply several different algorithms to accurately suggest IDE commands to novice users\cite{MurphyHill2012Improving}.
+The ``Most Widely Used'' algorithm recommends commands based on the collective usage profile of the team and performed nearly as well as the more sophisticated algorithms.
+This work takes the idea of making recommendations into a game context to influence how recommendations in a game influence developers to use recommended commands.
+
+\subsection{Gamification}
+
+When discussing gamification, concerns are raised about replacing an activity like software development, which is intrinsically motivated, with extrinsic motivation provided by points and achievements.
+
+Beecham and colleagues provide a thorough analysis of existing literature on studies of motivation in Software Engineering \cite{Beecham2008Motivation}.
+They claim that the most common motivator found in cited references is ''the work''.
+The list of de-motivators is also useful with common job satisfaction items like stress and inequity in recognition, plus poor quality software (low accomplishment) and lack of influence in decision making.
+The paper also lists characteristics of people in the professions including need for independence (autonomy), desire to learn new skills and tackle new challenges.
+
+Maehr proposes an affirmative theory whereby individuals achieve as a member of a social group, choosing the behavior that meets the expectations and values of the group that are significant to them\cite{wbsnipes:MaehrCulture}.
+A key factor in raising achievement motivation is establishing a social group where the person is motivated to belong and excel.
+%Maehr also states that since achievement behaviors can be triggered by circumstance, they are present in most people regardless of background.
+
+Hamari et al.
+analyze achievement patterns in game systems and provide guidance on defining successful achievement systems~\cite{wbsnipes:Hamari2011Framework}.
+They recommend achievement awards be setup as a secondary form of scoring so players have multiple objectives available for a particular game.
+Players should feel like they earned achievements in the game instead given trivial tasks.
+
+Singer and Schneider demonstrate the use of a message board and points for encouraging students to increase their frequency of commits to the source code repository \cite{Singer2012It}.
+The communication mechanism enabled students to see each other's progress and resulted in more frequent commits than baseline.
+Participants valued the communication and collaboration aspect and some valued the competition enough to change their opinion on optimum commit frequency.
+The subsequent thesis by Singer \cite{Singer2013a} describes results of an experiment conducted across two iterations of a class where active feedback on commits was deployed to one course and another course served as the control group where commit frequency was simply monitored.
+Results show an increase in the frequency of commits at a statistically significant level.
+This study uses fine grained instrumentation to identify practices used between commits such as structured navigation.
+
+Dubois and Tamburrelli discuss theoretical concepts behind applying gamification to software development \cite{Dubois2013Understanding}.
+Their approach is to consider how gamification could be applied throughout the software development life-cycle to motivate many practices in different phases.
+They provide guidance on assessing gamification in the software development domain including analyzing the actors and game mechanics, defining integration with existing tools and processes, and designing the evaluation of results.
+Dubois and Tamburrelli demonstrate this guidance using a study of students working on a class project.
+The study results indicate that gamified environments for software engineering practices may improve results by incorporating competitive elements like leaderboards.
+
 \section{Threats to Validity}
-The threats to construct validity occur when measurements may not reflect the intended operational definition.The key measures of navigation ratio and points in our study could have results differing from their definition.
+The threats to construct validity occur when measurements may not reflect the intended operational definition.
+The key measures of navigation ratio and points in our study could have results differing from their definition.
 
@@
 The categorization of events is specific to this study and documented herein, however, it may not be accurate to the intent of the operational definition of structured navigation depending on how the developer uses some commands like Find.
-Another factor affecting this measurement may occur when developers repeat a command such as hitting the down arrow key many times, that may be something other than navigation (though it is classified as such for this study).
+Another factor affecting this measurement may occur when developers repeat a command such as hitting the down arrow key many times, which may be something other than navigation (though it is classified as such for this study).
 In feedback, we learned developers were reviewing and reading code which may account for these repetitive actions.
@@
 Threats to internal validity are those conditions that create alternative reasons for the conclusions, or confound the conclusion.  
-The Hawthorne effect, where improvement occurs because we are seeking a change in a particular practice, could impact our results because a presentation  to the pilot organization was required in advance of deployment.
+The Hawthorne effect, where improvement occurs because we are seeking a change in a particular practice, could impact our results because participants received a presentation on the study objectives prior to the start of the study.
 By making the purpose of our measurement and contest known,  we may have triggered the participants to think about navigation commands they use before establishing a baseline.
@@
 
-Participants in the pilot were not randomly selected, they are all volunteers from one intact team chosen due to their advanced practices and understanding of metrics.
-Therefore the participants may be predisposed to have an interest in the kind of tool and methodology we are piloting and have bias towards a favorable opinion.
+Participants in the study were not randomly selected, they are all volunteers from one intact team chosen due to their advanced practices and understanding of metrics.
+Therefore the participants may be predisposed to have an interest in the kind of tool and methodology we are studying and have bias towards a favorable opinion.
 
 The intact team may be more familiar with the code they maintain thus be less likely to change their navigation practices.
-The post-pilot survey responses confirmed developers are less inclined to use structured navigation when they work in code they are very familiar with.
+The post-study survey responses confirmed developers are less inclined to use structured navigation when they work in code they are very familiar with.
 
@@
 
- Threats to generalizable results exist because we conducted the pilot of the tool with six developers thus the population size is too small to draw statistically significant inferences from.
+ Threats to generalizable results exist because we conducted the study with six developers thus the population size is too small to draw statistically significant inferences from.
 
- The pilot was conducted in India, the country with the most positive response in our pre-study survey, perhaps where this activity was most acceptable.
+ The study was conducted in India, the country with the most positive response in our pre-study survey, perhaps where this activity was most acceptable.
 Thus, other countries who rated their view of gamification lower may be less receptive.
 
-Participants in the pilot were professional software engineers with at least seven years of experience.
+Participants in the study were professional software engineers with at least seven years of experience.
 In other settings where developers have different levels of experience, the results may differ.
@@
 We set out to demonstrate whether gamification of software development practices could succeed in the industrial environment at ABB.
-Through a series of three steps we conducted a pre-study survey with 130 developers, a pilot of a gamified navigation practices with six developers in an intact team using the Blaze tool, and post-pilot survey with the same six developers.
+Through a series of three steps we conducted a pre-study survey with 130 developers, a study of a gamified navigation practices with six developers in an intact team using the Blaze tool, and post-study survey with the same six developers.
 A common thread of findings through all three steps is developers are interested in the idea of using games to help them learn and improve their practices.
-On average, 74\% of developers were interested in gamified workspace in our pre-study survey, and five out of six developers indicated elements like points and leaderboards influenced them in our post-pilot survey.
-The recorded usage data shows two out of six developers in our pilot responded positively when they started receiving points feedback.
+On average, 74\% of developers were interested in gamified workspace in our pre-study survey, and five out of six developers indicated elements like points and leaderboards influenced them in our post-study survey.
+The recorded usage data shows two out of six developers in our study responded positively when they started receiving points feedback.
 We learned developers need more detailed and immediate feedback from Blaze on ways to improve their practices.
@@
 \section{Acknowledgments}
-The authors wish to thank ABB and the pilot participants in India for their support of this research.
+The authors wish to thank ABB and the study participants in India for their support of this research.
+Thanks specifically to Vinay Augustine, David Shepherd, Suzanne Snipes, and Balaji W for your support and contributions to Blaze.
 
@@
 \balance
-\bibliography{../mousetap} 
+\bibliography{mousetap} 
  % sigproc.bib is the name of the Bibliography in this case

### 2014-03-17 paper_id51.tex
cleanup and final license updates

@@
 % --- Author Metadata here ---
+\permission{Permission to make digital or hard copies of all or part of this work for personal or classroom use is granted without fee provided that copies are not made or distributed for profit or commercial advantage and that copies bear this notice and the full citation on the first page.
+Copyrights for components of this work owned by others than the author(s) must be honored.
+Abstracting with credit is permitted.
+To copy otherwise, or republish, to post on servers or to redistribute to lists, requires prior specific permission and/or a fee.
+Request permissions from Permissions@acm.org.}
 \conferenceinfo{ICSE}{'14, May 31 -- June 7, 2014, Hyderabad, India}
@@
 \balance
-\bibliography{mousetap} 
+%\bibliography{mousetap} 
  % sigproc.bib is the name of the Bibliography in this case
@@
 %
+\begin{thebibliography}{10}
+
+\bibitem{2013Oxford}
+{\em {Oxford Dictionaries Online - American English (US)}}.
+\newblock Oxford University Press, 2013.
+
+\bibitem{Beecham2008Motivation}
+S.~Beecham, N.~Baddoo, T.~Hall, H.~Robinson, and H.~Sharp.
+\newblock {Motivation in Software Engineering: A Systematic Literature Review}.
+\newblock {\em Inf. Softw. Technol.}, 50(9-10):860--878, Aug. 2008.
+
+\bibitem{Coman2009Casestudy}
+I.~D. Coman, A.~Sillitti, and G.~Succi.
+\newblock {A Case-study on Using an Automated In-process Software Engineering
+  Measurement and Analysis System in an Industrial Environment}.
+\newblock In {\em International Conference on Software Engineering, ICSE},
+  pages 89--99. IEEE, May 2009.
+
+\bibitem{Dubois2013Understanding}
+D.~J. Dubois and G.~Tamburrelli.
+\newblock {Understanding Gamification Mechanisms for Software Development}.
+\newblock In {\em Foundations of Software Engineering}. ACM, 2013.
+
+\bibitem{Hall2009WEKA}
+M.~Hall, E.~Frank, G.~Holmes, B.~Pfahringer, P.~Reutemann, and I.~H. Witten.
+\newblock {The WEKA Data Mining Software: An Update}.
+\newblock {\em SIGKDD Explorations}, 11(1), 2009.
+
+\bibitem{wbsnipes:Hamari2011Framework}
+J.~Hamari and V.~Eranti.
+\newblock {Framework for Designing and Evaluating Game Achievements}.
+\newblock In {\em Digra Conference}, Sept. 2011.
+
+\bibitem{RefWorks:118}
+T.~Hill and P.~Lewicki.
+\newblock {\em {Statistics: Methods and Applications}}.
+\newblock StatSoft, Inc, 2007.
+
+\bibitem{V:Johnson2007Automated}
+P.~M. Johnson and H.~Kou.
+\newblock {Automated Recognition of Test-Driven Development with Zorro}.
+\newblock In {\em Agile Conference (AGILE)}, pages 15--25. IEEE, Aug. 2007.
+
+\bibitem{Kou2010Operational}
+H.~Kou, P.~Johnson, and H.~Erdogmus.
+\newblock {Operational Definition and Automated Inference of Test-driven
+  Development with Zorro}.
+\newblock {\em Automated Software Engineering}, 17(1):57--85, 2010.
+
+\bibitem{wbsnipes:MaehrCulture}
+M.~Maehr.
+\newblock {Culture and Achievement Motivation.}
+\newblock {\em American Psychologist}, 1974(29):887--896, 1974.
+
+\bibitem{Moser2008Case}
+R.~Moser, P.~Abrahamsson, W.~Pedrycz, A.~Sillitti, and G.~Succi.
+\newblock {A Case Study on the Impact of Refactoring on Quality and
+  Productivity in an Agile Team}.
+\newblock In B.~Meyer and et~al., editors, {\em Balancing Agility and Formalism
+  in Software Engineering}, volume 5082 of {\em Lecture Notes in Comp. Sci.},
+  pages 252--266. Springer], 2008.
+
+\bibitem{Moser2006Does}
+R.~Moser, A.~Sillitti, P.~Abrahamsson, and G.~Succi.
+\newblock {Does Refactoring Improve Reusability?}
+\newblock In M.~Morisio, editor, {\em Reuse of Off-the-Shelf Components},
+  volume 4039 of {\em Lecture Notes in Comp. Sci.}, pages 287--297. Springer
+  Berlin Heidelberg, 2006.
+
+\bibitem{MurphyHill2012Improving}
+E.~Murphy-Hill, R.~Jiresal, and G.~C. Murphy.
+\newblock {Improving Software Developers' Fluency by Recommending Development
+  Environment Commands}.
+\newblock In {\em Foundations of Software Engineering}. ACM, Nov. 2012.
+
+\bibitem{wbsnipes:Hill2011Peer}
+E.~Murphy-Hill and G.~C. Murphy.
+\newblock {Peer Interaction Effectively, Yet Infrequently, Enables Programmers
+  to Discover New Tools}.
+\newblock In {\em ACM Conference on Computer supported cooperative work}, CSCW.
+  ACM, 2011.
+
+\bibitem{V:MurphyHill2012How}
+E.~Murphy-Hill, C.~Parnin, and A.~P. Black.
+\newblock {How We Refactor, and How We Know It}.
+\newblock {\em IEEE Transactions on Software Engineering}, 38:5--18, Jan. 2012.
+
+\bibitem{Rcitation}
+{R Core Team}.
+\newblock {\em R: A Language and Environment for Statistical Computing}.
+\newblock R Foundation for Statistical Computing, Vienna, Austria, 2013.
+
+\bibitem{wbsnipes:Robillard2004How}
+M.~P. Robillard, W.~Coelho, and G.~C. Murphy.
+\newblock {How Effective Developers Investigate Source Code: an Exploratory
+  Study}.
+\newblock {\em IEEE Transactions on Software Engineering}, 30:889--903, Dec.
+  2004.
+
+\bibitem{wbsnipes:SaatchiGameification}
+{Saatchi and Saatchi S}.
+\newblock Gameification for business brands and loyalty, June 2011.
+
+\bibitem{Shepherd2012Sando}
+D.~Shepherd, K.~Damevski, B.~Ropski, and T.~Fritz.
+\newblock {Sando: an Extensible Local Code Search Framework}.
+\newblock In {\em Foundations of Software Engineering}. ACM, 2012.
+
+\bibitem{Singer2013a}
+L.~Singer.
+\newblock {\em {I}mproving the {A}doption of {S}oftware {E}ngineering
+  {P}ractices {T}hrough {P}ersuasive {I}nterventions}.
+\newblock PhD thesis, Gottfried Wilhelm Leibniz Universität Hannover, 2013.
+
+\bibitem{Singer2012It}
+L.~Singer and K.~Schneider.
+\newblock {It was a Bit of a Race: Gamification of Version Control}.
+\newblock In {\em Second International Workshop on Games and Software
+  Engineering: Realizing User Engagement with Game Engineering Techniques
+  (GAS)}. IEEE, June 2012.
+
+\bibitem{Snipes2013Towards}
+W.~Snipes, V.~Augustine, A.~R. Nair, and E.~M. Hill.
+\newblock {Towards Recognizing and Rewarding Efficient Developer Work
+  Patterns}.
+\newblock In {\em International Conference on Software Engineering}, ICSE.
+  IEEE, 2013.
+
+\end{thebibliography}
 
